---
title: "Getting Data In: Files, Import Checks, Dates"
subtitle: "BIOS 6301 · Introduction to Statistical Computing"
author: "Department of Biostatistics, Vanderbilt University Medical Center"
date: "Session 5"
date-format: "[Session 5]"
format:
  revealjs:
    theme: simple
    css: ../../slides.css
    slide-number: c/t
    show-slide-number: all
    incremental: false
    width: 1280
    height: 720
    fig-align: center
    code-line-numbers: false
    highlight-style: github
    footer: "BIOS 6301 · Session 5 · Getting data in"
    navigation-mode: linear
execute:
  enabled: true
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false
library(knitr)

# Synthetic teaching data for BIOS 6301. Not real patients.
set.seed(6301)
n <- 60
trial <- data.frame(
  id        = sprintf("P%03d", 1:n),
  arm       = rep(c("Placebo", "Treatment"), each = n / 2),
  site      = sample(1:4, n, replace = TRUE),
  age       = round(rnorm(n, 61, 9)),
  sex       = sample(c("F", "M"), n, replace = TRUE),
  weight_kg = round(rnorm(n, 82, 15), 1),
  sbp_0     = round(rnorm(n, 148, 14)),
  stringsAsFactors = FALSE
)
trial$sbp_12 <- round(trial$sbp_0 -
  ifelse(trial$arm == "Treatment", rnorm(n, 14, 8), rnorm(n, 3, 8)))

# ---------------------------------------------------------------------------
# Everything below manufactures the FILES this deck reads, so that every import
# demonstration on the slides is a genuine import of a genuine file rather than
# a transcript. Derived from `trial`; what we broke, and where each break is
# used:
#   creat      character, because 3 records hold "NA" and 2 hold "<0.30".
#              One of the "NA" records is row 2, so the defect is visible in
#              the first few lines of the file on the readLines() slide.
#   wt         -99 for "not recorded" in 4 records   (session 3 callback)
#   sbp12      -99 in 2 records
#   sex        an empty field in 2 records
#   visit_0    dates written as m/d/Y
#   visit_12   dates written as m/d/Y, except 3 records written as d-b-Y
#   dob_ser    dates written as Windows Excel serial numbers
# Plus three small files: a semicolon/decimal-comma export, a REDCap-style
# export with metadata lines above the header, and a Latin-1 encoded file.
# ---------------------------------------------------------------------------
dir.create("data", showWarnings = FALSE)

visit_0  <- as.Date("2023-01-09") + sample(0:180, n, replace = TRUE)
visit_12 <- visit_0 + 84 + sample(-6:6, n, replace = TRUE)
dob      <- visit_0 - round(trial$age * 365.25) - sample(0:300, n, replace = TRUE)

export <- data.frame(
  id       = trial$id,
  arm      = trial$arm,
  sex      = trial$sex,
  wt       = trial$weight_kg,
  sbp12    = trial$sbp_12,
  creat    = as.character(round(rnorm(n, 0.95, 0.25), 2)),
  visit_0  = format(visit_0, "%m/%d/%Y"),
  visit_12 = format(visit_12, "%m/%d/%Y"),
  dob_ser  = as.numeric(dob - as.Date("1899-12-30")),
  stringsAsFactors = FALSE
)
export$creat[c(2, 23, 41)]    <- "NA"
export$creat[c(12, 55)]       <- "<0.30"
export$wt[c(3, 19, 44, 58)]   <- -99
export$sbp12[c(9, 30)]        <- -99
export$sex[c(5, 27)]          <- ""
export$visit_12[c(4, 17, 38)] <- format(visit_12[c(4, 17, 38)], "%d-%b-%Y")
write.csv(export, "data/trial_export.csv", row.names = FALSE)

# A site that exports from a German-locale Excel: semicolons, decimal commas.
writeLines(c("id;wt;sbp12",
             "P061;90,4;160",
             "P062;67,8;162",
             "P063;101,2;139"), "data/site4_export.csv")

# A REDCap-style export: two metadata lines and a blank one above the header.
writeLines(c("Study 42 - Blood Pressure Substudy",
             "Exported 2024-03-04 09:12 by s.zhao",
             "",
             "Patient ID,Weight (kg),SBP @ 12wk,% change",
             "P061,90.4,142,-11.2",
             "P062,67.8,151,-6.8"), "data/redcap_export.csv")

# Latin-1 bytes, written as raw so the file really is mis-encoded.
writeBin(c(charToRaw("id,clinician\nP061,Bj"), as.raw(0xf6),
           charToRaw("rn Andr"), as.raw(0xe9), charToRaw("\n")),
         "data/clinicians.csv")
```

## Where we are {.smaller}

| Course arc | Sessions | |
|---|---:|---|
| Setup and workflow | 0–2 |  |
| **Data foundations** | **3–8** | **← today** |
| Modern tools | 9–10 |  |
| R programming | 11–13 |  |
| Statistical computing | 14–16 |  |

Today's deck develops one part of the current arc; later-session tools are named only when they clarify where this idea will be used.


## What you will be able to do {.smaller}

::: {.smaller}
[1]{.objective-num} **Import** a delimited file with `read.csv()`, and **state**
rather than guess what type each column should be.

[2]{.objective-num} **Convert** missing-value codes to `NA` at import, and say
why the door is the right place to do it.

[3]{.objective-num} **Diagnose** a file that is not what it claims to be — wrong
delimiter, header junk, wrong encoding.

[4]{.objective-num} **Read** Excel, SPSS and Stata files, and know what each
format drags along with it.

[5]{.objective-num} **Choose** between `write.csv()` and `saveRDS()`, and explain
why the workspace is never saved.

[6]{.objective-num} **Parse** dates, do arithmetic on them, and rescue Excel
serial numbers.

[7]{.objective-num} **Recognise** the import failures that produce a wrong
answer rather than an error.
:::


## Roadmap for today {.smaller}

| | Block | Minutes |
|---|---|---|
| 1 | A file is not a data frame: `read.csv()` and its arguments | 15 |
| 2 | Missing-value codes, handled at the door | 12 |
| 3 | When the file is not a friendly CSV | 15 |
| 4 | Excel, SPSS, Stata — and `readr` | 10 |
| 5 | Getting data back out; never saving the workspace | 8 |
| 6 | Dates and times | 25 |
| 7 | Practice and summary | 5 |

: {tbl-colwidths="[6,64,30]"}

. . .

Blocks 1–3 are one habit repeated: **look, then state your intent, then verify.**


# 1 · A file is not a data frame {.section-divider}

## What `read.csv()` actually does {.smaller}

```{mermaid}
%%| echo: false
%%| fig-width: 10
flowchart LR
  F["<b>bytes on disk</b><br/>trial_export.csv"] --> D["split into<br/>lines and fields<br/><i>sep, quote, dec</i>"]
  D --> E["decode characters<br/><i>fileEncoding</i>"]
  E --> H["header row?<br/><i>header, skip</i>"]
  H --> M["which text<br/>means missing?<br/><i>na.strings</i>"]
  M --> G["<b>guess a type<br/>per column</b><br/><i>colClasses</i>"]
  G --> DF["data.frame"]
  style F fill:#f4efe4,stroke:#CFAE70
  style G fill:#CFAE70,stroke:#8a7448,color:#fff
  style DF fill:#CFAE70,stroke:#8a7448,color:#fff
```

. . .

**Six decisions, and you made none of them.** Each argument named under a box is
a place where R chose for you — correctly most of the time, and silently either way.


## Look at the file before you read it {.smaller}

```{r}
#| label: readlines
writeLines(readLines("data/trial_export.csv", n = 4))
```

. . .

::: {.smaller}
Four lines of raw text, and already visible:

- **quoted** character fields, unquoted numbers — so `write.csv()` made this
- `"NA"` as *text*, in a creatinine column
- `-99` where a weight should be
- dates as `04/08/2023` — **month first, or day first?**
:::


## `read.csv()`, argument by argument {.smaller}

::: {style="font-size:0.80em"}
| Argument | Default | Set it when |
|---|---|---|
| `file` | — | Always. A **relative** path, via `here::here()` |
| `header` | `TRUE` | The file has no header row → `FALSE`, then set `col.names` |
| `sep` | `","` | Tabs (`"\t"`), semicolons (`";"`), pipes (`"|"`) |
| `dec` | `"."` | A European export uses a decimal comma → `","` |
| `na.strings` | `"NA"` | **Almost always.** Block 2 |
| `colClasses` | guessed | You know what the types should be. Block 1 |
| `skip` | `0` | Metadata lines sit above the header |
| `nrows` | `-1` | Peeking at a large file, or a junk footer |
| `check.names` | `TRUE` | You want the original column names → `FALSE` |
| `fileEncoding` | native | Accented characters arrive as rubbish. Block 3 |
| `strip.white` | `FALSE` | Unquoted fields have padding spaces |
| `comment.char` | `""` | A `#` legend at the foot of the file |
:::


## What type guessing gave you {.smaller}

```{r}
#| label: naive-read
raw <- read.csv("data/trial_export.csv")
str(raw)
```


## State the types instead of guessing {.smaller}

```{r}
#| label: colclasses
raw <- read.csv(
  "data/trial_export.csv",
  colClasses = c(id = "character", arm = "factor", sex = "factor",
                 wt = "numeric", sbp12 = "integer", creat = "character",
                 visit_0 = "character", visit_12 = "character",
                 dob_ser = "integer")
)
str(raw[, c("id", "arm", "sbp12")])
```

. . .

::: {.smaller}
`colClasses` is a **contract**. If the file changes shape — a new value in a
numeric column, a renamed field — you find out at import instead of three chunks
later.
:::


## `stringsAsFactors`: what changed in R 4.0 {.smaller}

::: {style="font-size:0.9em"}
| | Before R 4.0 | R 4.0 and later |
|---|---|---|
| `read.csv()` default | character → **factor** | character stays **character** |
| `data.frame()` default | character → **factor** | character stays **character** |
:::

. . .

::: {.smaller}
So a script written in 2018 and run today gives you *different objects*. Symptoms:
`levels()` returning `NULL`, a model with a numeric term where a categorical one
was intended, and the `as.numeric(as.factor(x))` catastrophe from session 4 —
which used to happen by accident and now takes effort.
:::

. . .

::: {.callout-note appearance="simple"}
Write `stringsAsFactors = FALSE` explicitly in inherited code you are repairing.
Not because you need it — because it documents that you noticed.
:::


# 2 · Missing values, handled at the door {.section-divider}

## "Missing" is written a dozen ways {.smaller}

::: {style="font-size:0.82em"}
| In the file | Comes from | Import default |
|---|---|---|
| `NA` | R, or a well-behaved export | `NA` ✓ |
| *empty field* | Excel, REDCap, almost everything | `NA` if numeric, `""` if character |
| `.` | SAS | text → **column becomes character** |
| `-99`, `-999`, `999`, `9999` | Fortran-era CRFs, SPSS conventions | a number ✗ |
| `NULL`, `N/A`, `#N/A`, `n/a` | databases, Excel formulas | text → character ✗ |
| `Unknown`, `Not recorded`, `Missing` | free-text data entry | text → character ✗ |
| `-1` for a count, `0` for a date | a field that "cannot be blank" | a number ✗ |
:::

. . .

Session 3 showed you how to **find** these. Today they never get in.


## `na.strings` — do it at the door {.smaller}

```{r}
#| label: na-strings
clean <- read.csv("data/trial_export.csv",
                  na.strings = c("NA", "", "-99", "999"))
summary(clean$wt)
colSums(is.na(clean))
```


## What `na.strings` cannot fix {.smaller}

```{r}
#| label: na-strings-limits
str(clean$creat)
clean$creat[c(2, 12)]
```

. . .

::: {.smaller}
The three `"NA"` records became `NA`. The two `"<0.30"` records did not, because
`"<0.30"` is not a missing value — it is **a measurement below the assay's limit
of detection**, and throwing it away is a modelling decision, not a cleaning one.
:::

. . .

::: {.callout-warning appearance="simple"}
The column is still `character`. One unhandled value is enough — session 3.
:::


## Then prove it {.smaller}

```{r}
#| label: prove-it
summary(clean[, c("wt", "sbp12")])
table(clean$sex, useNA = "ifany")
```

. . .

::: {.smaller}
`na.strings` states an **intention**. These two lines are the **evidence**. An
import is not finished until the ranges are plausible and the `NA` counts are
ones you can explain.
:::


# 3 · When the file is not a friendly CSV {.section-divider}

## Delimiters and the decimal comma {.smaller}

```{r}
#| label: euro-bad
writeLines(readLines("data/site4_export.csv", n = 2))
str(read.csv("data/site4_export.csv"))
```

. . .

```{r}
#| label: euro-good
str(read.csv("data/site4_export.csv", sep = ";", dec = ","))
```


## Header junk, and names R rewrites {.smaller}

```{r}
#| label: junk-bad
#| error: true
writeLines(readLines("data/redcap_export.csv", n = 4))
read.csv("data/redcap_export.csv")
```

. . .

```{r}
#| label: junk-good
names(read.csv("data/redcap_export.csv", skip = 3))
names(read.csv("data/redcap_export.csv", skip = 3, check.names = FALSE))
```


## Encodings {.smaller}

```{r}
#| label: encoding
lines <- readLines("data/clinicians.csv")
validUTF8(lines)
```

. . .

::: {.smaller}
Line 2 is not valid UTF-8. It is **Latin-1** — one byte per character — and R
read it as though it were UTF-8.
:::

```r
# the fix: tell the reader what the bytes mean
read.csv("data/clinicians.csv", fileEncoding = "latin1")
#>     id      clinician
#> 1 P061  Björn André
```


# 4 · Excel, SPSS, Stata — and readr {.section-divider}

## Excel via `readxl` {.smaller}

```r
library(readxl)

# What sheets are in here?
excel_sheets("data/study42.xlsx")
#> [1] "ReadMe"  "Enrolment"  "Labs"  "Sheet3"

labs <- read_excel(
  "data/study42.xlsx",
  sheet = "Labs",
  range = "A4:H210",              # skip the title block, stop before the totals
  col_types = c("text", "text", "date", rep("numeric", 5)),
  na = c("", "NA", "-99")
)
```

. . .

::: {.callout-warning appearance="simple"}
Excel is a **data-entry application**, not a file format. It silently reformats
IDs as numbers, turns gene names into dates, stores a date as a number whose
origin depends on which decade the workbook was made in, and lets a total row sit
below the data.
:::


## SPSS and Stata via `haven` {.smaller}

```r
library(haven)

d <- read_sav("data/study42.sav")     # SPSS   .sav
d <- read_dta("data/study42.dta")     # Stata  .dta
d <- read_xpt("data/adsl.xpt")        # SAS transport  .xpt
```

. . .

::: {.smaller}
These formats carry **metadata R has no native place for**: variable labels,
value labels, user-defined missing codes. `haven` keeps them on a `labelled`
column.
:::

```r
attr(d$arm, "labels")      #  Placebo = 1, Treatment = 2
as_factor(d$arm)           #  labelled -> factor, using the labels
zap_labels(d$arm)          #  labelled -> plain numeric, labels discarded
d$sex <- as_factor(d$sex)  #  do this once, deliberately, at import
```


## `readr::read_csv()` {.smaller}

```r
library(readr)
d <- read_csv("data/trial_export.csv")
#> Rows: 60 Columns: 9
#> -- Column specification ------------------------------------
#> chr (6): id, arm, sex, creat, visit_0, visit_12
#> dbl (3): wt, sbp12, dob_ser
```

. . .

::: {.smaller}
That message is the **best thing about it** — the parser tells you what it
decided, every time. Then you replace the guess with a contract:
:::

```r
d <- read_csv(
  "data/trial_export.csv",
  col_types = cols(id = col_character(), wt = col_double(),
                   visit_0 = col_date(format = "%m/%d/%Y")),
  na = c("", "NA", "-99")
)
```

```r
# the base R way
read.csv("data/trial_export.csv", colClasses = c(id = "character"),
         na.strings = c("", "NA", "-99"))
```


# 5 · Getting data back out {.section-divider}

## What CSV cannot carry {.smaller}

```{r}
#| label: roundtrip
small <- data.frame(mrn = c("0042", "0007"),
                    visit = as.Date(c("2024-01-03", "2024-02-11")),
                    arm = factor(c("Placebo", "Treatment")))
write.csv(small, "data/small.csv", row.names = FALSE)
str(read.csv("data/small.csv"))
```

. . .

::: {.smaller}
Three columns in, three columns wrecked. A CSV holds **text**, and nothing else —
no types, no factor levels, no attributes. `write.csv()` then `read.csv()` is a
round trip through a format that cannot represent what you had.
:::


## `saveRDS()` for R objects {.smaller}

```{r}
#| label: rds
saveRDS(small, "data/small.rds")
str(readRDS("data/small.rds"))
```

. . .

::: {style="font-size:0.84em"}
| | Use for | Keeps types | Readable by |
|---|---|---|---|
| `write.csv()` | handing data to **someone else** | no | everything |
| `saveRDS()` | one object, for **your own next step** | yes | R only |
| `save()` / `load()` | several objects at once | yes | R only — **and it renames your objects** |
:::


## Never save the workspace {.smaller}

::: {.smaller}
- `.RData` restored at startup means your session begins with objects **no code in
  your project created**
- The object in memory and the code on the page drift apart, and nothing warns you
- `save.image()` at the end of a script is the same bug, made deliberate
:::

. . .

::: {.callout-important appearance="simple"}
Tools → Global Options → **Restore .RData: never**, **Save workspace: never**.
:::

. . .

::: {.smaller}
The reproducible alternative, already yours since session 1: **the script is the
state.** Anything you need, you re-create by running code. Anything expensive,
you cache with `saveRDS()` and a named file you can point at.
:::


# 6 · Dates and times {.section-divider}

## A date is a number {.smaller}

```{r}
#| label: date-is-number
d <- as.Date("2024-03-04")
class(d)
as.numeric(d)
d + 84
```

. . .

::: {.smaller}
A `Date` is **days since 1970-01-01**, stored as a number and printed as a date.
That is why `d + 84` works and why `d1 - d2` gives you an interval.
:::

. . .

::: {.callout-note appearance="simple"}
A character column that *looks* like a date supports none of this. `"2024-03-04"
+ 84` is an error, and `"03/04/2024" < "12/31/2023"` is `TRUE`.
:::


## Parsing with `as.Date()` {.smaller}

```{r}
#| label: parse-dates
as.Date("2024-03-04")                              # ISO: no format needed
as.Date("03/04/2024", format = "%m/%d/%Y")
as.Date("04-Mar-2024", format = "%d-%b-%Y")
as.Date("Mar 4, 2024", format = "%b %d, %Y")
```

. . .

::: {style="font-size:0.82em"}
| Code | Means | Matches |
|---|---|---|
| `%Y` / `%y` | year, 4-digit / 2-digit | `2024` / `24` |
| `%m` | month, as a number | `03` |
| `%b` / `%B` | month name, short / full | `Mar` / `March` |
| `%d` | day of month | `04` |
| `%H:%M:%S` | time of day | `14:30:00` |
:::

. . .

`%b` and `%B` are **locale-dependent**. A French export gives you `mars`, and
your parse returns `NA`.


## `3/4/24` {.smaller}

```{r}
#| label: ambiguous
x <- c("3/4/24", "12/31/23", "7/15/24")
as.Date(x, format = "%m/%d/%y")
as.Date(x, format = "%d/%m/%y")
```

. . .

::: {.smaller}
The first value parsed **both ways, silently, into different dates**. The other
two failed loudly under the wrong format — because 31 and 15 cannot be months.
:::

. . .

::: {.callout-important appearance="simple"}
A day-first file read as month-first is wrong on every date where both parts are
≤ 12 — about **a third of the year** — and correct-looking everywhere else.
:::


## Two formats in one column {.smaller}

```{r}
#| label: mixed-formats
v <- read.csv("data/trial_export.csv")$visit_12
d <- as.Date(v, format = "%m/%d/%Y")
sum(is.na(d))
v[is.na(d)]
```

. . .

```{r}
#| label: mixed-fix
d[is.na(d)] <- as.Date(v[is.na(d)], format = "%d-%b-%Y")
sum(is.na(d))
range(d)
```


## Components and arithmetic {.smaller}

```{r}
#| label: date-arith
d0  <- as.Date(read.csv("data/trial_export.csv")$visit_0, format = "%m/%d/%Y")
d12 <- as.Date(read.csv("data/trial_export.csv")$visit_12, format = "%m/%d/%Y")
head(as.numeric(format(d0, "%Y")))                    # year as a number
head(difftime(d12, d0, units = "days"), 3)            # explicit units
head(as.numeric(d12 - d0), 3)                         # plain numbers
summary(as.numeric(d12 - d0))
```


## Excel serial dates {.smaller}

```{r}
#| label: excel-serial
head(read.csv("data/trial_export.csv")$dob_ser, 3)
head(as.Date(read.csv("data/trial_export.csv")$dob_ser,
             origin = "1899-12-30"), 3)
```

. . .

::: {style="font-size:0.84em"}
| Workbook | Serial 1 is | `origin =` |
|---|---|---|
| Windows Excel, and everything modern | 1900-01-01 | `"1899-12-30"` |
| Legacy Mac Excel (pre-2011) | 1904-01-02 | `"1904-01-01"` |
:::

. . .

Use the wrong one and every date lands **four years and a day** out — plausibly,
and with no warning.


## Times, and time zones {.smaller}

```{r}
#| label: posix
t1 <- as.POSIXct("2024-03-04 14:30:00", tz = "America/Chicago")
as.numeric(t1)                                     # seconds since 1970
format(t1, tz = "UTC", usetz = TRUE)
difftime(as.POSIXct("2024-03-04 14:30:00", tz = "UTC"), t1, units = "hours")
```

. . .

::: {.smaller}
`POSIXct` is **seconds** since 1970-01-01 UTC. A time with no `tz` uses your
machine's zone — so the same code gives different answers on different laptops.
:::

::: {.callout-warning appearance="simple"}
Always name the zone. `"America/Chicago"`, not `"CST"`.
:::


## `lubridate`, the friendlier version {.smaller}

```r
library(lubridate)

ymd("2024-03-04");  mdy("03/04/2024");  dmy("4 March 2024")   # named parsers
year(d); month(d); day(d); wday(d, label = TRUE)              # components
d + days(84);  d %m+% months(3)                               # safe arithmetic
interval(dob, visit) / years(1)                               # exact age
with_tz(t1, "UTC");  force_tz(t1, "UTC")                      # two different jobs
```

. . .

```r
# the base R way
as.Date("03/04/2024", format = "%m/%d/%Y")
as.numeric(format(d, "%Y"))
```

. . .

::: {.smaller}
`mdy()` versus `dmy()` puts the ambiguity **in the function name**, where a
reader can see it. That is the real gain, not the typing.
:::


# 7 · Wrap {.section-divider}

## Import problems that do not error {.smaller}

::: {style="font-size:0.78em"}
| Symptom | Cause | Fix |
|---|---|---|
| Data frame has **one column** | Wrong `sep` | `sep = ";"` / `"\t"`, or `read.csv2()` |
| Numeric column arrives as `chr` | One non-numeric value in it | Find it, then `na.strings` or repair |
| Leading zeros gone from an ID | Type guessing made it an integer | `colClasses = c(mrn = "character")` |
| Mean is implausible; `min` is `-99` | Sentinel value | `na.strings = c("NA", "-99")` |
| `table()` shows a nameless category | Empty fields, not `NA` | Add `""` to `na.strings` |
| Column name not found | `check.names` mangled it | `names(df)` right after import |
| Dates all `NA` | `format` does not match the file | `readLines()`, then match it exactly |
| Dates *partly* `NA` | Two formats, or day-first read month-first | Two passes; check `range()` |
| A column of five-digit integers | Excel serial dates | `as.Date(x, origin = "1899-12-30")` |
| Accented names are rubbish | Encoding mismatch | `fileEncoding = "latin1"` |
| First column called `X` | `write.csv()` without `row.names = FALSE` | Re-export |
| Row count is 1 more than expected | A totals row at the foot of the sheet | `nrows =`, or `range =` in `read_excel()` |
:::


## What a defensible import looks like {.smaller}

```r
raw <- read.csv(
  here::here("data", "trial_export.csv"),
  na.strings  = c("NA", "", "-99", "999"),   # per the data dictionary, v2.1
  colClasses  = c(id = "character"),
  strip.white = TRUE
)
raw$visit_0 <- as.Date(raw$visit_0, format = "%m/%d/%Y")

str(raw)
dim(raw)
summary(raw)
table(raw$arm, useNA = "ifany")
colSums(is.na(raw))
range(raw$visit_0)
```

. . .

::: {.smaller}
Ten lines. The comment naming the data-dictionary version is not decoration — it
is the only record of **why** those codes and not others.
:::


## Summary {.smaller}

- Import is a translation governed by explicit assumptions.
- Inspect the file and the returned object before analysis.
- Declare missing-value codes and date formats at the boundary.
- Validate dimensions, keys, types, ranges, categories, and missingness.
- Choose export format according to what must be preserved.

. . .

**Next connection:** Session 6 expresses familiar row, column, transformation, and summary operations with dplyr.

