---
title: "Tidyverse Foundations and the dplyr Verbs"
subtitle: "BIOS 6301 · Introduction to Statistical Computing"
author: "Department of Biostatistics, Vanderbilt University Medical Center"
date: "Session 6"
date-format: "[Session 6]"
format:
  revealjs:
    theme: simple
    css: ../../slides.css
    slide-number: c/t
    show-slide-number: all
    incremental: false
    width: 1280
    height: 720
    fig-align: center
    code-line-numbers: false
    highlight-style: github
    footer: "BIOS 6301 · Session 6 · Tidyverse foundations and the dplyr verbs"
    navigation-mode: linear
execute:
  enabled: true
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(ggplot2)
library(knitr)
theme_set(theme_minimal(base_size = 15))

# Synthetic teaching data for BIOS 6301. Not real patients.
set.seed(6301)
n <- 60
trial <- data.frame(
  id        = sprintf("P%03d", 1:n),
  arm       = rep(c("Placebo", "Treatment"), each = n / 2),
  site      = sample(1:4, n, replace = TRUE),
  age       = round(rnorm(n, 61, 9)),
  sex       = sample(c("F", "M"), n, replace = TRUE),
  weight_kg = round(rnorm(n, 82, 15), 1),
  sbp_0     = round(rnorm(n, 148, 14)),
  stringsAsFactors = FALSE
)
trial$sbp_12 <- round(trial$sbp_0 -
  ifelse(trial$arm == "Treatment", rnorm(n, 14, 8), rnorm(n, 3, 8)))

# --- Session-6 additions, derived from the shared cohort above. -------------
# What I broke, deliberately, and where it is used:
#   sbp_12    three missed 12-week visits -> filter()/NA, the honest n
#   weight_kg two weights never recorded  -> across(..., na.rm = TRUE)
#   enrol     an enrolment log in which one CRF page was keyed twice,
#             so P003 and P004 appear on it twice -> distinct(), count()
trial$sbp_12[c(7, 23, 41)] <- NA
trial$weight_kg[c(12, 50)] <- NA
enrol <- rbind(
  trial[1:6, c("id", "arm", "site")],
  trial[3:4, c("id", "arm", "site")]
)
```

## Where we are {.smaller}

| Course arc | Sessions | |
|---|---:|---|
| Setup and workflow | 0–2 |  |
| **Data foundations** | **3–8** | **← today** |
| Modern tools | 9–10 |  |
| R programming | 11–13 |  |
| Statistical computing | 14–16 |  |

Today's deck develops one part of the current arc; later-session tools are named only when they clarify where this idea will be used.


## What you will be able to do {.smaller}

::: {.smaller}
[1]{.objective-num} **Explain** what `library()` does, and **diagnose** a masked
function in under a minute.

[2]{.objective-num} **Distinguish** a tibble from a `data.frame` and say what
actually differs.

[3]{.objective-num} **Write** pipelines with `|>` — and **read** `%>%`.

[4]{.objective-num} **Use** `select`, `filter`, `arrange`, `mutate`, `if_else`,
`case_when`, `rename`, `distinct`, and `count`.

[5]{.objective-num} **Compute** grouped summaries with an honest *n*, and know
when you still need `ungroup()`.

[6]{.objective-num} **Translate** a base-R data-handling script into dplyr, line
by line.

[7]{.objective-num} **Explain** why `filter(trial, age > 70)` works with no
quotes, and where that convenience bites.

:::


## Roadmap for today

| | Block | Minutes |
|---|---|---|
| 1 | Packages, namespaces, and masking | 15 |
| 2 | Data frames and tibbles | 10 |
| 3 | The pipe, and the design principles | 10 |
| 4 | The verbs: `select` `filter` `arrange` `mutate` `if_else` `case_when` … | 25 |
| 5 | Grouping: `group_by` `summarise` `count` `across` | 20 |
| 6 | Translating session 4, line by line | 5 |
| 7 | Tidy evaluation in ten minutes | 10 |

: {tbl-colwidths="[6,64,30]"}

. . .

Same 60-patient cohort as every other session. Three 12-week visits are missing,
and **that is deliberate**.


# 1 · Packages, namespaces, and masking {.section-divider}

## Installed, loaded, or reached into {.smaller}

::: {style="font-size:0.88em"}
| | Command | How often | Changes |
|---|---|---|---|
| **Install** | `install.packages("dplyr")` | Once per machine | Your hard disk |
| **Load** | `library(dplyr)` | Every session | This R session |
| **Reach in** | `dplyr::n_distinct(x)` | Per call | Nothing |
:::

. . .

```{r}
#| label: search-path
head(search(), 5)
```

::: {.smaller}
`search()` is the list of places R looks up a name, **in order**, from left to
right. `library()` pushes a package onto the front of that list; `::` skips the
list entirely and **documents the dependency** where it is used.
:::


## What masking actually looks like {.smaller}

```{r}
#| label: masking-stats
round(stats::filter(trial$sbp_0[1:8], rep(1/3, 3)), 1)
```

. . .

::: {.smaller}
`stats::filter()` is a **time-series filter** — a moving average. It has nothing
to do with rows.
:::

. . .

```{r}
#| label: masking-error
#| error: true
stats::filter(trial, age > 70)
```


## A tibble *is* a data frame {.smaller}

```{r}
#| label: tibble-class
trial <- tibble::as_tibble(trial)   # from here on, trial is a tibble
class(trial)
trial
```

::: {.smaller}
Three classes, and `data.frame` is the **last** one — so everything from session
4 still applies. What changed is the printing: ten rows, a row count, and the
type of every column for free.
:::


## Three surprises a tibble removes {.smaller}

```{r}
#| label: tibble-nosurprise
#| warning: true
trial$ag                    # session 4: partial matching gave you `age`
class(trial[, "age"])       # session 4: dropped to a vector
```

. . .

::: {style="font-size:0.88em"}
| Session 4 surprise | Tibble |
|---|---|
| `$` partially matches a typo | Warns, returns `NULL` |
| `[` drops one column to a vector | Always a tibble — no `drop = FALSE` needed |
| `stringsAsFactors` history | Never converts anything |
:::


# 2 · The pipe {.section-divider}

## The pipe, mechanically {.smaller}

`x |> f(y)` is **exactly** `f(x, y)`. Nothing more.

```{r}
#| label: pipe-mech
round(mean(trial$sbp_0), 1)
trial$sbp_0 |> mean() |> round(1)
```

. . .

::: {.smaller}
Read `|>` as **"and then"**. It moves the innermost thing in a nested call to the
front, where you read first.
:::


## `%>%` — because you will read it everywhere {.smaller}

```r
# magrittr / tidyverse pipe — needs library(magrittr) or library(dplyr)
trial %>% filter(age > 70) %>% summarise(m = mean(sbp_0))

# native pipe — base R since 4.1, nothing to load
trial |> filter(age > 70) |> summarise(m = mean(sbp_0))
```

. . .

::: {style="font-size:0.86em"}
| | `%>%` | <code>\|></code> |
|---|---|---|
| Comes from | the **magrittr** package | base R, since 4.1 (2021) |
| Placeholder | `.` — anywhere, repeatable | `_` — named argument only |
| Bare function name | `x %>% mean` works | <code>x \|> mean</code> is an error; needs `mean()` |
:::

. . .

::: {.smaller}
**Write `|>`. Read `%>%`.** Ninety percent of the dplyr code written before 2021
uses `%>%`, and that includes most of the internet.
:::


## Four design principles {.smaller}

::: {style="font-size:0.9em"}
| Principle | Consequence |
|---|---|
| **Data comes first** | Every verb pipes |
| **Functions are verbs** | The code says what it does: `filter`, `arrange` |
| **Data frame in, data frame out** | Any verb can follow any verb |
| **Columns are bare names** | No `trial$` on every line |
:::

. . .

::: {.callout-important appearance="simple"}
Consistency is the feature. Learning the fifth verb costs nothing once you know
the first four, because all four rules hold for all of them.
:::


# 3 · The verbs {.section-divider}

## One grammar {.smaller}

```{mermaid}
%%| echo: false
%%| fig-width: 10
flowchart LR
  D["<b>trial</b><br/>60 rows"] --> F["<b>filter()</b><br/>fewer rows"]
  F --> M["<b>mutate()</b><br/>more columns"]
  M --> G["<b>group_by()</b><br/>+ metadata"]
  G --> S["<b>summarise()</b><br/>one row per group"]
  S --> R["<b>result</b><br/>2 rows"]
  style D fill:#f4efe4,stroke:#CFAE70
  style R fill:#CFAE70,stroke:#8a7448,color:#fff
```

::: {style="font-size:0.86em"}
| Verb | Changes | Base-R relative |
|---|---|---|
| `select`, `rename`, `relocate` | which columns, and their names | `df[, cols]`, `names()` |
| `filter`, `distinct`, `slice` | which rows | `df[i, ]`, `subset()`, `duplicated()` |
| `arrange` | row order | `order()` |
| `mutate` | adds columns | `transform()`, `df$new <-` |
| `summarise`, `count` | collapses rows | `aggregate()`, `table()` |
:::


## `select()` {.smaller}

```{r}
#| label: select-basic
trial |> select(id, arm, sbp_0, sbp_12) |> head(3)
```

. . .

```{r}
#| label: select-helpers
trial |> select(id, starts_with("sbp")) |> names()
trial |> select(where(is.numeric)) |> names()
trial |> select(!c(site, sex)) |> names()
```

::: {.callout-note appearance="simple"}
## The base R way
```r
trial[, c("id", "arm", "sbp_0", "sbp_12")]
trial[, grep("^sbp", names(trial))]
```
:::


## `filter()` {.smaller}

```{r}
#| label: filter-basic
trial |> filter(arm == "Treatment", age > 70) |> select(id, age, sbp_12)
```

. . .

::: {.smaller}
Commas mean **and**. Use `|` for or, and `%in%` for a set:
`filter(site %in% c(1, 2))`.
:::

::: {.callout-note appearance="simple"}
## The base R way
```r
subset(trial, arm == "Treatment" & age > 70, select = c(id, age, sbp_12))
trial[which(trial$arm == "Treatment" & trial$age > 70), c("id", "age", "sbp_12")]
```
:::


## `filter()` and missing values {.smaller}

```{r}
#| label: filter-na
nrow(trial |> filter(sbp_12 < 120))
nrow(trial[trial$sbp_12 < 120, ])
```

. . .

::: {.smaller}
Five versus eight. `NA < 120` is `NA` in both dialects — but `filter()` **keeps
only `TRUE`**, while `[` returns a row of `NA`s for every `NA` in the index.
:::

. . .

```{r}
#| label: filter-na-explicit
trial |> filter(is.na(sbp_12)) |> select(id, arm, sbp_0, sbp_12)
```


## `arrange()` {.smaller}

```{r}
#| label: arrange-basic
trial |> arrange(desc(sbp_12)) |> select(id, arm, sbp_12) |> head(3)
```

. . .

```{r}
#| label: arrange-na
trial |> arrange(desc(sbp_12)) |> select(id, sbp_12) |> tail(3)
```

::: {.callout-note appearance="simple"}
## The base R way
```r
trial[order(-trial$sbp_12), c("id", "arm", "sbp_12")]
trial[order(trial$arm, -trial$sbp_0), ]
```
:::


## `mutate()` {.smaller}

```{r}
#| label: mutate-basic
trial <- trial |>
  mutate(
    change    = sbp_12 - sbp_0,
    responder = change < -10
  )
trial |> select(id, sbp_0, sbp_12, change, responder) |> head(3)
```

. . .

```{r}
#| label: mutate-transform-fail
#| error: true
transform(trial, sbp_x2 = sbp_0 * 2, sbp_x3 = sbp_x2 + sbp_0)
```

::: {.callout-note appearance="simple"}
## The base R way
```r
trial$change <- trial$sbp_12 - trial$sbp_0
trial$responder <- trial$change < -10
```
:::


## `if_else()` is strict where `ifelse()` is permissive {.smaller}

```{r}
#| label: if-else-strict
#| error: true
ifelse(c(TRUE, FALSE), 1, "x")
if_else(c(TRUE, FALSE), 1, "x")
```

. . .

```{r}
#| label: if-else-missing
if_else(c(TRUE, FALSE, NA), 1, 0, missing = -1)
```

::: {.callout-note appearance="simple"}
`if_else()` requires compatible types and sizes and gives missing conditions an
explicit result. It stops where `ifelse()` might silently coerce or recycle.
:::


## `case_when()` {.smaller}

```{r}
#| label: case-when
trial |>
  mutate(bp_cat = case_when(
    is.na(sbp_12) ~ "missing",
    sbp_12 <  120 ~ "normal",
    sbp_12 <  140 ~ "elevated",
    .default      = "hypertensive"
  )) |>
  count(bp_cat)
```

. . .

::: {.smaller}
Read top to bottom: **the first `TRUE` wins**. Put `is.na()` first, or the
missing values fall through into whichever category comes last.
:::


## The base R category assignment {.smaller}

```{r}
#| label: case-when-base
bp_cat <- rep("hypertensive", nrow(trial))
bp_cat[trial$sbp_12 < 140] <- "elevated"
bp_cat[trial$sbp_12 < 120] <- "normal"
bp_cat[is.na(trial$sbp_12)] <- "missing"

table(bp_cat)
```

. . .

The base version initializes an output vector and replaces values in sequence.
The order still matters—but here **later assignments overwrite earlier ones**.


## `case_when()` fails loudly {.smaller}

```{r}
#| label: case-when-error
#| error: true
trial |>
  mutate(bp_cat = case_when(
    sbp_12 < 120 ~ "normal",
    sbp_12 < 140 ~ 1,
    .default     = "high"
  ))
```

. . .

::: {.smaller}
A character and a number in the same `case_when()`. In session 3, `ifelse()`
would have **coerced** them and handed you a plausible column.
:::


## `rename()`, and which way round it goes {.smaller}

```{r}
#| label: rename
trial |> rename(sbp_baseline = sbp_0, sbp_week12 = sbp_12) |> names()
```

. . .

::: {.callout-note appearance="simple"}
## The base R way
```r
names(trial)[names(trial) == "sbp_0"] <- "sbp_baseline"
```
:::


## `distinct()` and `count()` {.smaller}

```{r}
#| label: distinct
nrow(enrol)                          # the enrolment log: one page keyed twice
nrow(distinct(enrol))
enrol |> count(id) |> filter(n > 1)
```

. . .

```{r}
#| label: count-two
trial |> count(arm, sex)
```

::: {.callout-note appearance="simple"}
## The base R way
```r
nrow(unique(enrol));  table(enrol$id)[table(enrol$id) > 1]
table(trial$arm, trial$sex)
```
:::


# 4 · Grouping {.section-divider}

## `group_by()` and `summarise()` {.smaller}

```{r}
#| label: group-summarise
trial |>
  group_by(arm) |>
  summarise(
    n        = n(),
    mean_sbp = mean(sbp_12, na.rm = TRUE),
    sd_sbp   = sd(sbp_12, na.rm = TRUE)
  )
```

::: {.callout-note appearance="simple"}
## The base R way
```r
aggregate(sbp_12 ~ arm, data = trial, FUN = mean)
tapply(trial$sbp_12, trial$arm, mean, na.rm = TRUE)
```
:::


## `group_by()` does not compute anything {.smaller}

```{r}
#| label: group-metadata
trial |> group_by(arm) |> select(id, arm, sbp_12)
```

. . .

::: {.smaller}
Same 60 rows, same columns. All that changed is the line reading
**`# Groups: arm [2]`** — a note attached to the data saying how a later verb
should split it.
:::


## `summarise()` peels one layer {.smaller}

```{r}
#| label: summarise-peel
#| message: true
by_arm_sex <- trial |> group_by(arm, sex) |> summarise(n = n())
group_vars(by_arm_sex)
```

. . .

```{r}
#| label: summarise-peel-bug
by_arm_sex |> mutate(pct = round(100 * n / sum(n), 1))
```


## The forgotten `ungroup()` {.smaller}

```{r}
#| label: ungroup
trial |>
  group_by(arm, sex) |>
  summarise(n = n(), .groups = "drop") |>
  mutate(pct = round(100 * n / sum(n), 1))
```

. . .

::: {.callout-important appearance="simple"}
`.groups = "drop"` in `summarise()`, or `ungroup()` on the next line. Pick one
and use it **every time you group by more than one variable**.
:::


## Every grouped summary gets an *n* {.smaller}

```{r}
#| label: honest-n
trial |>
  group_by(arm) |>
  summarise(
    n_rand   = n(),                   # randomised
    n_seen   = sum(!is.na(sbp_12)),   # actually measured at 12 weeks
    mean_sbp = round(mean(sbp_12, na.rm = TRUE), 1),
    .groups  = "drop"
  )
```

. . .

::: {.smaller}
`n()` counts **rows**. `sum(!is.na(x))` counts **observations**. When they differ,
the difference is a finding.
:::


## `across()` — the same thing to many columns {.smaller}

```{r}
#| label: across-basic
trial |>
  group_by(arm) |>
  summarise(across(c(sbp_0, sbp_12, weight_kg),
                   \(x) round(mean(x, na.rm = TRUE), 1)),
            .groups = "drop")
```

. . .

```{r}
#| label: across-where
trial |> summarise(across(where(is.numeric), \(x) sum(is.na(x))))
```


## Grouped `mutate()` — the thing `aggregate()` cannot do {.smaller}

```{r}
#| label: grouped-mutate
trial |>
  group_by(arm) |>
  mutate(sbp_0_centred = round(sbp_0 - mean(sbp_0), 1)) |>
  ungroup() |>
  select(id, arm, sbp_0, sbp_0_centred) |> head(3)
```

. . .

::: {.smaller}
Sixty rows out, not two. The group mean is computed per arm and **recycled back**
across every row in that arm.
:::


## The Rosetta table {.smaller}

::: {style="font-size:0.82em"}
| Task | Session 4, base R | Today, dplyr |
|---|---|---|
| Keep some rows | `subset(trial, age > 70)` | `filter(trial, age > 70)` |
| Keep some columns | `trial[, c("id", "age")]` | `select(trial, id, age)` |
| Both at once | `subset(trial, age > 70, select = c(id, age))` | <code>filter() \|> select()</code> |
| Sort | `trial[order(-trial$sbp_0), ]` | `arrange(trial, desc(sbp_0))` |
| Derive a column | `transform(trial, chg = sbp_12 - sbp_0)` | `mutate(trial, chg = sbp_12 - sbp_0)` |
| Rename | `names(trial)[3] <- "study_site"` | `rename(trial, study_site = site)` |
| Drop duplicates | `unique(enrol)` | `distinct(enrol)` |
| Cross-tabulate | `table(trial$arm, trial$sex)` | `count(trial, arm, sex)` |
| Grouped summary | `aggregate(sbp_12 ~ arm, trial, mean)` | <code>group_by(arm) \|> summarise()</code> |
| Group summary per row | `merge()` it back on | <code>group_by(arm) \|> mutate()</code> |
:::


## Summary {.smaller}

- Packages, namespaces, tibbles, and pipes explain the tidyverse workbench.
- The dplyr verbs express row, column, ordering, derivation, and summary operations.
- Grouped metadata changes the meaning of later verbs.
- `n()` and observed n answer different denominator questions.
- Base R and dplyr can express the same statistical inclusion rule.

. . .

**Next connection:** Session 7 applies both dialects to reshaping and joining multiple tables.

