---
title: "Tidyverse Foundations and the dplyr Verbs"
subtitle: "BIOS 6301 · Introduction to Statistical Computing"
author: "Department of Biostatistics, Vanderbilt University Medical Center"
date: "Session 6"
date-format: "[Session 6]"
format:
  revealjs:
    theme: simple
    css:
      - ../../slides.css
      - slides-session06.css
    slide-number: c/t
    show-slide-number: all
    incremental: false
    width: 1280
    height: 720
    fig-align: center
    code-line-numbers: false
    highlight-style: github
    footer: "BIOS 6301 · Session 6 · Tidyverse foundations and the dplyr verbs"
    navigation-mode: linear
execute:
  enabled: true
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(ggplot2)
library(knitr)
theme_set(theme_minimal(base_size = 15))

# Synthetic teaching data for BIOS 6301. Not real patients.
set.seed(6301)
n <- 60
trial <- data.frame(
  id        = sprintf("P%03d", 1:n),
  arm       = rep(c("Placebo", "Treatment"), each = n / 2),
  site      = sample(1:4, n, replace = TRUE),
  age       = round(rnorm(n, 61, 9)),
  sex       = sample(c("F", "M"), n, replace = TRUE),
  weight_kg = round(rnorm(n, 82, 15), 1),
  sbp_0     = round(rnorm(n, 148, 14)),
  stringsAsFactors = FALSE
)
trial$sbp_12 <- round(trial$sbp_0 -
  ifelse(trial$arm == "Treatment", rnorm(n, 14, 8), rnorm(n, 3, 8)))

# --- Session-6 additions, derived from the shared cohort above. -------------
# What I broke, deliberately, and where it is used:
#   sbp_12    three missed 12-week visits -> filter()/NA, the honest n
#   weight_kg two weights never recorded  -> across(..., na.rm = TRUE)
#   enrol     an enrolment log in which one CRF page was keyed twice,
#             so P003 and P004 appear on it twice -> distinct(), count()
trial$sbp_12[c(7, 23, 41)] <- NA
trial$weight_kg[c(12, 50)] <- NA
enrol <- rbind(
  trial[1:6, c("id", "arm", "site")],
  trial[3:4, c("id", "arm", "site")]
)
```

## Where we are {.smaller}

| Course arc | Sessions | |
|---|---:|---|
| Setup and workflow | 0–2 |  |
| **Data foundations** | **3–8** | **← today** |
| Modern tools | 9–10 |  |
| R programming | 11–13 |  |
| Statistical computing | 14–16 |  |

Today's deck develops one part of the current arc; later-session tools are named only when they clarify where this idea will be used.


## A familiar analysis question {.smaller}

Among patients aged 60 or older with a recorded week-12 blood pressure:

1. Keep the eligible observations.
2. Calculate change from baseline.
3. Find the mean change in each treatment arm.

You already have the base-R tools to do this.


## One answer using base R {.smaller}

```{r}
#| label: opening-base
eligible <- trial[trial$age >= 60 & !is.na(trial$sbp_12), ]
eligible$change <- eligible$sbp_12 - eligible$sbp_0
base_summary <- aggregate(change ~ arm, data = eligible, FUN = mean)
base_summary
```

Keep rows → add a column → summarize by group.


## The same answer using dplyr {.smaller}

```{r}
#| label: opening-dplyr
tidy_summary <- trial |>
  filter(age >= 60, !is.na(sbp_12)) |>
  mutate(change = sbp_12 - sbp_0) |>
  group_by(arm) |>
  summarise(mean_change = mean(change), .groups = "drop")
tidy_summary
```

Read `|>` as “and then.” We will unpack each step today.


## Why develop another set of tools? {.smaller}

Base R already supports data analysis. The tidyverse aims to make common
workflows easier to express and read:

- consistent interfaces across related tasks;
- names that describe the operation;
- small steps that work together;
- shared conventions for tabular data.

These are design goals, not a promise that one style is always best.


## What is the tidyverse? {.smaller}

A coordinated collection of **R packages** for data science.

It is not another programming language, and it does not replace R.

Use base R, tidyverse, or a clear combination. The goal is correct,
understandable analysis.


## History 1 · Plotting and reshaping {.smaller}

In the 2000s, Hadley Wickham developed tools for two practical problems:

- **reshape:** getting data into a useful arrangement;
- **ggplot, then ggplot2:** expressing plots through a grammar of graphics.

So ggplot2 is a major early foundation—but the story also begins with data preparation.


## History 2 · A broader analysis workflow {.smaller}

Other needs led to more packages: text, dates, grouped calculations, and import.

- `plyr` preceded `dplyr` in group-wise data manipulation.
- `reshape` and `reshape2` preceded `tidyr`.
- The collection received the **tidyverse** name in **2016**.

The ecosystem developed gradually; it was not designed all at once.


## History 3 · From individual packages to a community {.smaller}

Hadley Wickham played a central early role. Development expanded through
collaborators, the RStudio/Posit team, and open-source contributors.

The packages share ideas, but do not all have the same authors.

**Why this matters:** familiar conventions can carry from one package to another.


## Package map · The main data workflow {.smaller}

| Job | Package | Course connection |
|---|---|---|
| Read CSV/TSV files | `readr` | Session 5 |
| Work with tabular objects | `tibble` | Today |
| Filter, transform, summarize | `dplyr` | Today |
| Reshape tables | `tidyr` | Session 7 |
| Visualize | `ggplot2` | Session 8 |

Know each package's job; do not memorize its functions yet.


## Package map · Supporting tools {.smaller}

| Job | Package |
|---|---|
| Work with text | `stringr` |
| Work with factors | `forcats` |
| Work with dates and times | `lubridate` |
| Apply operations repeatedly | `purrr` |

Together with the previous slide, these are the nine core packages.

`readxl`, used for Excel in Session 5, is a related package outside the core
set attached by `library(tidyverse)`.


## What we will learn today {.smaller}

- Read tibbles and step-by-step pipelines.
- Use dplyr verbs for rows, columns, and derived variables.
- Summarize within groups and check the number of observations.
- Connect the results to equivalent base-R operations.

Installation and loading are guided steps in the **lab**.


## Roadmap for today {.smaller}

| Minutes | Focus |
|---:|---|
| 0–15 | Motivation, brief history, package map |
| 15–25 | Tibbles and pipes |
| 25–40 | Row/column verbs, selection helpers, intermediate results |
| 40–50 | Derived variables, if_else, and case_when |
| 50–65 | Grouping, summarise, mutate, and ungroup |
| 65–75 | across: reuse one operation |
| 75–82 | Counts and duplicate records |
| 82–90 | Complete example and recap |


# 1 · Tables and pipelines {.section-divider}

## A tibble *is* a data frame {.smaller}

```{r}
#| label: tibble-class
trial <- tibble::as_tibble(trial)   # from here on, trial is a tibble
class(trial)
print(trial, n = 6, width = 80)
```

::: {.smaller}
A tibble inherits from `data.frame`. Many familiar operations still work.
The printout shows dimensions and column types; here we display six rows.
:::


## A small example: table or vector? {.smaller}

```{r}
#| label: tibble-subset-example
plain <- data.frame(age = c(50, 60), arm = c("A", "B"))
tbl <- tibble::as_tibble(plain)
class(plain[, "age"])
class(tbl[, "age"])
tbl[["age"]]
```

A tibble's `[` keeps a table **by default**; `[[` extracts a column vector.


## Exact names reduce accidental matches {.smaller}

```{r}
#| label: tibble-exact-names
#| warning: true
plain$ag
tbl$ag
```

The data frame partially matches `age`. The tibble warns and returns `NULL`.

A tibble combines familiar data-frame storage with different defaults for
printing, subsetting, and name checking—not just shorter printing.


# 2 · The pipe {.section-divider}

## The pipe, mechanically {.smaller}

`x |> f(y)` is **exactly** `f(x, y)`. Nothing more.

```{r}
#| label: pipe-mech
round(mean(trial$sbp_0), 1)
trial$sbp_0 |> mean() |> round(1)
```

. . .

::: {.smaller}
Read `|>` as **"and then"**. It moves the innermost thing in a nested call to the
front, where you read first.
:::


## `%>%` — because you will read it everywhere {.smaller}

```r
# magrittr / tidyverse pipe — needs library(magrittr) or library(dplyr)
trial %>% filter(age > 70) %>% summarise(m = mean(sbp_0))

# native pipe — base R since 4.1, nothing to load
trial |> filter(age > 70) |> summarise(m = mean(sbp_0))
```

. . .

## Two pipe styles {.smaller}

::: {.comparison}
| | `%>%` | <code>\|></code> |
|---|---|---|
| Comes from | the **magrittr** package | base R, since 4.1 (2021) |
| Bare function name | `x %>% mean` works | <code>x \|> mean</code> is an error; needs `mean()` |
:::

. . .

::: {.smaller}
**Write `|>`. Read `%>%`.** Both appear frequently in R examples.
:::


## A consistent pattern for dplyr verbs {.smaller}

- The table is the first argument.
- Function names describe the task.
- Most common verbs return another table.
- Column names can be used directly inside these verbs.

This makes step-by-step pipelines convenient. Still check what each step
does to rows, columns, and grouping.


# 3 · The verbs {.section-divider}

## One grammar {.smaller}

```{mermaid}
%%| echo: false
%%| fig-width: 10
flowchart LR
  D["<b>trial</b><br/>60 rows"] --> F["<b>filter()</b><br/>fewer rows"]
  F --> M["<b>mutate()</b><br/>more columns"]
  M --> G["<b>group_by()</b><br/>+ metadata"]
  G --> S["<b>summarise()</b><br/>one row per group"]
  S --> R["<b>result</b><br/>2 rows"]
  style D fill:#f4efe4,stroke:#CFAE70
  style R fill:#CFAE70,stroke:#8a7448,color:#fff
```

## What each verb changes {.smaller}

::: {.comparison}
| Verb | Changes | Base-R relative |
|---|---|---|
| `select`, `rename`, `relocate` | which columns, and their names | `df[, cols]`, `names()` |
| `filter`, `distinct`, `slice` | which rows | `df[i, ]`, `subset()`, `duplicated()` |
| `arrange` | row order | `order()` |
| `mutate` | adds columns | `transform()`, `df$new <-` |
| `summarise`, `count` | collapses rows | `aggregate()`, `table()` |
:::


## `select()`: choose columns {.smaller}

```{r}
#| label: select-basic
trial |> select(id, arm, sbp_0, sbp_12) |> head(3)
```

Columns appear in the order requested. Rows stay the same.

```r
# Base R: the same column selection
trial[, c("id", "arm", "sbp_0", "sbp_12")]
```

## Select columns by their names {.smaller}

```{r}
#| label: select-name-helpers
trial |> select(starts_with("sbp")) |> names()
trial |> select(ends_with("_12")) |> names()
trial |> select(contains("weight")) |> names()
```

These are **selection helpers**, not separate data-cleaning steps.


## `where()`: choose columns by type {.smaller}

```{r}
#| label: select-where
is.numeric(trial$age)
is.numeric(trial$arm)
trial |> select(where(is.numeric)) |> names()
```

`where(is.numeric)` applies the test to each **column**.

Pass the function `is.numeric`, not an empty call `is.numeric()`.


## `all_of()`: names stored in a vector {.smaller}

```{r}
#| label: select-all-of
wanted <- c("id", "sbp_12")
trial |> select(all_of(wanted)) |> head(3)
```

Use `all_of(wanted)` when **all requested columns are required**.

The character vector contains names; it is not itself a data column.

## `all_of()` versus `any_of()` {.smaller}

```{r}
#| label: select-all-of-error
#| error: true
wanted <- c("id", "not_recorded")
trial |> select(all_of(wanted))
```

```{r}
#| label: select-any-of
trial |> select(any_of(wanted)) |> names()
```

`any_of()` skips absent names. Use it only when their absence is acceptable.


## `filter()` {.smaller}

```{r}
#| label: filter-basic
trial |> filter(arm == "Treatment", age > 70) |> select(id, age, sbp_12)
```

. . .

::: {.smaller}
Commas mean **and**. Use `|` for or, and `%in%` for a set:
`filter(site %in% c(1, 2))`.
:::

::: {.callout-note appearance="simple"}
## The base R way
```r
subset(trial, arm == "Treatment" & age > 70, select = c(id, age, sbp_12))
trial[which(trial$arm == "Treatment" & trial$age > 70), c("id", "age", "sbp_12")]
```
:::


## Row conditions are different from column helpers {.smaller}

```{r}
#| label: filter-between
trial |>
  filter(between(age, 50, 70), site %in% c(1, 2)) |>
  select(id, age, site) |> head(3)
```

- `between(age, 50, 70)`: includes both 50 and 70.
- `site %in% c(1, 2)`: membership in a set.
- `select(where(is.numeric))`: chooses **columns**, not rows.


## `filter()` and missing values {.smaller}

```{r}
#| label: filter-na
nrow(trial |> filter(sbp_12 < 120))
nrow(trial[trial$sbp_12 < 120, ])
```

. . .

::: {.smaller}
Five versus eight. `NA < 120` is `NA` in both dialects — but `filter()` **keeps
only `TRUE`**, while `[` returns a row of `NA`s for every `NA` in the index.
:::

. . .

```{r}
#| label: filter-na-explicit
trial |> filter(is.na(sbp_12)) |> select(id, arm, sbp_0, sbp_12)
```


## `arrange()` {.smaller}

```{r}
#| label: arrange-basic
trial |> arrange(desc(sbp_12)) |>
  select(id, arm, sbp_12) |> head(3)
```

. . .

## Missing values sort last {.smaller}

```{r}
#| label: arrange-na
trial |> arrange(desc(sbp_12)) |> select(id, sbp_12) |> tail(3)
```

::: {.callout-note appearance="simple"}
## The base R way
```r
trial[order(-trial$sbp_12), c("id", "arm", "sbp_12")]
trial[order(trial$arm, -trial$sbp_0), ]
```
:::


## `mutate()` {.smaller}

```{r}
#| label: mutate-basic
trial <- trial |>
  mutate(
    change    = sbp_12 - sbp_0,
    responder = change < -10
  )
trial |> select(id, sbp_0, sbp_12, change, responder) |> head(3)
```

. . .

## Sequential calculations in `mutate()` {.smaller}

```{r}
#| label: mutate-transform-fail
#| error: true
transform(trial, sbp_x2 = sbp_0 * 2, sbp_x3 = sbp_x2 + sbp_0)
```

::: {.callout-note appearance="simple"}
## Sequential assignment in base R
```r
trial$change <- trial$sbp_12 - trial$sbp_0
trial$responder <- trial$change < -10
```
:::


## Follow one pipeline: six patients {.smaller}

```{r}
#| label: small-pipeline-data
demo <- tibble::tibble(
  id = paste0("P", 1:6), arm = rep(c("A", "B"), each = 3),
  age = c(45, 65, 70, 55, 68, 72),
  sbp_0 = c(140, 150, 160, 130, 145, 155),
  sbp_12 = c(135, 140, NA, 125, 135, 140)
)
knitr::kable(demo)
```

Which rows meet `age >= 60` **and** have a recorded follow-up?

## Step 1: keep eligible rows {.smaller}

```{r}
#| label: small-pipeline-filter
step1 <- demo |> filter(age >= 60, !is.na(sbp_12))
knitr::kable(step1)
```

Six rows become three. No columns have been added or removed.


## Step 2: derive a value, then choose columns {.smaller}

```{r}
#| label: small-pipeline-mutate
step2 <- step1 |> mutate(change = sbp_12 - sbp_0)
step3 <- step2 |> select(id, arm, change)
knitr::kable(step3)
```

`mutate()` adds `change`; `select()` keeps three columns.

Inspect intermediate objects when a longer pipeline gives a surprising answer.

## `if_else()`: first, a working example {.smaller}

```{r}
#| label: if-else-working
flag <- c(TRUE, FALSE)
if_else(flag, 1, 0)
```

The arguments are **condition, value if true, value if false**.

Both output branches can be combined into one numeric type.

## Incompatible output types are a separate issue {.smaller}

```{r}
#| label: if-else-strict
#| error: true
ifelse(c(TRUE, FALSE), 1, "x")
if_else(c(TRUE, FALSE), 1, "x")
```

Base `ifelse()` converts to character here; `if_else()` rejects the
numeric/character combination.

Adding an `NA` condition does **not** fix incompatible output types.

## Missing conditions: decide explicitly {.smaller}

```{r}
#| label: if-else-missing
flag <- c(TRUE, FALSE, NA)
if_else(flag, 1, 0)
if_else(flag, "yes", "no", missing = "unknown")
```

By default, an unknown condition gives a missing result.
`missing =` specifies an alternative of a compatible type.


## `case_when()` {.smaller}

```{r}
#| label: case-when
trial |>
  mutate(bp_cat = case_when(
    is.na(sbp_12) ~ "missing",
    sbp_12 <  120 ~ "normal",
    sbp_12 <  140 ~ "elevated",
    .default      = "hypertensive"
  )) |>
  count(bp_cat)
```

. . .

::: {.smaller}
Read top to bottom: **the first `TRUE` wins**. Put `is.na()` first, or the
missing values fall through into whichever category comes last.
:::


## The base R category assignment {.smaller}

```{r}
#| label: case-when-base
bp_cat <- rep("hypertensive", nrow(trial))
bp_cat[trial$sbp_12 < 140] <- "elevated"
bp_cat[trial$sbp_12 < 120] <- "normal"
bp_cat[is.na(trial$sbp_12)] <- "missing"

table(bp_cat)
```

. . .

The base version initializes an output vector and replaces values in sequence.
The order still matters—but here **later assignments overwrite earlier ones**.


## `case_when()` fails loudly {.smaller}

```{r}
#| label: case-when-error
#| error: true
trial |>
  mutate(bp_cat = case_when(
    sbp_12 < 120 ~ "normal",
    sbp_12 < 140 ~ 1,
    .default     = "high"
  ))
```

. . .

::: {.smaller}
A character and a number in the same `case_when()`. In session 3, `ifelse()`
would have **coerced** them and handed you a plausible column.
:::


## `rename()`, and which way round it goes {.smaller}

```{r}
#| label: rename
trial |> rename(sbp_baseline = sbp_0, sbp_week12 = sbp_12) |> names()
```

. . .

::: {.callout-note appearance="simple"}
## The base R way
```r
names(trial)[names(trial) == "sbp_0"] <- "sbp_baseline"
```
:::


## `distinct()` and `count()` {.smaller}

```{r}
#| label: distinct
nrow(enrol)                          # the enrolment log: one page keyed twice
nrow(distinct(enrol))
enrol |> count(id) |> filter(n > 1)
```

## `count()`: frequencies of combinations {.smaller}

```{r}
#| label: count-two
trial |> count(arm, sex)
```

::: {.callout-note appearance="simple"}
## The base R way
```r
nrow(unique(enrol));  table(enrol$id)[table(enrol$id) > 1]
table(trial$arm, trial$sex)
```
:::


# 4 · Grouping {.section-divider}

## `group_by()` and `summarise()` {.smaller}

```{r}
#| label: group-summarise
trial |>
  group_by(arm) |>
  summarise(
    n        = n(),
    mean_sbp = mean(sbp_12, na.rm = TRUE),
    sd_sbp   = sd(sbp_12, na.rm = TRUE)
  )
```

::: {.callout-note appearance="simple"}
## The base R way
```r
aggregate(sbp_12 ~ arm, data = trial, FUN = mean)
tapply(trial$sbp_12, trial$arm, mean, na.rm = TRUE)
```
:::


## `group_by()` does not compute anything {.smaller}

```{r}
#| label: group-metadata
trial |> group_by(arm) |> select(id, arm, sbp_12)
```

. . .

::: {.smaller}
Same 60 rows, same columns. All that changed is the line reading
**`# Groups: arm [2]`** — a note attached to the data saying how a later verb
should split it.
:::


## `summarise()` peels one layer {.smaller}

```{r}
#| label: summarise-peel
#| message: true
by_arm_sex <- trial |> group_by(arm, sex) |> summarise(n = n())
group_vars(by_arm_sex)
```

. . .

```{r}
#| label: summarise-peel-bug
by_arm_sex |> mutate(pct = round(100 * n / sum(n), 1))
```


## The forgotten `ungroup()` {.smaller}

```{r}
#| label: ungroup
trial |>
  group_by(arm, sex) |>
  summarise(n = n(), .groups = "drop") |>
  mutate(pct = round(100 * n / sum(n), 1))
```

. . .

::: {.callout-important appearance="simple"}
`.groups = "drop"` in `summarise()`, or `ungroup()` on the next line. Pick one
and use it **every time you group by more than one variable**.
:::


## Every grouped summary gets an *n* {.smaller}

```{r}
#| label: honest-n
trial |>
  group_by(arm) |>
  summarise(
    n_rand   = n(),                   # randomised
    n_seen   = sum(!is.na(sbp_12)),   # actually measured at 12 weeks
    mean_sbp = round(mean(sbp_12, na.rm = TRUE), 1),
    .groups  = "drop"
  )
```

. . .

::: {.smaller}
`n()` counts **rows**. `sum(!is.na(x))` counts **observations**. When they differ,
the difference is a finding.
:::


## Grouping on six simple measurements {.smaller}

```{r}
#| label: group-small-data
group_demo <- tibble::tibble(
  id = 1:6, arm = rep(c("A", "B"), each = 3),
  value = c(2, 4, 6, 10, 12, 14)
)
grouped <- group_demo |> group_by(arm)
grouped
```

Still six rows. The group means will be 4 and 12.

## `summarise()`: one row per group {.smaller}

```{r}
#| label: group-small-summary
grouped |>
  summarise(n = n(), group_mean = mean(value), .groups = "drop")
```

Six input rows → two summary rows.

## Grouped `mutate()`: keep the original rows {.smaller}

```{r}
#| label: grouped-mutate
grouped |>
  mutate(group_mean = mean(value),
         deviation = value - group_mean) |>
  ungroup()
```

Six input rows → six rows, with calculations performed **within each arm**.


## Base R can also calculate a group mean per row {.smaller}

```{r}
#| label: grouped-base-ave
with(group_demo, ave(value, arm, FUN = mean))
```

`ave()` returns one value per original row.

Different interfaces; the underlying calculation is the same.

## Before `across()`: repeat the same operation {.smaller}

```{r}
#| label: across-manual
demo |> summarise(
  sbp_0 = mean(sbp_0, na.rm = TRUE),
  sbp_12 = mean(sbp_12, na.rm = TRUE)
)
```

Two columns, the same calculation. What if there were twenty?

## `across()`: columns plus an operation {.smaller}

```{r}
#| label: across-simple
demo |> summarise(across(c(sbp_0, sbp_12), mean))
```

Select the columns with the first argument; apply `mean` to each.

The missing follow-up makes its mean `NA`. Next we specify how to handle it.

## Supply arguments through a small function {.smaller}

```{r}
#| label: across-function
demo |> summarise(
  across(c(sbp_0, sbp_12),
         function(x) mean(x, na.rm = TRUE))
)
```

Here `x` is **one selected column at a time**.

`function(x) mean(x, na.rm = TRUE)` is a function without a separate name.

## `\(x)` is shorthand for `function(x)` {.smaller}

```{r}
#| label: across-short-function
demo |> summarise(
  across(c(sbp_0, sbp_12), \(x) mean(x, na.rm = TRUE))
)
```

Same operation, same result. Both are base-R function syntax.

Full function writing comes in Session 11; either spelling is fine here.

## Reuse the selection helpers inside `across()` {.smaller}

```{r}
#| label: across-helpers
demo |> summarise(
  across(starts_with("sbp"), \(x) mean(x, na.rm = TRUE))
)
demo |> summarise(
  across(where(is.numeric), \(x) sum(is.na(x)))
)
```

`all_of(wanted)` also works when column names are stored in a vector.


## Combine grouping and `across()` {.smaller}

```{r}
#| label: across-grouped
demo |> group_by(arm) |>
  summarise(
    across(starts_with("sbp"), \(x) mean(x, na.rm = TRUE)),
    .groups = "drop"
  )
```

The operation runs separately for each selected column **within each group**.

## Same ID does not necessarily mean a duplicate {.smaller}

```{r}
#| label: visits-duplicates
visits <- tibble::tibble(
  id = c("P1", "P1", "P2", "P2"),
  visit = c(0, 12, 0, 0), sbp = c(140, 130, 150, 150)
)
knitr::kable(visits)
```

P1 has two legitimate visits. P2 has one exactly repeated record.

## `unique()` and `distinct()`: define what is repeated {.smaller}

```{r}
#| label: visits-distinct
visits |> distinct()
visits |> distinct(id, .keep_all = TRUE)
```

`unique(visits)` also removes identical rows.
`distinct(id, .keep_all = TRUE)` keeps the **first** row per ID,
discarding P1's follow-up here.


## Finish with one analysis question {.smaller}

Among patients aged 60 or older, report by arm:

- the number eligible;
- the number with observed change;
- the mean observed change.

Use the six-patient `demo` table. Predict the result before running the code.


## A complete pipeline, with visible denominators {.smaller}

```{r}
#| label: final-pipeline
demo |> filter(age >= 60) |>
  mutate(change = sbp_12 - sbp_0) |>
  group_by(arm) |>
  summarise(
    n_eligible = n(),
    n_observed = sum(!is.na(change)),
    mean_change = mean(change, na.rm = TRUE),
    .groups = "drop"
  )
```

Do the counts and the signs match your prediction?

## Base R and dplyr: rows and columns {.smaller}

| Task | Base R | dplyr |
|---|---|---|
| Keep rows | `subset(d, age > 70)` | `filter(d, age > 70)` |
| Keep columns | `d[, c("id", "age")]` | `select(d, id, age)` |
| Sort rows | `d[order(d$age), ]` | `arrange(d, age)` |
| Rename a column | `names(d)[3] <- "site"` | `rename(d, site = old_name)` |

Here `d` stands for a data frame. The rename examples assume the target is
column 3, called `old_name`.


## Base R and dplyr: calculations {.smaller}

| Task | Base R tool | dplyr tool |
|---|---|---|
| Derive a column | `transform()` or `$<-` | `mutate()` |
| Remove identical rows | `unique()` | `distinct()` |
| Count categories | `table()` | `count()` |
| Group summary | `tapply()` or `aggregate()` | `group_by()` + `summarise()` |
| Group value per original row | `ave()` | `group_by()` + `mutate()` |


## Summary {.smaller}

- The tidyverse connects packages through shared conventions.
- Tibbles and pipes support step-by-step data analysis.
- The dplyr verbs express row, column, ordering, derivation, and summary operations.
- Grouped metadata changes the meaning of later verbs.
- `n()` and observed n answer different denominator questions.
- Base R and dplyr can express the same statistical inclusion rule.

. . .

**Next connection:** Session 7 applies both dialects to reshaping and joining multiple tables.

