---
title: "Homework 2: From a raw data file to a reproducible analysis"
subtitle: "Due Wednesday, October 14, 2026, at 11:59 PM Central Time"
author: "REPLACE WITH YOUR FULL NAME"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S %Z')`"
output:
  html_document:
    toc: true
    self_contained: true
---

The full questions and data dictionary are on the Homework 2 page.
Use your existing personal project and keep the original CSV unchanged.
Set `data_file` below to your CSV location; no particular folder layout is required.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = FALSE)
library(dplyr)
data_file <- "data/bp_export.csv"
```

## Part 1: Import and inspect

```{r inspect-file}
# Display the first six lines of data_file.
```

Header line and number of lines to skip: **YOUR ANSWER**

```{r import-checks}
# Import as raw with the correct skip, na.strings, and colClasses.
# Keep id as character.
# Show dim(raw), str(raw), and missingness for the blood-pressure columns.
# Create dat <- raw for the next parts; keep raw unchanged.
```

## Part 2: Factors

```{r factor-order}
# Convert dat$arm to a factor: control first, then treated.
# Create arm_relevelled with treated first; keep dat$arm unchanged.
# Print both level orders and verify identical observed labels.
```

Why is assigning reversed text to levels(dat$arm) different? **YOUR ANSWER**

## Part 3: Prepare and summarize with base R

```{r duplicates}
# Identify the repeated ID with table(), then display its full records.
# After checking agreement across columns, use unique() to create analysis.
# Verify row count and one record per participant.
```

Duplicate decision and unit of observation: **YOUR ANSWER**

```{r base-summary}
# Add analysis$change = follow-up minus baseline.
# Use tapply() or aggregate() to show mean observed change by arm.
# Participant counts will be calculated in Part 4.
```

## Part 4: Work with dplyr

```{r review-table}
# Use select() to create review with id, arm, and change.
# Add observed = !is.na(change) with mutate().
# Display missing-change records with filter(); keep review unchanged.
```

```{r tidy-summary}
# Create tidy_summary: one row per arm, with n_total, n_observed, mean_change.
# Remove grouping explicitly.
```

Comparison with Part 3: **YOUR ANSWER**

## Part 5: Interpret the results

Write two or three sentences, including both arm means, contributing counts,
the direction of change, and the limitations: **YOUR ANSWER**

## Part 6: Reproducibility and Git

This supplied code identifies the exact input file used in this render.

```{r input-record}
data.frame(
  file = data_file,
  md5 = unname(tools::md5sum(data_file))
)
```

What would a changed checksum mean, and what would it not prove? **YOUR ANSWER**

```{r software}
sessionInfo()
```

Before submission: replace the author and answer placeholders, restart R,
knit/render, and verify that source, HTML, and unchanged CSV are pushed to your
existing private course repository's default branch. Make at least two meaningful
commits as you work. Check that the instructor and TA have access.

The latest version pushed to GitHub by **October 14, 2026, at 11:59 PM Central
Time** is the submission. Local commits alone do not count. If submitting late
or requesting that a later revision be graded, notify the TA; the homework's
late-work policy applies.
