---
title: "Joins and Reshaping"
subtitle: "BIOS 6301 · Introduction to Statistical Computing"
author: "Department of Biostatistics, Vanderbilt University Medical Center"
date: "Session 7"
date-format: "[Session 7]"
format:
  revealjs:
    theme: simple
    css: ../../slides.css
    slide-number: c/t
    show-slide-number: all
    incremental: false
    width: 1280
    height: 720
    fig-align: center
    code-line-numbers: false
    highlight-style: github
    footer: "BIOS 6301 · Session 7 · Joins and Reshaping"
    navigation-mode: linear
execute:
  enabled: true
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(ggplot2)
library(knitr)
options(width = 68)
theme_set(theme_minimal(base_size = 15))

# Synthetic teaching data for BIOS 6301. Not real patients.
set.seed(6301)
n <- 60
trial <- data.frame(
  id        = sprintf("P%03d", 1:n),
  arm       = rep(c("Placebo", "Treatment"), each = n / 2),
  site      = sample(1:4, n, replace = TRUE),
  age       = round(rnorm(n, 61, 9)),
  sex       = sample(c("F", "M"), n, replace = TRUE),
  weight_kg = round(rnorm(n, 82, 15), 1),
  sbp_0     = round(rnorm(n, 148, 14)),
  stringsAsFactors = FALSE
)
trial$sbp_12 <- round(trial$sbp_0 -
  ifelse(trial$arm == "Treatment", rnorm(n, 14, 8), rnorm(n, 3, 8)))

# Added for session 7: a 6-week visit, so the repeated-measures reshape has
# three timepoints rather than two, and diastolic pressure, so names_sep has
# something two-part to split.
trial$sbp_6 <- round(trial$sbp_0 -
  ifelse(trial$arm == "Treatment", rnorm(n, 9, 7), rnorm(n, 2, 7)))
set.seed(70301)
trial$dbp_0 <- round(rnorm(n, 88, 9))
trial$dbp_12 <- round(trial$dbp_0 -
  ifelse(trial$arm == "Treatment", rnorm(n, 7, 5), rnorm(n, 1, 5)))
trial <- trial[, c("id", "arm", "site", "age", "sex", "weight_kg",
                   "sbp_0", "sbp_6", "sbp_12", "dbp_0", "dbp_12")]

# A site lookup table, as a study coordinator would send it. Two deliberate
# defects: site 4 is missing (15 patients will not match) and site 5 has no
# patients (a site that opened but never enrolled).
sites <- data.frame(
  site      = c(1, 2, 3, 5),
  site_name = c("VUMC", "Meharry", "Regional", "Satellite"),
  n_beds    = c(1019, 240, 130, 44)
)

# The same table after somebody pasted a row in twice. This is the duplicate
# key that multiplies rows in a join.
sites_dup <- rbind(sites, data.frame(site = 2, site_name = "Meharry",
                                     n_beds = 245))

# A long-format lab table: two rows per patient.
set.seed(6307)
labs <- data.frame(
  id    = rep(trial$id, each = 2),
  week  = rep(c(0, 12), times = n),
  creat = round(rnorm(2 * n, 0.95, 0.22), 2)
)

wide3 <- trial[, c("id", "arm", "sbp_0", "sbp_6", "sbp_12")]
bp_wide <- trial[1:3, c("id", "sbp_0", "sbp_12", "dbp_0", "dbp_12")]
```

## Where we are {.smaller}

| Course arc | Sessions | |
|---|---:|---|
| Setup and workflow | 0–2 |  |
| **Data foundations** | **3–8** | **← today** |
| Modern tools | 9–10 |  |
| R programming | 11–13 |  |
| Statistical computing | 14–16 |  |

Today's deck develops one part of the current arc; later-session tools are named only when they clarify where this idea will be used.


## What you will be able to do {.smaller}

::: {.smaller}
[1]{.objective-num} **Predict** the number of rows a join will return, *before*
you run it.

[2]{.objective-num} **Choose** among `left_join()`, `inner_join()`,
`full_join()`, `semi_join()` and `anti_join()` for a stated question.

[3]{.objective-num} **Detect** duplicate keys and unmatched rows before they
become a wrong answer.

[4]{.objective-num} **Assert** what you believe with `relationship=` and
`unmatched=`.

[5]{.objective-num} **Reshape** between wide and long with `pivot_longer()` and
`pivot_wider()`.

[6]{.objective-num} **Read and write** `merge()`, and **recognise** `reshape()`.

[7]{.objective-num} **Compute** within-patient quantities with `lag()`,
`cumsum()` and `row_number()`.
:::


## Roadmap for today

| | Block | Minutes |
|---|---|---|
| 1 | Data arrives in pieces — keys and grain | 10 |
| 2 | The join family | 20 |
| 3 | **Predicting the row count** | 20 |
| 4 | Safety rails: `relationship=` and `unmatched=` | 10 |
| 5 | The base R way — `merge()`, head to head | 15 |
| 6 | Reshaping: `pivot_longer()`, `pivot_wider()`, `reshape()` | 20 |
| 7 | Window functions, if there is room | 5 |

: {tbl-colwidths="[6,64,30]"}


# 1 · Data arrives in pieces {.section-divider}

## Three tables, one analysis {.smaller}

```{r}
#| label: three-tables
head(trial[, c("id", "arm", "site", "sbp_0", "sbp_12")], 3)
sites
head(labs, 4)
```


## What is one row? {.smaller}

Before you join anything, say out loud what one row of each table is. This is
the table's **grain**.

::: {style="font-size:0.92em"}
| Table | One row is | Key |
|---|---|---|
| `trial` | one **patient** | `id` |
| `sites` | one **site** | `site` |
| `labs` | one **patient-visit** | `id` + `week` |
:::

. . .

A **key** is the set of columns that identifies a row uniquely. If it does not
identify a row uniquely, it is not a key — it is just some columns.

. . .

::: {.callout-important appearance="simple"}
Joining changes the grain unless you can prove it does not. Say what the grain
of the *result* should be before you type the join.
:::


## Check the keys first {.smaller}

The check takes one line and it goes **before** the join, not after:

```{r}
#| label: dup-check
sites_dup |> count(site) |> filter(n > 1)
```

. . .

```{r}
#| label: dup-check-2
trial |> count(id) |> filter(n > 1)
```

An empty result is the answer you want: **`id` is a key of `trial`**.

. . .

::: {.callout-note appearance="simple"}
`count(key) |> filter(n > 1)` is the single highest-value habit in this session.
Run it on both tables, every time, before you join.
:::


# 2 · The join family {.section-divider}

## Six joins, two kinds {.smaller}

::: {style="font-size:0.88em"}
| | Join | Keeps | Adds columns from `y`? |
|---|---|---|---|
| **Mutating** | `left_join(x, y)` | every row of `x` | yes |
| | `inner_join(x, y)` | rows matched in **both** | yes |
| | `right_join(x, y)` | every row of `y` | yes |
| | `full_join(x, y)` | every row of **either** | yes |
| **Filtering** | `semi_join(x, y)` | rows of `x` **with** a match | **no** |
| | `anti_join(x, y)` | rows of `x` **without** a match | **no** |
:::

. . .

The split is the useful thing here. A **mutating** join changes the columns and
may change the rows. A **filtering** join changes only the rows — it is
`filter()` with the condition stored in another table.


## Which join do you want? {.smaller}

```{mermaid}
%%| echo: false
%%| fig-width: 8.6
flowchart TD
  A{"Do you need<br/><b>columns</b> from y?"}
  A -- "no, just to filter rows" --> B{"Keep rows that<br/>match, or that don't?"}
  B -- match --> SEMI["<b>semi_join</b>"]
  B -- "don't match" --> ANTI["<b>anti_join</b><br/><i>your diagnostic</i>"]
  A -- yes --> C{"Which rows<br/>must survive?"}
  C -- "all of x" --> L["<b>left_join</b><br/><i>the default choice</i>"]
  C -- "only matched" --> I["<b>inner_join</b>"]
  C -- "all of both" --> F["<b>full_join</b>"]
  style L fill:#CFAE70,stroke:#8a7448,color:#fff
  style ANTI fill:#CFAE70,stroke:#8a7448,color:#fff
```


## `left_join()`: keep every row of `x` {.smaller}

```{r}
#| label: left-join
trial_sites <- left_join(trial, sites, by = "site")

nrow(trial)
nrow(trial_sites)
head(trial_sites[, c("id", "arm", "site", "site_name", "n_beds")], 5)
```

::: {.callout-note appearance="simple"}
**The base R way**

```r
merge(trial, sites, by = "site", all.x = TRUE)
```
:::


## `inner_join()`: rows that vanish {.smaller}

```{r}
#| label: inner-join
nrow(inner_join(trial, sites, by = "site"))
```

. . .

Fifteen patients are gone, and **the result does not say so**.

```{r}
#| label: inner-join-arm
inner_join(trial, sites, by = "site") |>
  count(arm)
```

. . .

::: {.callout-warning appearance="simple"}
An `inner_join()` is a silent exclusion criterion. If the rows it drops are not
missing at random — and site is *never* missing at random — your analysis
population has changed and your table 1 will not match your consort diagram.
:::


## `anti_join()`: the diagnostic {.smaller}

Which patients failed to match?

```{r}
#| label: anti-join
anti_join(trial, sites, by = "site") |>
  count(site)
```

. . .

And in the other direction — which sites have no patients?

```{r}
#| label: anti-join-rev
anti_join(sites, trial, by = "site")
```

. . .

::: {.callout-tip appearance="simple"}
Run `anti_join()` **both ways** after every join you care about. It costs two
lines and it is the difference between knowing your data and hoping.
:::


## The other three {.smaller}

```{r}
#| label: other-joins
c(left  = nrow(left_join(trial, sites, by = "site")),
  inner = nrow(inner_join(trial, sites, by = "site")),
  right = nrow(right_join(trial, sites, by = "site")),
  full  = nrow(full_join(trial, sites, by = "site")),
  semi  = nrow(semi_join(trial, sites, by = "site")),
  anti  = nrow(anti_join(trial, sites, by = "site")))
```

. . .

::: {.smaller}
- `full_join` gives **61** — the 60 patients plus one all-`NA` patient row
  carrying site 5.
- `right_join(x, y)` is `left_join(y, x)` with the columns in a different order.
  **Use `left_join` and swap the arguments** — then the table your analysis is
  about is always on the left.
- `semi_join` gives 45 rows and **the same columns as `trial`**. Compare with
  `inner_join`, which gives 45 rows and three more columns.
:::


# 3 · Predicting the row count {.section-divider}

## The only rule you need {.smaller}

For each row of `x`, count how many rows of `y` share its key. **Add those
counts up.** That is the number of rows a mutating join returns.

. . .

::: {style="font-size:0.9em"}
| Multiplicity in `y` | Rows out (`left_join`) | Name |
|---|---|---|
| exactly one for every `x` row | `nrow(x)` | one-to-one |
| at most one, sometimes none | `nrow(x)` | **many-to-one** |
| several for some `x` rows | **more** than `nrow(x)` | one-to-many |
| several on both sides | **many** more | many-to-many |
:::

. . .

::: {.callout-important appearance="simple"}
A `left_join()` **never loses** rows. It very much can gain them. Those are two
different facts and students conflate them constantly.
:::


## Many-to-one: the safe case {.smaller}

`sites` has **at most one** row per `site`, so 60 rows in, 60 rows out.

```{r}
#| label: m2o
nrow(trial)
nrow(left_join(trial, sites, by = "site"))
```

. . .

We are entitled to expect that only because we checked:

```{r}
#| label: m2o-check
sites |> count(site) |> filter(n > 1)
```

. . .

::: {.callout-note appearance="simple"}
Most joins you write should be many-to-one: many patients, one lookup row each.
When a join is *supposed* to be many-to-one and is not, you have found a data
error — not a join problem.
:::


## One-to-many: rows multiply on purpose {.smaller}

`labs` has **two** rows per patient — baseline and Session 12.

```{r}
#| label: o2m
nrow(labs)
trial_labs <- left_join(trial, labs, by = "id")
nrow(trial_labs)
```

. . .

`60 × 2 = 120`, predicted before running it.

```{r}
#| label: o2m-head
head(trial_labs[, c("id", "arm", "week", "creat")], 4)
```

. . .

::: {.smaller}
This is correct and often exactly what you want — the result's grain is now
**one row per patient-visit**. But every patient-level column is now
**repeated**, so `mean(trial_labs$age)` is a weighted mean, not a patient mean.
:::


## Many-to-many: rows multiply by accident {.smaller}

`sites_dup` is `sites` with **one row pasted in twice**.

```{r}
#| label: m2m
#| warning: true
nrow(left_join(trial, sites_dup, by = "site"))
```

. . .

Site 2 has **18** patients. `60 + 18 = 78`.

. . .

::: {.callout-important appearance="simple"}
Eighteen patients are now in the data twice. Every count, every mean, every
model is wrong — and the only visible symptom is a row count that nobody looked
at.
:::


## Predict, then check {.smaller}

Write the prediction into the code. It costs one line:

```{r}
#| label: predict
#| error: true
joined <- left_join(trial, sites, by = "site")
stopifnot(nrow(joined) == nrow(trial))
```

. . .

```{r}
#| label: predict-fail
#| error: true
joined_bad <- left_join(trial, sites_dup, by = "site")
stopifnot(nrow(joined_bad) == nrow(trial))
```

. . .

::: {.smaller}
`stopifnot()` is session 12 material — borrowed here as a one-liner. It does
nothing when the condition holds and stops the render when it does not, which is
exactly what you want from an assumption.
:::


# 4 · Safety rails {.section-divider}

## The `relationship` argument {.smaller}

State what you believe, and let the join enforce it:

```{r}
#| label: relationship
#| error: true
left_join(trial, sites_dup, by = "site", relationship = "many-to-one")
```

. . .

::: {.smaller}
Four values: `"one-to-one"`, `"one-to-many"`, `"many-to-one"`,
`"many-to-many"`. The last one is not a safety rail — it is how you tell dplyr
you meant it, and it silences the warning.
:::


## The `unmatched` argument {.smaller}

```{r}
#| label: unmatched
#| error: true
left_join(trial, sites, by = "site", unmatched = "error")
```

. . .

::: {.smaller}
Read the message carefully. For a `left_join()`, rows of `x` are **never**
dropped — so `unmatched=` guards the only side that can lose rows: **`y`**. It
just told us site 5 is unused.
:::

. . .

To assert that every *patient* matched, use the diagnostic:

```{r}
#| label: unmatched-anti
nrow(anti_join(trial, sites, by = "site")) == 0
```


## The join checklist {.smaller}

Five lines. Every join that matters.

```r
# 1. What is the grain of each table, and of the result?
# 2. Duplicate keys?
sites |> count(site) |> filter(n > 1)
# 3. Predict the row count. Write the number down.
# 4. Join, asserting the relationship.
joined <- left_join(trial, sites, by = "site",
                    relationship = "many-to-one")
# 5. What failed to match, both ways?
anti_join(trial, sites, by = "site")
anti_join(sites, trial, by = "site")
```

. . .

::: {.callout-note appearance="simple"}
Steps 2 and 5 belong in your document, not your console. They are part of the
analysis: they are how a reader knows the join was safe.
:::


# 5 · Head to head: `merge()` {.section-divider}

## `merge()` anatomy {.smaller}

```r
merge(x, y,
      by      = NULL,       # key columns, in BOTH tables
      by.x    = ,  by.y = , # when the key has different names
      all.x   = FALSE,      # keep unmatched rows of x?
      all.y   = FALSE,      # keep unmatched rows of y?
      all     = FALSE,      # shorthand for both
      suffixes = c(".x", ".y"),
      sort    = TRUE)       # sorts the result by the key
```

. . .

::: {.smaller}
Everything a join can do is a combination of `all.x` and `all.y`. There is no
`how = "left"` argument; the two logicals *are* the join type.
:::


## The join family in both dialects {.smaller}

::: {style="font-size:0.86em"}
| dplyr | `merge()` | Rows |
|---|---|---|
| `inner_join(x, y, by = "k")` | `merge(x, y, by = "k")` | 45 |
| `left_join(x, y, by = "k")` | `merge(x, y, by = "k", all.x = TRUE)` | 60 |
| `right_join(x, y, by = "k")` | `merge(x, y, by = "k", all.y = TRUE)` | 46 |
| `full_join(x, y, by = "k")` | `merge(x, y, by = "k", all = TRUE)` | 61 |
| `semi_join(x, y, by = "k")` | `x[x$k %in% y$k, ]` | 45 |
| `anti_join(x, y, by = "k")` | `x[!(x$k %in% y$k), ]` | 15 |
:::

. . .

```{r}
#| label: merge-counts
c(inner = nrow(merge(trial, sites, by = "site")),
  left  = nrow(merge(trial, sites, by = "site", all.x = TRUE)),
  right = nrow(merge(trial, sites, by = "site", all.y = TRUE)),
  full  = nrow(merge(trial, sites, by = "site", all   = TRUE)))
```


## The same job, twice {.smaller}

**Task.** Attach site names, keep only patients at a known site, and report mean
12-week SBP by site.

::: columns
::: {.column width="49%"}
```{r}
#| label: hh-tidy
trial |>
  inner_join(sites, by = "site") |>
  group_by(site_name) |>
  summarise(
    n    = n(),
    mean = round(mean(sbp_12), 1)
  )
```
:::

::: {.column width="49%"}
```{r}
#| label: hh-base
m <- merge(trial, sites,
           by = "site")
aggregate(sbp_12 ~ site_name,
          data = m,
          FUN  = mean)
```
:::
:::


## `merge()` gotchas {.smaller}

::: {style="font-size:0.88em"}
| Behaviour | Consequence |
|---|---|
| `sort = TRUE` by default | The result is **reordered by the key**. dplyr preserves `x`'s order |
| `by = NULL` by default | Joins on *all* common column names — or gives a **cross join** if there are none |
| No warning on duplicate keys | 78 rows, silently. dplyr 1.1 warns |
| `suffixes = c(".x", ".y")` | Same-named non-key columns collide into `age.x` / `age.y` |
:::

. . .

```{r}
#| label: merge-cross
nrow(merge(trial[1:3, c("id", "arm")], sites))   # no common column
```

Three rows and four rows gave **twelve**. That is every pairing.


# 6 · Reshaping {.section-divider}

## The same data, two shapes {.smaller}

::: columns
::: {.column width="47%"}
**Wide** — one row per patient

```{r}
#| label: wide-shape
#| echo: false
head(wide3, 3)
```

One column per timepoint. Good for reading, for table 1, for a paired *t*-test.
:::

::: {.column width="47%"}
**Long** — one row per patient-visit

```{r}
#| label: long-shape
#| echo: false
sbp_long <- reshape(wide3,
  direction = "long", idvar = "id",
  varying = list(c("sbp_0", "sbp_6", "sbp_12")),
  v.names = "sbp", timevar = "week",
  times = c(0, 6, 12))
sbp_long <- sbp_long[order(sbp_long$id, sbp_long$week), ]
rownames(sbp_long) <- NULL
head(sbp_long, 6)
```

Timepoint is a **value in a column**. Good for plotting, grouping, and mixed
models.
:::
:::

. . .

Same 180 numbers. `60 × 3` cells becomes `180 × 1` column.


## Which shape do you need? {.smaller}

::: {style="font-size:0.9em"}
| You want to | Shape |
|---|---|
| Read the data as a human | **wide** |
| A baseline characteristics table | **wide** |
| `sbp_12 - sbp_0` | **wide** |
| Plot outcome against time | **long** |
| A summary at each timepoint | **long** |
| Fit a mixed model or GEE | **long** |
| Send it to a clinical collaborator | **wide**, always |
:::

. . .

::: {.callout-note appearance="simple"}
Reshaping is not a step in a pipeline you do once. It is a thing you do **back
and forth**, several times, in a single analysis.
:::


## Wide to long: `pivot_longer()` {.smaller}

```r
library(tidyr)

sbp_long <- wide3 |>
  pivot_longer(
    cols             = c(sbp_0, sbp_6, sbp_12),
    names_to         = "week",
    names_prefix     = "sbp_",
    values_to        = "sbp",
    names_transform  = list(week = as.integer)
  )
```

. . .

```
# A tibble: 180 × 4
   id    arm      week   sbp
   <chr> <chr>   <int> <dbl>
 1 P001  Placebo     0   160
 2 P001  Placebo     6   146
 3 P001  Placebo    12   144
 4 P002  Placebo     0   162
 5 P002  Placebo     6   165
 6 P002  Placebo    12   156
 7 P003  Placebo     0   157
 8 P003  Placebo     6   151
 9 P003  Placebo    12   156
10 P004  Placebo     0   157
# ℹ 170 more rows
```

::: {.dim style="font-size:0.72em"}
`tidyr` is not installed on the teaching machine, so this block is not executed.
The output is the real thing — base `reshape()`, three slides on, builds the
identical table and *is* executed.
:::


## `names_sep`: two facts in one name {.smaller}

```{r}
#| label: bp-wide
#| echo: false
bp_wide
```

`sbp_12` encodes **two** things: which measurement, and which week.

. . .

```r
bp_wide |>
  pivot_longer(
    cols      = -id,
    names_to  = c("measure", "week"),
    names_sep = "_"
  )
```

```
# A tibble: 12 × 4
   id    measure week  value
   <chr> <chr>   <chr> <dbl>
 1 P001  sbp     0       160
 2 P001  sbp     12      144
 3 P001  dbp     0        98
 4 P001  dbp     12       92
 5 P002  sbp     0       162
 6 P002  sbp     12      156
 7 P002  dbp     0        78
 8 P002  dbp     12       74
```

::: {.dim style="font-size:0.72em"}
Rows 1–8 of 12. Not executed here, for the same reason as the previous slide.
:::


## `pivot_wider()` and the round trip {.smaller}

```r
sbp_long |>
  pivot_wider(
    names_from   = week,
    values_from  = sbp,
    names_prefix = "sbp_"
  )
```

. . .

::: {.smaller}
`pivot_wider()` is the inverse. It needs to know which column supplies the new
**names** and which supplies the new **values**; everything else becomes the
identifier.
:::

. . .

::: {.callout-warning appearance="simple"}
If the id columns plus `names_from` do not identify a row uniquely,
`pivot_wider()` cannot put one value in one cell — so it puts a **list** in the
cell and warns. That warning means you have duplicate measurements, which is a
data question, not a code question.
:::


## Base `reshape()`: recognition only {.smaller}

```{r}
#| label: base-reshape
sbp_long <- reshape(wide3,
  direction = "long",
  idvar     = "id",
  varying   = list(c("sbp_0", "sbp_6", "sbp_12")),
  v.names   = "sbp",
  timevar   = "week",
  times     = c(0, 6, 12))
dim(sbp_long)
```

. . .

```{r}
#| label: base-reshape-back
back <- reshape(sbp_long, direction = "wide", idvar = "id",
                timevar = "week", v.names = "sbp", sep = "_")
dim(back)
```

. . .

::: {.callout-important appearance="simple"}
`reshape()` works, and it round-trips exactly. The `idvar` / `timevar` /
`varying` interface is genuinely punishing, it is easy to get subtly wrong, and
**`pivot_longer()` is the tool you will actually use.** Learn to recognise this
call. Do not learn to write it.
:::


## Why long: the plot {.smaller}

```{r}
#| label: fig-spaghetti
#| fig-cap: "Mean systolic BP by visit and arm. Long format, three lines of ggplot2."
#| fig-width: 8
#| fig-height: 3.0
ggplot(sbp_long, aes(week, sbp, colour = arm)) +
  stat_summary(fun = mean, geom = "line", linewidth = 1.1) +
  labs(x = "Week", y = "Mean SBP (mmHg)", colour = NULL)
```

::: {.smaller}
Borrowed from session 1's minimum-viable plot — the grammar is session 8.
:::


# 7 · Window functions {.section-divider}

## Within patient, in order {.smaller}

`lag()` looks back one row. `group_by()` says *back within what*.

```{r}
#| label: lag-demo
visits <- sbp_long |>
  arrange(id, week) |>
  group_by(id) |>
  mutate(
    prev_sbp = lag(sbp),
    change   = sbp - lag(sbp)
  ) |>
  ungroup()

head(as.data.frame(visits), 6)
```


## Numbering, looking forward, and baselines {.smaller}

```{r}
#| label: window-more
sbp_long |>
  arrange(id, week) |>
  group_by(id) |>
  mutate(
    visit      = row_number(),
    next_sbp   = lead(sbp),
    from_base  = sbp - first(sbp)
  ) |>
  ungroup() |>
  head(6) |>
  as.data.frame()
```

::: {.smaller}
`row_number()` numbers rows within group · `lead()` looks forward · `first()`
takes the group's first value · `cumsum()` accumulates.
:::


## Summary {.smaller}

- State the grain and keys of every table before joining.
- Predict row counts and detect duplicate keys before the join.
- Use anti-joins to inspect unmatched records in both directions.
- Wide and long forms encode different observation units.
- Order explicitly before lagged or cumulative operations.

. . .

**Next connection:** Session 8 uses the long, validated analysis table to build statistical figures.

