---
title: "Joining, Reshaping, and Cleaning Data"
subtitle: "BIOS 6301 · Introduction to Statistical Computing"
author: "Department of Biostatistics, Vanderbilt University Medical Center"
date: "Session 7"
date-format: "[Session 7]"
format:
  revealjs:
    theme: simple
    css:
      - ../../slides.css
      - slides-session07.css
    slide-number: c/t
    show-slide-number: all
    incremental: false
    width: 1280
    height: 720
    fig-align: center
    code-line-numbers: false
    highlight-style: github
    footer: "BIOS 6301 · Session 7 · Tidyverse Data Tools"
    navigation-mode: linear
execute:
  enabled: true
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(tidyr)
library(stringr)
library(forcats)
library(lubridate)
library(ggplot2)
library(knitr)
options(width = 68)
theme_set(theme_minimal(base_size = 15))

# Synthetic teaching data for BIOS 6301. Not real patients.
set.seed(6301)
n <- 60
trial <- data.frame(
  id        = sprintf("P%03d", 1:n),
  arm       = rep(c("Placebo", "Treatment"), each = n / 2),
  site      = sample(1:4, n, replace = TRUE),
  age       = round(rnorm(n, 61, 9)),
  sex       = sample(c("F", "M"), n, replace = TRUE),
  weight_kg = round(rnorm(n, 82, 15), 1),
  sbp_0     = round(rnorm(n, 148, 14))
)
trial$sbp_12 <- round(trial$sbp_0 -
  ifelse(trial$arm == "Treatment", rnorm(n, 14, 8), rnorm(n, 3, 8)))

# Added for session 7: a 6-week visit, so the repeated-measures reshape has
# three timepoints rather than two, and diastolic pressure, so names_sep has
# something two-part to split.
trial$sbp_6 <- round(trial$sbp_0 -
  ifelse(trial$arm == "Treatment", rnorm(n, 9, 7), rnorm(n, 2, 7)))
set.seed(70301)
trial$dbp_0 <- round(rnorm(n, 88, 9))
trial$dbp_12 <- round(trial$dbp_0 -
  ifelse(trial$arm == "Treatment", rnorm(n, 7, 5), rnorm(n, 1, 5)))
trial <- trial[, c("id", "arm", "site", "age", "sex", "weight_kg",
                   "sbp_0", "sbp_6", "sbp_12", "dbp_0", "dbp_12")]

# A site lookup table, as a study coordinator would send it. Two deliberate
# defects: site 4 is missing (15 patients will not match) and site 5 has no
# patients (a site that opened but never enrolled).
sites <- data.frame(
  site      = c(1, 2, 3, 5),
  site_name = c("VUMC", "Meharry", "Regional", "Satellite"),
  n_beds    = c(1019, 240, 130, 44)
)

# The same table after somebody pasted a row in twice. This is the duplicate
# key that multiplies rows in a join.
sites_dup <- rbind(sites, data.frame(site = 2, site_name = "Meharry",
                                     n_beds = 245))

# A long-format lab table: two rows per patient.
set.seed(6307)
labs <- data.frame(
  id    = rep(trial$id, each = 2),
  week  = rep(c(0, 12), times = n),
  creat = round(rnorm(2 * n, 0.95, 0.22), 2)
)

wide3 <- trial[, c("id", "arm", "sbp_0", "sbp_6", "sbp_12")]
bp_wide <- trial[1:3, c("id", "sbp_0", "sbp_12", "dbp_0", "dbp_12")]
```


```{r}
#| label: core-data
#| include: false
# Small synthetic cohort used throughout the main lesson and lab.
patient_lookup <- tibble(
  id = c("P01", "P02", "P03", "P04"),
  arm = c("control", "treated", "treated", "control")
)
marker_wide <- tibble(
  id = c("P01", "P02", "P03", "P05"),
  marker_w0 = c(4.1, 5.2, 6.0, 4.8),
  marker_w4 = c(4.4, 4.7, 5.1, 4.9)
)
raw_patients <- tibble(
  id = c(" p01 ", "p02", " P03", "p04 "),
  arm = c(" Control ", "TREATED", " treated ", "control"),
  enrolled = c("2026-09-01", "2026-09-03", "2026-10-01", "2026-10-02"),
  note = c(" first   visit ", " repeat sample", "repeat  visit ", "first visit")
)
```

## Session 6 skills we will reuse {.smaller}

- `filter()` and `select()` choose rows and columns.
- `mutate()` creates or changes columns.
- `count()` and `summarise()` describe groups.
- Pipes connect those steps.

Today we combine tables, change their shape, and clean particular column types.


## Packages and their jobs {.smaller}

| Package | Today's job |
|---|---|
| dplyr | Join related tables and check the result |
| tidyr | Reshape measurements between wide and long |
| stringr | Clean and search text |
| forcats | Change factor labels and level order |
| lubridate | Parse dates and extract date components |

These packages work together inside the same `mutate()` and pipe workflows.

## The main teaching route {.smaller}

| Minutes | Topic |
|---:|---|
| 0–5 | Recap and package roles |
| 5–30 | Basic joins and checks |
| 30–50 | tidyr and basic pivots |
| 50–65 | stringr |
| 65–75 | forcats |
| 75–85 | lubridate |
| 85–90 | Connected example and summary |

The optional join block and final appendix sit outside this 90-minute route.


# Joining tables {.section-divider}

## Two tables from the same study {.smaller}

```{r}
#| label: core-input-tables
#| echo: false
patient_lookup
marker_wide
```

Each table has one row per participant, but their participant lists differ.


## Join keys {.smaller}

The `by` argument tells R which values identify a match.

```{r}
#| label: core-key-check
patient_lookup |> count(id) |> filter(n > 1)
marker_wide |> count(id) |> filter(n > 1)
```

Zero rows here means no repeated IDs in that table.

A visit-level table can contain repeated IDs legitimately. Its row identifier
would also need a visit variable.

## Three joins for today's work {.smaller}

| Question | Function |
|---|---|
| Keep all measurement rows and add metadata | `left_join()` |
| Keep only participants appearing in both tables | `inner_join()` |
| Find measurement rows without metadata | `anti_join()` |

`left_join(x, y, by = "id")` starts with `x` and matches its IDs against `y`.


## A left join, row by row {.smaller}

```{r}
#| label: core-left-join
course_joined <- marker_wide |>
  left_join(patient_lookup, by = "id")
course_joined
```

P05 stays. Its `arm` is `NA` because the metadata table has no matching ID.

## An inner join changes the retained participants {.smaller}

```{r}
#| label: core-inner-join
marker_wide |> inner_join(patient_lookup, by = "id")
```

Three participants remain. P05 is absent.

Choose an inner join when restricting to matched records is the intended rule.

## Unmatched records in both directions {.smaller}

```{r}
#| label: core-anti-joins
marker_wide |> anti_join(patient_lookup, by = "id")
patient_lookup |> anti_join(marker_wide, by = "id")
```

The first result asks who lacks metadata.
The second asks whose metadata has no measurements.


## Duplicate lookup rows can repeat measurements {.smaller}

```{r}
#| label: core-duplicate-lookup
extra_lookup <- tibble(
  id = c("P01", "P01"), arm = c("control", "treated")
)
marker_wide |> filter(id == "P01") |>
  left_join(extra_lookup, by = "id")
```

Which treatment label is correct? Resolve that from the data source.
Keeping the first row would make an unsupported choice.


## A short join checklist {.smaller}

- What does one row represent in each input?
- Which columns identify the intended match?
- Does the lookup have repeated keys?
- How many rows remain, and which IDs failed to match?

```{r}
#| label: core-row-count
c(before = nrow(marker_wide), after = nrow(course_joined))
```

Here the lookup ID is unique, so four measurement rows stay four.

# Optional join details {.section-divider #optional-join-details}

[Skip optional joins and continue to tidyr](#tidyr-and-data-shape)

## Optional examples: a larger synthetic study {.smaller}

The next slides use a separate synthetic dataset:

- `trial`: 60 participants, one row per participant.
- `sites`: one row per site, missing site 4 and including unused site 5.
- `labs`: two visits per participant.
- `sites_dup`: a conflicting extra record for site 2.

These examples are for reference. The core lab does not require them.

[Continue to tidyr](#tidyr-and-data-shape)

## Optional: Six joins, two kinds {.smaller}

::: {.comparison}
| | Join | Keeps | Adds columns from `y`? |
|---|---|---|---|
| **Mutating** | `left_join(x, y)` | every row of `x` | yes |
| | `inner_join(x, y)` | rows matched in **both** | yes |
| | `right_join(x, y)` | every row of `y` | yes |
| | `full_join(x, y)` | every row of **either** | yes |
| **Filtering** | `semi_join(x, y)` | rows of `x` **with** a match | **no** |
| | `anti_join(x, y)` | rows of `x` **without** a match | **no** |
:::

. . .

The split is the useful thing here. A **mutating** join changes the columns and
may change the rows. A **filtering** join changes only the rows — it is
`filter()` with the condition stored in another table.


## Optional: The other three {.smaller}

```{r}
#| label: other-joins
c(left  = nrow(left_join(trial, sites, by = "site")),
  inner = nrow(inner_join(trial, sites, by = "site")),
  right = nrow(right_join(trial, sites, by = "site")),
  full  = nrow(full_join(trial, sites, by = "site")),
  semi  = nrow(semi_join(trial, sites, by = "site")),
  anti  = nrow(anti_join(trial, sites, by = "site")))
```

. . .

::: {.smaller}
- `full_join` gives **61** — the 60 patients plus one all-`NA` patient row
  carrying site 5.
- `right_join(x, y)` is `left_join(y, x)` with the columns in a different order.
  **Use `left_join` and swap the arguments** — then the table your analysis is
  about is always on the left.
- `semi_join` gives 45 rows and **the same columns as `trial`**. Compare with
  `inner_join`, which gives 45 rows and two more columns.
:::


## Optional: A small join you can inspect row by row {.smaller}

```{r}
#| label: small-join-data
#| echo: false
visit_demo <- tibble(id = c("P01", "P01", "P02"),
                     week = c(0L, 4L, 0L),
                     marker = c(4.1, 4.4, 5.2))
meta_demo <- tibble(id = c("P01", "P02"),
                    arm = c("control", "treated"))
visit_demo
meta_demo
```

There are three visits but only two participants.

## Optional: The metadata repeats, the visits stay {.smaller}

```{r}
#| label: small-join-result
visit_demo |>
  left_join(meta_demo, by = "id",
            relationship = "many-to-one")
```

The metadata table has at most one row per ID.
The visit table supplies the three rows we preserve.

## Optional: Different names for the same key {.smaller}

```{r}
#| label: different-key-names
meta_named <- meta_demo |> rename(patient_id = id)
visit_demo |>
  left_join(meta_named, by = c("id" = "patient_id"),
            relationship = "many-to-one")
```

The left name belongs to the left table; the right name belongs to the right table.
Check that the values and types represent the same identifiers.


## Optional: A visit-level join needs two keys {.smaller}

```{r}
#| label: two-key-data
assay_demo <- tibble(id = c("P01", "P01", "P02"),
                     week = c(0L, 4L, 0L),
                     assay = c(10, 12, 20))
visit_demo |>
  left_join(assay_demo, by = c("id", "week"),
            relationship = "one-to-one")
```

`by = "id"` alone would pair P01's baseline with both assay visits.

## Optional: What happens when week is omitted? {.smaller}

```{r}
#| label: two-key-error
#| error: true
visit_demo |>
  left_join(assay_demo, by = "id",
            relationship = "one-to-one")
```

Both tables contain repeated IDs for a valid reason.
Correct the **join key**, rather than deleting valid visits.


## Optional: Row counts in an equality join {.smaller}

For each row of `x`, let `m` be its number of matches in `y`.

| Join | Rows contributed by that row of `x` |
|---|---|
| `inner_join()` | `m` |
| `left_join()` | `max(1, m)` |

Add these contributions across `x`. An unmatched left-hand row contributes
one row to a left join, with missing values in the added columns.


## Optional: Relationships depend on both directions {.smaller}

| Expected relationship | Match limit |
|---|---|
| one-to-one | Each row on either side matches at most one on the other |
| many-to-one | Each `x` row matches at most one `y` row |
| one-to-many | Each `y` row matches at most one `x` row |
| many-to-many | Multiple matches can occur in both directions |

Many patients can share a site. A unique site lookup therefore supports a
**many-to-one** patient-to-site join.


## Optional: Many-to-one: the safe case {.smaller}

`sites` has **at most one** row per `site`, so 60 rows in, 60 rows out.

```{r}
#| label: m2o
nrow(trial)
nrow(left_join(trial, sites, by = "site"))
```

. . .

We are entitled to expect that only because we checked:

```{r}
#| label: m2o-check
sites |> count(site) |> filter(n > 1)
```

. . .

::: {.callout-note appearance="simple"}
A common lookup task is many-to-one: many patients, one lookup row each.
When a join is *supposed* to be many-to-one and is not, you have found a data
error — not a join problem.
:::


## Optional: One-to-many: rows multiply on purpose {.smaller}

`labs` has **two** rows per patient — baseline and week 12.

```{r}
#| label: o2m
nrow(labs)
trial_labs <- left_join(trial, labs, by = "id")
nrow(trial_labs)
```

. . .

`60 × 2 = 120`, predicted before running it.

```{r}
#| label: o2m-head
head(trial_labs[, c("id", "arm", "week", "creat")], 4)
```

. . .

::: {.smaller}
This is correct and often exactly what you want — the result's grain is now
**one row per patient-visit**. But every patient-level column is now
**repeated**. A mean on this table weights patients by their visit counts.
Here everyone has two visits, so it equals the patient-level mean.
:::


## Optional: Many-to-many: rows multiply by accident {.smaller}

`sites_dup` contains **two conflicting records for site 2** (240 versus 245 beds).

```{r}
#| label: m2m
#| warning: true
nrow(left_join(trial, sites_dup, by = "site"))
```

. . .

Site 2 has **18** patients. `60 + 18 = 78`.

. . .

::: {.callout-important appearance="simple"}
Eighteen patients are now in the data twice. Patient counts are inflated and summaries or models can change — and the only visible symptom is a row count that nobody looked
at.
:::


## Optional: Predict, then check {.smaller}

Write the prediction into the code. It costs one line:

```{r}
#| label: predict
#| error: true
joined <- left_join(trial, sites, by = "site")
stopifnot(nrow(joined) == nrow(trial))
```

. . .

```{r}
#| label: predict-fail
#| error: true
joined_bad <- left_join(trial, sites_dup, by = "site")
stopifnot(nrow(joined_bad) == nrow(trial))
```

. . .

::: {.smaller}
`stopifnot()` is session 12 material — borrowed here as a one-liner. It does
nothing when the condition holds and stops the render when it does not, which is
exactly what you want from an assumption.
:::


## Optional: The `relationship` argument {.smaller}

State what you believe, and let the join enforce it:

```{r}
#| label: relationship
#| error: true
left_join(trial, sites_dup, by = "site", relationship = "many-to-one")
```

. . .

::: {.smaller}
Four values: `"one-to-one"`, `"one-to-many"`, `"many-to-one"`,
`"many-to-many"`. The last one is not a safety rail — it is how you tell dplyr
you meant it, and it silences the warning.
:::


## Optional: The `unmatched` argument {.smaller}

```{r}
#| label: unmatched
#| error: true
left_join(trial, sites, by = "site", unmatched = "error")
```

. . .

::: {.smaller}
Read the message carefully. For a `left_join()`, rows of `x` are **never**
dropped — so `unmatched=` guards the only side that can lose rows: **`y`**. It
just told us site 5 is unused.
:::

. . .

To assert that every *patient* matched, use the diagnostic:

```{r}
#| label: unmatched-anti
nrow(anti_join(trial, sites, by = "site")) == 0
```


## Optional: `merge()` arguments {.smaller}

```r
# Same key name in both tables:
merge(trial, sites, by = "site", all.x = TRUE)

# Different key names:
merge(trial, site_lookup,
      by.x = "site", by.y = "site_code", all.x = TRUE)
```

- `all.x = TRUE`: keep unmatched rows from `x`.
- `all.y = TRUE`: keep unmatched rows from `y`.
- `all = TRUE`: keep unmatched rows from both.
- `sort = TRUE` is the default.


## Optional: The join family in both dialects {.smaller}

::: {.comparison}
| dplyr | `merge()` | Rows |
|---|---|---|
| `inner_join(x, y, by = "k")` | `merge(x, y, by = "k")` | 45 |
| `left_join(x, y, by = "k")` | `merge(x, y, by = "k", all.x = TRUE)` | 60 |
| `right_join(x, y, by = "k")` | `merge(x, y, by = "k", all.y = TRUE)` | 46 |
| `full_join(x, y, by = "k")` | `merge(x, y, by = "k", all = TRUE)` | 61 |
| `semi_join(x, y, by = "k")` | `x[x$k %in% y$k, ]` | 45 |
| `anti_join(x, y, by = "k")` | `x[!(x$k %in% y$k), ]` | 15 |
:::

. . .

```{r}
#| label: merge-counts
c(inner = nrow(merge(trial, sites, by = "site")),
  left  = nrow(merge(trial, sites, by = "site", all.x = TRUE)),
  right = nrow(merge(trial, sites, by = "site", all.y = TRUE)),
  full  = nrow(merge(trial, sites, by = "site", all   = TRUE)))
```


## Optional: The same job, twice {.smaller}

**Task.** Attach site names, keep only patients at a known site, and report mean
12-week SBP by site.

::: columns
::: {.column width="49%"}
```{r}
#| label: hh-tidy
trial |>
  inner_join(sites, by = "site") |>
  group_by(site_name) |>
  summarise(
    n    = n(),
    mean = round(mean(sbp_12), 1)
  )
```
:::

::: {.column width="49%"}
```{r}
#| label: hh-base
m <- merge(trial, sites,
           by = "site")
aggregate(sbp_12 ~ site_name,
          data = m,
          FUN  = mean)
```
:::
:::


## Optional: `merge()` gotchas {.smaller}

::: {.comparison}
| Behaviour | Consequence |
|---|---|
| `sort = TRUE` by default | The result is **reordered by the key**. dplyr preserves `x`'s order |
| Omitted `by` | Matches all common column names; no common names gives a Cartesian product |
| Explicit `by = NULL` | Always requests a Cartesian product |
| Duplicated join keys | Can multiply rows; dplyr warns on unexpected many-to-many equality joins |
| `suffixes = c(".x", ".y")` | Same-named non-key columns collide into `age.x` / `age.y` |
:::

. . .

```{r}
#| label: merge-cross
nrow(merge(trial[1:3, c("id", "arm")], sites))   # no common column
```

Three rows and four rows gave **twelve**. That is every pairing.



# tidyr and data shape {.section-divider #tidyr-and-data-shape}

## What is tidyr? {.smaller}

`tidyr` helps organize data into a usable table:

- Each variable has a column.
- Each observation has a row.
- Each cell holds one value.

Today: `pivot_longer()` and `pivot_wider()`.

Other tidyr tools split columns or make missing rows explicit.
They are optional here.


## Wide measurements {.smaller}

```{r}
#| label: core-wide
marker_wide
```

One row per participant. `marker_w0` and `marker_w4` store the same
measurement at different weeks.

To summarize by week, it helps to give week its own column.

## Wide to long with pivot_longer {.smaller}

```{r}
#| label: core-long
marker_long <- marker_wide |>
  pivot_longer(
    cols = starts_with("marker_w"),
    names_to = "week",
    names_prefix = "marker_w",
    values_to = "marker"
  ) |>
  mutate(week = as.integer(week))
```

The column names supply `week`. The cell values supply `marker`.


## Long measurements {.smaller}

```{r}
#| label: core-long-output
#| echo: false
marker_long
```

Four participants × two weeks = eight rows.
The pair `id` and `week` identifies a row.


## A summary by week {.smaller}

```{r}
#| label: core-week-summary
marker_long |>
  group_by(week) |>
  summarise(mean_marker = mean(marker), .groups = "drop")
```

This reuses Session 6's `group_by()` and `summarise()`.

Long form is also useful for plotting measurements against time in Session 8.

## Long to wide with pivot_wider {.smaller}

```{r}
#| label: core-wide-return
marker_back <- marker_long |>
  pivot_wider(
    id_cols = id,
    names_from = week,
    values_from = marker,
    names_prefix = "marker_w"
  )
```

`id_cols` identifies an output row.
`names_from` supplies column names. `values_from` supplies cell contents.


## A check after reshaping {.smaller}

```{r}
#| label: core-round-trip
marker_back
```

Compare this with `marker_wide`:

- Same participant IDs?
- Same measurement columns and values?
- Same row count after returning to wide?

Wide and long formats are useful for different tasks.

# stringr: text columns {.section-divider}

## Text differences that affect matching {.smaller}

```{r}
#| label: core-raw-text
raw_patients |> select(id, arm)
" p01 " == "P01"
```

Our study's documented ID convention is uppercase with no surrounding spaces.
The raw export does not consistently follow it.


## Spaces and capitalization {.smaller}

```{r}
#| label: core-string-functions
str_trim("  Repeat   visit ")
str_squish("  Repeat   visit ")
str_to_lower("TREATED")
str_to_upper("p01")
```

`str_trim()` removes outer spaces.
`str_squish()` also reduces internal whitespace to one space.


## Cleaning inside mutate {.smaller}

```{r}
#| label: core-clean-text
patients_text <- raw_patients |>
  mutate(
    id = str_to_upper(str_trim(id)),
    arm = str_to_lower(str_squish(arm)),
    note = str_to_lower(str_squish(note))
  )
```

Keep `raw_patients` unchanged so the original values remain available.

## The cleaned text columns {.smaller}

```{r}
#| label: core-clean-text-output
patients_text |> select(id, arm, note)
```

The IDs now match the study's convention. Spacing and capitalization are
consistent, while each participant keeps the same treatment category.

## Detecting literal text {.smaller}

```{r}
#| label: core-detect
patients_text |>
  mutate(has_repeat = str_detect(note, fixed("repeat"))) |>
  select(id, note, has_repeat)
```

`str_detect()` returns TRUE/FALSE.
`fixed("repeat")` requests a literal substring.


## A text condition inside filter {.smaller}

```{r}
#| label: core-text-filter
patients_text |>
  filter(str_detect(note, fixed("repeat"))) |>
  select(id, note)
```

`mutate()` can retain the flag. `filter()` can retain just matching rows.

As in Session 6, `filter()` drops conditions that evaluate to `NA`.

# forcats: factors {.section-divider}

## Factor labels and level order {.smaller}

Session 4 introduced `factor()`, `levels()`, and `relevel()`.

`forcats` provides functions for the same kinds of categorical-data tasks.

```{r}
#| label: core-factor-start
arm_factor <- factor(
  patients_text$arm, levels = c("treated", "control")
)
arm_factor
```

Each participant has a category. The factor also stores an order of levels.

## fct_relevel changes the level order {.smaller}

```{r}
#| label: core-relevel
arm_ordered <- fct_relevel(arm_factor, "control")
levels(arm_ordered)
as.character(arm_factor) == as.character(arm_ordered)
```

The participant labels stay the same.

Under the usual treatment contrasts, the first level is the model reference.

For a full order: `fct_relevel(x, "c", "b", "a")`.


## fct_recode changes the labels {.smaller}

```{r}
#| label: core-recode
arm_labeled <- fct_recode(
  arm_ordered,
  Control = "control",
  Treatment = "treated"
)
arm_labeled
```

The syntax is `new_label = "old_label"`.
This deliberately changes category labels, rather than only their order.


## Factor operations inside mutate {.smaller}

```{r}
#| label: core-factor-table
patients_factor <- patients_text |>
  mutate(
    arm = fct_recode(arm,
      Control = "control", Treatment = "treated"),
    arm = fct_relevel(arm, "Control")
  )
patients_factor |> count(arm)
```

`fct_recode()` can accept a character column and returns a factor.

## Frequency order for a display {.smaller}

```{r}
#| label: core-factor-frequency
reported_site <- factor(c("East", "West", "West", "West", "East"))
levels(fct_infreq(reported_site))
```

`fct_infreq()` puts the most frequent level first.

Useful for displays. Use a deliberate scientific reference category for modeling.


# lubridate: dates {.section-divider}

## Date parsers describe the input order {.smaller}

```{r}
#| label: core-date-parsers
ymd("2026-10-18")
mdy("10/18/2026")
dmy("18/10/2026")
```

All three represent October 18, 2026.

Choose the parser using the source's date convention.


## An ambiguous date still needs a decision {.smaller}

```{r}
#| label: core-date-ambiguity
mdy("03/04/2026")
dmy("03/04/2026")
```

Both conversions succeed, but they mean different days.

A parser cannot tell you which convention the data producer intended.

## Parsing a column and extracting components {.smaller}

```{r}
#| label: core-date-column
patients_clean <- patients_factor |>
  mutate(
    enrolled_date = ymd(enrolled),
    enrolled_year = year(enrolled_date),
    enrolled_month = month(enrolled_date)
  )
```

The original text column remains available for checking.

## Dates and their components {.smaller}

```{r}
#| label: core-date-column-output
patients_clean |>
  select(id, enrolled_date, enrolled_month)
```

`enrolled_date` contains Date values. `enrolled_month` contains month numbers.

## A check after date conversion {.smaller}

```{r}
#| label: core-date-check
patients_clean |>
  filter(is.na(enrolled_date)) |>
  select(id, enrolled)
```

Zero rows means no missing parsed dates in this example.

Also inspect the range and whether the dates make sense for the study.
A successful parse alone does not establish correctness.


## One connected analysis {.smaller}

```{r}
#| label: core-connected
analysis <- marker_long |>
  left_join(
    patients_clean |> select(id, arm, enrolled_date),
    by = "id"
  )
```

The measurement table defines the rows. The cleaned metadata supplies each
participant's treatment label and enrollment date.

## The week-4 summary {.smaller}

```{r}
#| label: core-connected-output
analysis |> filter(week == 4) |>
  group_by(arm) |>
  summarise(mean_marker = mean(marker), .groups = "drop")
```

P05 still has missing metadata. Keep that visible and investigate.


## Summary {.smaller #core-summary}

- `dplyr` joins match records by keys.
- `tidyr` pivots change how measurements occupy rows and columns.
- `stringr` cleans and searches text.
- `forcats` distinguishes relabeling from level ordering.
- `lubridate` parses dates using a known input convention.

The core lesson ends here. Detailed pivots and window functions follow as
optional reference material.


# Optional reshaping and window examples {.section-divider #optional-appendix}

[Return to the core summary](#core-summary)

```{r}
#| label: appendix-setup
#| include: false
sbp_long <- wide3 |>
  pivot_longer(starts_with("sbp_"),
    names_to = "week", names_prefix = "sbp_",
    names_transform = list(week = as.integer), values_to = "sbp")
```

## Optional: Two measurements encoded in column names {.smaller}

```{r}
#| label: bp-wide
bp_wide
```

`sbp_12` contains a measurement name and a visit week.
The separator is an underscore.

## Optional: `names_sep`: one value column {.smaller}

```{r}
#| label: names-sep-live
bp_long <- bp_wide |>
  pivot_longer(
    cols = -id,
    names_to = c("measure", "week"),
    names_sep = "_",
    names_transform = list(week = as.integer),
    values_to = "value"
  )
```

One row now represents a **participant, measurement, and week**.


## Optional: The result of splitting the names {.smaller}

```{r}
#| label: names-sep-output
head(bp_long, 6)
```

There are 3 participants × 2 measurements × 2 visits = 12 rows.

## Optional: One output cell needs one value {.smaller}

```{r}
#| label: pivot-duplicate-input
repeat_readings <- tibble(
  id = c("P01", "P01"),
  week = c(4L, 4L),
  marker = c(4.4, 4.6)
)
repeat_readings |> count(id, week)
```

Two readings target the same participant-week cell.
Decide whether these are valid repeats, a correction, or an error.

## Optional: A duplicate-cell warning {.smaller}

```{r}
#| label: pivot-duplicate-warning
#| warning: true
repeat_readings |>
  pivot_wider(id_cols = id, names_from = week,
              values_from = marker)
```

The output contains a list column because both values need somewhere to go.
Using `values_fn = mean` would impose an averaging rule.
Use that rule only when the analysis calls for it.


## Optional: Missing measurements remain in the long table {.smaller}

```{r}
#| label: pivot-missing
missing_wide <- tibble(id = c("P01", "P02"),
                       marker_w0 = c(4.1, 5.2),
                       marker_w4 = c(NA, 4.7))
missing_long <- missing_wide |>
  pivot_longer(starts_with("marker_w"),
    names_to = "week", names_prefix = "marker_w",
    names_transform = list(week = as.integer),
    values_to = "marker")
```

The default is `values_drop_na = FALSE`: keep the missing measurement.

## Optional: A missing value still has a row {.smaller}

```{r}
#| label: pivot-missing-output
missing_long
```

`values_drop_na = TRUE` would remove the P01 week-4 row.
That changes the available denominator for visit-level summaries.


## Optional: Base `reshape()`: recognition only {.smaller}

```{r}
#| label: base-reshape
sbp_long <- reshape(wide3,
  direction = "long",
  idvar     = "id",
  varying   = list(c("sbp_0", "sbp_6", "sbp_12")),
  v.names   = "sbp",
  timevar   = "week",
  times     = c(0, 6, 12))
dim(sbp_long)
```

. . .

```{r}
#| label: base-reshape-back
back <- reshape(sbp_long, direction = "wide", idvar = "id",
                timevar = "week", v.names = "sbp", sep = "_")
dim(back)
```

. . .

::: {.callout-important appearance="simple"}
`reshape()` provides a base-R alternative. Its `idvar`, `timevar`, and
`varying` arguments describe the same reshaping task with different syntax.
Today's exercises use `pivot_longer()`. Either dialect is useful in other work.
:::


## Optional: Why long: the plot {.smaller}

```{r}
#| label: fig-spaghetti
#| fig-cap: "Mean systolic BP by visit and arm. Long format, three lines of ggplot2."
#| fig-width: 8
#| fig-height: 3.0
ggplot(sbp_long, aes(week, sbp, colour = arm)) +
  stat_summary(fun = mean, geom = "line", linewidth = 1.1) +
  labs(x = "Week", y = "Mean SBP (mmHg)", colour = NULL)
```

::: {.smaller}
Borrowed from session 1's minimum-viable plot — the grammar is session 8.
:::



## Optional: Within patient, in order {.smaller}

`lag()` looks back one row. `group_by()` says *back within what*.

```{r}
#| label: lag-demo
visits <- sbp_long |>
  arrange(id, week) |>
  group_by(id) |>
  mutate(
    prev_sbp = lag(sbp),
    change   = sbp - lag(sbp)
  ) |>
  ungroup()
```

## Optional: Previous-visit changes: the output {.smaller}

```{r}
#| label: lag-demo-output
head(as.data.frame(visits), 6)
```

Each participant starts with a missing previous value and change.


## Optional: Numbering, looking forward, and baselines {.smaller}

```{r}
#| label: window-more
numbered_visits <- sbp_long |>
  arrange(id, week) |>
  group_by(id) |>
  mutate(
    visit      = row_number(),
    next_sbp   = lead(sbp),
    from_base  = sbp - first(sbp)
  ) |>
  ungroup()
```

## Optional: Visit numbers and baseline changes: the output {.smaller}

```{r}
#| label: window-more-output
head(as.data.frame(numbered_visits), 6)
```

::: {.smaller}
`row_number()` numbers rows within group · `lead()` looks forward · `first()`
takes the group's first value · `cumsum()` accumulates.
:::


## Optional: What does change compare? {.smaller}

For each participant, after `arrange(id, week)`:

| Expression | Meaning |
|---|---|
| `marker - lag(marker)` | Change since the previous recorded row |
| `marker - first(marker)` | Change from the first recorded row |

The first recorded row is baseline only if baseline is present.
With irregular visits, a previous-row change is not automatically a four-week change.


## Optional: Cumulative observed measurements {.smaller}

```{r}
#| label: cumulative-data
tracking <- tibble(id = rep(c("P01", "P02"), each = 3),
                   week = rep(c(0L, 4L, 8L), 2),
                   marker = c(4.1, NA, 4.5, 5.2, 4.7, NA))
tracked <- tracking |>
  arrange(id, week) |>
  group_by(id) |>
  mutate(observed_so_far = cumsum(!is.na(marker))) |>
  ungroup()
```

## Optional: Cumulative counts: the output {.smaller}

```{r}
#| label: cumulative-output
tracked
```

`!is.na(marker)` gives TRUE/FALSE; cumulative sums count observed visits.



## Optional: stacking rows with `bind_rows()` {.smaller}

```{r}
#| label: bind-rows-demo
batch_a <- tibble(id = "P01", marker = 4.1)
batch_b <- tibble(id = "P02", marker = 5.2, lab = "central")
stacked <- bind_rows(A = batch_a, B = batch_b, .id = "batch")
stacked
```

Columns align by name. Missing columns get `NA`.
`.id` records which input supplied each row.


## Optional: stacking versus matching {.smaller}

| Files contain | Operation |
|---|---|
| The same variables for different participants | `bind_rows()` |
| Participant measurements and participant metadata | A join by ID |
| Two visit-level measurements | A join by ID and week |

After stacking, check keys and column types again.
Overlapping exports can introduce duplicated observations.

## Optional: competing columns after a join {.smaller}

```{r}
#| label: suffix-demo
clinic <- tibble(id = c("P01", "P02"), age = c(NA, 61))
registry <- tibble(id = c("P01", "P02"), age = c(58, 62))
ages <- left_join(clinic, registry, by = "id",
                   suffix = c("_clinic", "_registry"),
                   relationship = "one-to-one")
ages
```

Both age columns survive with suffixes. P02's sources disagree.

## Optional: `coalesce()` follows a priority rule {.smaller}

```{r}
#| label: coalesce-demo
ages |>
  mutate(age = coalesce(age_clinic, age_registry))
```

The stated rule here is clinic first, registry only if clinic is missing.
P02 keeps 61. `coalesce()` does not detect or resolve that disagreement.


## Optional: separate measurement columns with `.value` {.smaller}

```{r}
#| label: pivot-value-demo
bp_by_visit <- bp_wide |>
  pivot_longer(-id,
    names_to = c(".value", "week"),
    names_sep = "_",
    names_transform = list(week = as.integer))
head(bp_by_visit, 4)
```

`.value` uses the first name component to create `sbp` and `dbp` columns.


## Optional: a missing row versus a missing value {.smaller}

```{r}
#| label: complete-input
incomplete <- tibble(id = c("P01", "P01", "P02"),
                     week = c(0L, 4L, 0L),
                     marker = c(4.1, NA, 5.2))
incomplete
```

P01 week 4 exists with a missing marker.
P02 week 4 has no row.

## Optional: `complete()` makes expected visits explicit {.smaller}

```{r}
#| label: complete-demo
completed <- incomplete |>
  mutate(record_present = TRUE) |>
  complete(id, week = c(0L, 4L),
           fill = list(record_present = FALSE))
completed
```

The added marker stays `NA`. No measurement was invented.
Use an expected visit schedule, not every imaginable ID-time combination.

