---
title: "Session 9 lab: Copilot review"
format: html
---

```{r}
visit <- data.frame(
  id = rep(sprintf("P%02d", 1:8), each = 2),
  arm = rep(rep(c("control", "treated"), each = 4), each = 2),
  week = rep(c(0, 4), 8),
  marker = c(4.1, 4.3, 4.8, NA, 5.0, 5.2, 4.5, 4.7,
             5.2, 4.6, 5.8, 5.0, 6.1, NA, 5.5, 4.9)
)
```

## Statistical request

At week 4, report total `n`, observed `n`, mean marker, and SD by randomized
arm. Do not remove missing rows before counting total `n`.

## First candidate

Ask Copilot to suggest code below the comment. If unavailable, begin with this
candidate:

```r
visit |>
  dplyr::filter(week == 4, !is.na(marker)) |>
  dplyr::summarise(n = dplyr::n(), mean = mean(marker), sd = sd(marker), .by = arm)
```

Explain what is correct and incorrect about the candidate.

## Corrected analysis

```{r}
#| eval: false
# Add corrected dplyr or base-R code.
```

## Independent verification

```{r}
# Verify total and observed counts using a different, simple approach.
```

## AI disclosure

Replace this sentence with the required disclosure.

