---
title: "Debugging and Reliable Code"
subtitle: "BIOS 6301 · Introduction to Statistical Computing"
author: "Department of Biostatistics, Vanderbilt University Medical Center"
date: "Session 12"
date-format: "[Session 12]"
format:
  revealjs:
    theme: simple
    css: ../../slides.css
    slide-number: c/t
    show-slide-number: all
    incremental: false
    width: 1280
    height: 720
    fig-align: center
    code-line-numbers: false
    highlight-style: github
    footer: "BIOS 6301 · Session 12 · Debugging and Reliable Code"
    navigation-mode: linear
execute:
  enabled: true
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(ggplot2)
library(knitr)

# Synthetic teaching data for BIOS 6301. Not real patients.
set.seed(6301)
n <- 60
trial <- data.frame(
  id        = sprintf("P%03d", 1:n),
  arm       = rep(c("Placebo", "Treatment"), each = n / 2),
  site      = sample(1:4, n, replace = TRUE),
  age       = round(rnorm(n, 61, 9)),
  sex       = sample(c("F", "M"), n, replace = TRUE),
  weight_kg = round(rnorm(n, 82, 15), 1),
  sbp_0     = round(rnorm(n, 148, 14)),
  stringsAsFactors = FALSE
)
trial$sbp_12 <- round(trial$sbp_0 -
  ifelse(trial$arm == "Treatment", rnorm(n, 14, 8), rnorm(n, 3, 8)))

# Added for this session: the derived outcome every demo summarises.
trial$change <- trial$sbp_12 - trial$sbp_0

# A copy carrying the defects a real extract arrives with. Same object as
# session 11, with one column added. What is broken, deliberately:
#   sbp_12    three missing 12-week outcomes (P007, P023, P048)
#   weight_kg one weight recorded as the sentinel -99 (P012)
#   creat_raw creatinine as *character*, carrying "NA" and a "<0.30" limit-of-
#             detection string, which is how the lab's export actually arrives
trial_raw <- trial
trial_raw$sbp_12[c(7, 23, 48)] <- NA
trial_raw$weight_kg[12] <- -99
trial_raw$change <- trial_raw$sbp_12 - trial_raw$sbp_0
trial_raw$creat_raw <- as.character(round(rnorm(n, 0.95, 0.22), 2))
trial_raw$creat_raw[c(4, 31)] <- "NA"
trial_raw$creat_raw[19] <- "<0.30"

theme_set(theme_minimal(base_size = 15))
```

## Where we are {.smaller}

| Course arc | Sessions | |
|---|---:|---|
| Setup and workflow | 0–2 |  |
| Data foundations | 3–8 |  |
| Modern tools | 9–10 |  |
| **R programming** | **11–13** | **← today** |
| Statistical computing | 14–16 |  |

Today's deck develops one part of the current arc; later-session tools are named only when they clarify where this idea will be used.


## What you will be able to do {.smaller}

::: {.smaller}
[1]{.objective-num} **Read** an R error or warning and say which of the three
conditions it is and where it came from.

[2]{.objective-num} **Isolate** a minimal failing example from a script that does
not work.

[3]{.objective-num} **Use** `traceback()`, `browser()`, and `debug()` to find a
bug rather than guessing at it.

[4]{.objective-num} **Write** code that fails fast, using guard clauses,
`stopifnot()`, and small tests.

[5]{.objective-num} **Explain** why `0.1 + 0.2 != 0.3`, and compare numbers
correctly.

[6]{.objective-num} **Justify** why `suppressWarnings()` is almost never the
right fix.

[7]{.objective-num} **Post** a reprex that somebody can actually answer.
:::


## Roadmap for today

| | Block | Minutes |
|---|---|---|
| 1 | What the machine is telling you | 12 |
| 2 | Isolating a minimal failing example | 18 |
| 3 | `traceback()`, `browser()`, `debug()` | 22 |
| 4 | Fail fast: assertions and small tests | 14 |
| 5 | Floating point | 20 |
| 6 | `suppressWarnings()`, naming, and the reprex | 12 |

: {tbl-colwidths="[6,64,30]"}

. . .

**In lab this week** you fix five genuinely broken functions — one for each way
code goes wrong.


# 1 · What the machine is telling you {.section-divider}

## Two kinds of broken {.smaller}

::: columns
::: {.column width="48%"}
### It stops

```
Error in sum(x) :
  invalid 'type' (character) of argument
```

**Loud.** You know instantly. The location is roughly right. The worst case is
that you lose an hour.
:::

::: {.column width="48%"}
### It finishes

```
[1] -7.88
```

**Silent.** A number, correctly formatted, plausible, and wrong. Nothing on
screen distinguishes it from the right answer.
:::
:::

. . .

::: {.callout-warning appearance="simple"}
Almost everything in this session is aimed at the second column. The first column
is a nuisance; the second column is how analyses get retracted.
:::


## The three conditions {.smaller}

R has exactly three, and they differ in **what happens next**:

```{r}
#| label: three-conditions
#| warning: true
#| message: true
#| error: true
check_dose <- function(mg) {
  message("checking ", length(mg), " doses")   # 1. carries on
  if (any(mg > 100)) warning("dose above 100 mg")  # 2. carries on
  if (any(mg < 0)) stop("negative dose")            # 3. stops here
  invisible(TRUE)
}

check_dose(c(10, 120, -5))
```

. . .

::: {.smaller}
| Condition | Execution | Goes to | Use it for |
|---|---|---|---|
| `message()` | continues | stderr | Progress, and what you did |
| `warning()` | continues | stderr, often **deferred** | "This is probably not what you meant" |
| `stop()` | **halts** | stderr | "I cannot produce a correct answer" |
:::


## Anatomy of an error message {.smaller}

```{r}
#| label: anatomy
#| warning: true
#| error: true
mean(trial_raw$creat_raw)
sum(trial_raw$creat_raw)
```

. . .

::: {.smaller}
```
Error in sum(trial_raw$creat_raw) : invalid 'type' (character) of argument
        └──────── where ────────┘   └──────────── what ─────────────────┘
```

- **`Error in`** names the call that raised it — *not necessarily the call that is
  wrong*.
- **The text after the colon** is R's vocabulary, and it is small. `invalid
  'type'`, `object not found`, `subscript out of bounds`, `non-numeric argument`,
  `argument of length zero`. You will learn all of them this term.
:::


## Warnings are the ones that cost you {.smaller}

```{r}
#| label: coercion-warning
#| warning: true
creat <- as.numeric(trial_raw$creat_raw)
mean(creat)
```

. . .

```{r}
#| label: coercion-diagnose
sum(is.na(creat))
trial_raw$creat_raw[is.na(creat)]
```

. . .

::: {.callout-important appearance="simple"}
The warning told you the truth: three values could not be read. Two are genuinely
missing. **One is a real measurement** — a below-limit-of-detection result — and
you have just silently deleted it.
:::


## Where it stops is not where it broke {.smaller}

```{r}
#| label: distance
#| error: true
# Mean change in SBP, for a subgroup, rounded for the table.
mean_change <- function(data) summarise_change(data$change)
summarise_change <- function(x) round(pooled_mean(x), 1)
pooled_mean <- function(x) sum(x) / length(x)

site_3 <- subset(trial_raw, site == 3)
site_3$change <- as.character(site_3$change)  # ← the actual bug, 3 lines up

mean_change(site_3)
```

. . .

::: {.smaller}
The error names `sum(x)`. `sum()` is three function calls away from the mistake
and did nothing wrong. **The distance between the symptom and the cause is what
makes debugging hard** — and it is what `traceback()` exists to close.
:::


# 2 · Isolating a minimal failing example {.section-divider}

## The procedure {.smaller}

The most valuable debugging skill, and the one nobody teaches:

::: {.incremental}
1. **Reproduce it.** From a clean session. If you cannot make it fail on demand,
   you cannot tell when you have fixed it.
2. **Cut the data down.** 60 rows → 6 rows → 1 row. Does it still fail?
3. **Cut the code down.** Delete a step. Still fails? Delete another.
4. **Keep cutting until removing anything makes the bug disappear.** What is left
   is the bug.
:::

. . .

::: {.callout-tip appearance="simple"}
Most of the time you never reach step 4. The bug becomes obvious somewhere in
step 3, and you never find out how small the example could have been.
:::


## Cut the data down {.smaller}

Start from the failure, and make the input smaller:

```{r}
#| label: cut-data
#| error: true
mean_change(site_3)                 # 11 rows — fails
mean_change(site_3[1:3, ])          # 3 rows  — still fails
mean_change(site_3[1, ])            # 1 row   — still fails
```

. . .

```{r}
#| label: cut-data-2
#| error: true
pooled_mean(site_3$change[1])       # drop the wrappers entirely
site_3$change[1]                    # look at the value itself
```

. . .

::: {.smaller}
One row still fails, so the bug is **not** about missingness, not about the
subgroup, and not about `n`. It is about a single value — and the quotes on the
last line are the answer.
:::


## Cut the code down {.smaller}

The other direction: keep the data, delete the pipeline.

```r
# what you have — and it fails somewhere
report <- trial |>
  filter(!is.na(sbp_12), site %in% c(2, 3)) |>
  mutate(change = sbp_12 - sbp_0,
         responder = change <= -10,
         band = cut(age, breaks = quantile(age, c(0, .5, 1)))) |>
  group_by(arm, band) |>
  summarise(n = n(), pct = 100 * mean(responder), .groups = "drop")
```

. . .

```r
# run it one verb at a time and look at the result of each
step1 <- trial |> filter(!is.na(sbp_12), site %in% c(2, 3))
nrow(step1)                                  # 0? then the filter is the bug
step2 <- step1 |> mutate(change = sbp_12 - sbp_0)
summary(step2$change)                        # all NA? then the mutate is
```

. . .

::: {.callout-note appearance="simple"}
A pipeline is unusually hard to debug precisely because the intermediate results
have no names. Breaking it apart is not a failure of style — it is the technique.
:::


## Bisection, and what it buys you {.smaller}

When you have no hypothesis at all, **halve it**:

::: {.smaller}
- Comment out the bottom half of the script. Does it still fail?
- No → the bug is in the half you removed. Yes → it is in the half you kept.
- Repeat. Two hundred lines becomes eight in about five steps.
:::

. . .

::: {.smaller}
The same move works on **history** rather than on code — `git bisect` finds the
commit that introduced a bug by the same halving. Your session 2 habit of small
commits is what makes it possible.
:::

. . .

::: {.callout-important appearance="simple"}
The minimal example you end up with is not a by-product. It is the **test** you
add so the bug cannot come back, and the **reprex** you post if you still need
help.
:::


# 3 · `traceback()`, `browser()`, `debug()` {.section-divider}

## `traceback()` {.smaller}

Run it **immediately** after an error. It prints the call stack as it was when R
gave up:

```r
> mean_change(site_3)
Error in sum(x) : invalid 'type' (character) of argument

> traceback()
3: pooled_mean(x)
2: summarise_change(data$change)
1: mean_change(site_3)
```

. . .

::: {.smaller}
**Read it from the bottom up.** Line 1 is what *you* typed. Line 3 is the last
thing R was doing. The bug is somewhere on the way down.
:::

. . .

::: {.callout-warning appearance="simple"}
`traceback()` shows the *most recent* error only, and only after a real error at
the console. It has nothing to say inside a rendering document.
:::


## Reading a stack {.smaller}

```{mermaid}
%%| echo: false
%%| fig-width: 9
flowchart TD
  A["<b>1:</b> mean_change(site_3)<br/><i>what you typed</i>"] --> B["<b>2:</b> summarise_change(x)<br/><i>x is data$change —<br/>already character</i>"]
  B --> C["<b>3:</b> pooled_mean(x)"]
  C --> D["<b>sum(x)</b><br/>error raised here"]
  style A fill:#f4efe4,stroke:#CFAE70
  style B fill:#CFAE70,stroke:#8a7448,color:#fff
  style D fill:#e8d7d7,stroke:#a05252
```

::: {.smaller}
The error is raised at the bottom. The **cause** entered at the top. `traceback()`
gives you the path between them, which is the thing you cannot get any other way.
:::


## `browser()` {.smaller}

`traceback()` tells you *where*. `browser()` lets you *look*.

```r
pooled_mean <- function(x) {
  browser()            # ← execution pauses here, with x available
  sum(x) / length(x)
}

mean_change(site_3)
```

. . .

```
Called from: pooled_mean(x)
Browse[1]> str(x)
 chr [1:11] "-2" "-14" "-25" "-9" ...
Browse[1]> class(x)
[1] "character"
Browse[1]> Q
```

. . .

::: {.smaller}
Inside the browser you are in a **normal R console**, sitting inside the paused
function. Every object in that frame is real and inspectable.
:::


## The browser commands {.smaller}

::: {style="font-size:0.92em"}
| Key | Does | You want it when |
|---|---|---|
| `n` | **next** — run this line, stop at the next | Walking through the body |
| `s` | **step in** — descend into the call on this line | The suspect is the function being called |
| `f` | **finish** — run to the end of this loop or function | You stepped in by mistake |
| `c` | **continue** — resume until the next `browser()` or the end | You have seen enough |
| `Q` | **quit** — abandon it, back to the console | Almost always, at the end |
| `where` | print the stack from here | You have lost track of how deep you are |
:::

. . .

::: {.callout-warning appearance="simple"}
A variable called `n`, `s`, `c`, `f` or `Q` cannot be printed by typing its name
— the browser reads it as a command. Use `print(n)`.
:::


## `debug()`, `debugonce()`, and breakpoints {.smaller}

When you cannot edit the function — it is in a package, or it is 300 lines away:

```r
debugonce(pooled_mean)   # browse the next call, then forget about it
mean_change(site_3)

debug(pooled_mean)       # browse EVERY call until you stop it
undebug(pooled_mean)     # ...which you must remember to do
```

. . .

::: {.smaller}
| | Edit the file? | Persists? |
|---|---|---|
| `browser()` in the body | yes | until you delete it |
| **`debugonce(f)`** | **no** | **one call — the safe default** |
| `debug(f)` | no | until `undebug(f)` |
| RStudio breakpoint | no (stored by the IDE) | until you click it off |
:::


## `options(error = recover)` {.smaller}

The heavy tool. On **any** error, R offers you every frame in the stack and drops
you into the one you choose:

```r
options(error = recover)

mean_change(site_3)
```

```
Error in sum(x) : invalid 'type' (character) of argument

Enter a frame number, or 0 to exit

1: mean_change(site_3)
2: summarise_change(data$change)
3: pooled_mean(x)

Selection: 2
Called from: summarise_change(data$change)
Browse[1]> str(data$change)
 chr [1:11] "-2" "-14" "-25" ...
```

. . .

::: {.smaller}
Turn it off with `options(error = NULL)`. **Never set it in a script or a `.qmd`.**
:::


## Which tool, when {.smaller}

::: {style="font-size:0.88em"}
| Situation | Reach for |
|---|---|
| An error, and you do not know which function raised it | `traceback()` |
| You know the function, and want to see its arguments | `browser()` or `debugonce()` |
| The function is in a package | `debugonce(pkg::f)` |
| No error, but the answer is wrong | `browser()` + the minimal example |
| An error you cannot reproduce cheaply | `options(error = recover)` |
| A warning and you want to know where from | `options(warn = 2)`, then `traceback()` |
| Anything at all, and you are in a hurry | `print()` and `str()` |
:::

. . .

::: {.callout-tip appearance="simple"}
The last row is not a joke. Well-placed `print()` calls solve most bugs, and there
is no shame in them.
:::


# 4 · Fail fast, fail loud {.section-divider}

## Fail fast, fail loud {.smaller}

Session 11 gave you guard clauses. The argument is worth restating in this
session's terms:

::: {.smaller}
- A function that **stops** costs you ten minutes.
- A function that returns a **plausible wrong number** costs you a paper.
:::

. . .

```{r}
#| label: fail-fast
#| error: true
# Responder rate: percentage of patients with a fall of at least 10 mmHg.
responder_rate <- function(change) {
  stopifnot(is.numeric(change))
  if (length(change) > 0 && all(is.na(change))) {
    stop("change is entirely missing", call. = FALSE)
  }

  n_obs <- sum(!is.na(change))
  100 * sum(change <= -10, na.rm = TRUE) / n_obs
}

responder_rate(as.character(trial$change))
```


## `stopifnot()` outside a function {.smaller}

Assertions are not only for functions. Put them in the **analysis** where an
assumption is made:

```{r}
#| label: inline-assertions
#| error: true
# One row per patient is assumed by every summary below.
stopifnot(!anyDuplicated(trial$id))

# The join in session 7 must not have created rows.
stopifnot(nrow(trial) == 60)

# Change must be derived, never imported.
stopifnot(all(trial$change == trial$sbp_12 - trial$sbp_0))

# And the one that actually fails:
stopifnot("no outcome may be missing" = !anyNA(trial_raw$sbp_12))
```

. . .

::: {.smaller}
Each line is a sentence you would otherwise have written as a comment — except
this one is **checked every time the document renders**.
:::


## Make the check permanent {.smaller}

You already built a minimal failing example. **Do not throw it away.**

```{r}
#| label: assertion-tests
# Known answers, worked out by hand, checked on every render.
stopifnot(
  "empty input gives NaN, not an error" =
    is.nan(responder_rate(numeric(0))),
  "a single responder gives 100" =
    responder_rate(-12) == 100,
  "a single non-responder gives 0" =
    responder_rate(-2) == 0,
  "missing values are excluded from the denominator" =
    responder_rate(c(-12, NA)) == 100
)
cat("all checks passed\n")
```

. . .

::: {.smaller}
Four lines, and the fourth is the one the lab's broken version fails. A test is
just **an example with a known answer**, written down.
:::


## `testthat`, in one slide {.smaller}

When the project is big enough to want real tests:

```r
library(testthat)

test_that("responder_rate handles missingness", {
  expect_equal(responder_rate(-12), 100)
  expect_equal(responder_rate(c(-12, NA)), 100)
  expect_error(responder_rate("a"), "is.numeric")
  expect_true(is.nan(responder_rate(numeric(0))))
})
```

. . .

::: {.smaller}
| | Does |
|---|---|
| `expect_equal(x, y)` | Equal **within tolerance** — the right choice for numbers |
| `expect_identical(x, y)` | Exactly equal, type included |
| `expect_error(f(), "text")` | The call fails, with a message matching `"text"` |
| `expect_warning()`, `expect_true()` | As they read |
:::

::: {.callout-note appearance="simple"}
Not installed on the course machines and not required here. Use
`stopifnot()` — the idea is identical, and it is what the previous slide runs.
:::


# 5 · Floating point {.section-divider}

## The slide that ruins everyone's afternoon {.smaller}

```{r}
#| label: fp-first
0.1 + 0.2 == 0.3
```

. . .

```{r}
#| label: fp-second
0.1 + 0.2 - 0.3
```

. . .

```{r}
#| label: fp-third
print(0.1, digits = 22)
```

. . .

::: {.smaller}
R has not made a mistake. **`0.1` is not 0.1** — it is the closest number to 0.1
that a 64-bit double can represent, and that number is slightly too big.
:::


## Binary cannot write one tenth {.smaller}

In base 10, ⅓ needs infinitely many digits: 0.3333…

::: {.smaller}
In base 2, **one tenth** needs infinitely many digits:
`0.0001100110011001100…`
:::

. . .

::: {.smaller}
A double gets 53 bits for the value. So R stores the closest 53-bit truncation,
which is off by about 1 part in 10^16^.
:::

. . .

```{r}
#| label: fp-representable
0.5 + 0.25 == 0.75    # exact: halves and quarters ARE representable
1 - 0.9 == 0.1        # not exact
0.3 - 0.1 == 0.2      # not exact
```

. . .

::: {.callout-note appearance="simple"}
Integers up to 2^53^ are exact. Halves, quarters, eighths are exact. **Tenths are
not** — and clinical data is measured in tenths.
:::


## Machine epsilon {.smaller}

```{r}
#| label: fp-eps
.Machine$double.eps
```

. . .

::: {.smaller}
The **smallest** `e` such that `1 + e != 1`. It is the width of the gap between
representable numbers near 1 — and it is the unit in which "close enough" is
measured.
:::

```{r}
#| label: fp-eps-2
1 + .Machine$double.eps == 1
1 + .Machine$double.eps / 2 == 1
```

. . .

::: {.smaller}
The gap **scales with magnitude**. Near 1 it is 2 × 10^-16^; near 10^6^ it is
about 10^-10^; near 10^-6^ it is far smaller. Absolute tolerances that work for
one scale are wrong for another — which is why `all.equal()` defaults to a
*relative* one.
:::


## `all.equal()`, and the trap in it {.smaller}

```{r}
#| label: fp-allequal
all.equal(0.1 + 0.2, 0.3)
isTRUE(all.equal(0.1 + 0.2, 0.3))
```

. . .

```{r}
#| label: fp-allequal-trap
all.equal(1, 2)          # NOT FALSE — a character description
```

. . .

::: {.callout-important appearance="simple"}
`all.equal()` returns `TRUE` **or a description of the difference**. A non-empty
string is truthy enough to break an `if`. Always wrap it: `isTRUE(all.equal(x, y))`.
:::

. . .

::: {.smaller}
Or compare by hand, when you know the scale you care about:

```{r}
#| label: fp-tolerance
abs((0.1 + 0.2) - 0.3) < 1e-8
```
:::


## It is not only `==` {.smaller}

```{r}
#| label: fp-seq
schedule <- seq(0.1, 1.0, by = 0.1)   # titration steps, as a fraction
which(schedule == 0.3)                 # integer(0) — no such step
```

. . .

```{r}
#| label: fp-accumulate
# Session 13's loop, accumulating a total dose.
total <- 0
for (i in 1:10) total <- total + 0.1
total == 1
print(total, digits = 18)
```

. . .

::: {.smaller}
`==` was never the risk on its own. **Any equality test hidden inside something
else** carries it: `which()`, `match()`, `%in%`, `switch()`, a lookup table
keyed on a decimal, `duplicated()`, and a `while` loop that stops "when the total
reaches 1".
:::


## Integer overflow {.smaller}

```{r}
#| label: fp-intmax
.Machine$integer.max
```

. . .

```{r}
#| label: fp-overflow
#| warning: true
x <- 2147483647L
x + 1L
```

. . .

```{r}
#| label: fp-overflow-double
x + 1        # a double: no overflow, but no longer exact either
```

. . .

::: {.smaller}
R's integers are 32-bit and stop at about 2.1 billion. Doubles carry on, but
above 2^53^ they cannot represent every whole number — so a count can start
*skipping*.
:::


## Why likelihoods are computed on the log scale {.smaller}

```{r}
#| label: fp-underflow
big <- rnorm(400, mean = 148, sd = 14)

prod(dnorm(big, 148, 14))        # the likelihood, directly
log(prod(dnorm(big, 148, 14)))   # ...and its log
```

. . .

```{r}
#| label: fp-logsum
sum(dnorm(big, 148, 14, log = TRUE))   # the same quantity, computed properly
```

. . .

::: {.smaller}
The true value is about 10^-705^. The smallest positive double is about
10^-308^, so the product **underflows to exactly zero** — and `log(0)` is `-Inf`.
Summing logs never leaves the representable range.
:::


## The floating-point rules {.smaller}

::: {style="font-size:0.88em"}
| Rule | |
|---|---|
| **Never `==` on computed doubles** | `isTRUE(all.equal(x, y))`, or `abs(x - y) < tol` |
| **`==` is fine on integers and strings** | `id == "P003"`, `site == 3L` |
| **No floating-point keys, indices, or counters** | Use integers, derive the decimal |
| **Identifiers are character** | Not numeric. `colClasses` at import |
| **Likelihoods and probabilities live on the log scale** | `dnorm(..., log = TRUE)`, then `sum()` |
| **`round()` is for display** | It does not make a number exact |
| **Report at the precision you measured** | Not at the precision R prints |
:::


# 6 · Habits that prevent debugging {.section-divider}

## `suppressWarnings()` {.smaller}

```{r}
#| label: suppress
#| warning: true
creat <- suppressWarnings(as.numeric(trial_raw$creat_raw))
mean(creat, na.rm = TRUE)
```

. . .

::: {.smaller}
This runs cleanly, produces a number, and has **silently deleted a real
measurement**. The warning was the only thing standing between you and that
result.
:::

. . .

::: {.smaller}
It has exactly one legitimate use: a warning you have **read, diagnosed, and
decided is not a problem** — with a comment saying so.

```r
# Sites 1-4 only; `site` is a factor with an unused level 5 from the CRF,
# which as.integer() warns about. Checked: no patient has site 5.
site_num <- suppressWarnings(as.integer(as.character(trial$site)))
```
:::


## Names and comments that debug for you {.smaller}

```r
# what you cannot debug
x <- d[d$g == 1, ]
y <- mean(x$v, na.rm = TRUE)
z <- y * 0.1
```

. . .

```r
# what you can
site_3 <- subset(trial, site == 3)
mean_weight_kg <- mean(site_3$weight_kg, na.rm = TRUE)

# Dose is 0.1 mg per kg of body weight (protocol v2, section 4.3).
mean_dose_mg <- mean_weight_kg * 0.1
```

. . .

::: {.smaller}
The second version cannot have a unit bug hiding in it, because the units are in
the names. The comment answers the question a reader would otherwise have to ask
you — and **"why 0.1" is not answerable from the code**.
:::


## How to write a reprex {.smaller}

A **rep**roducible **ex**ample. Four requirements, and they are not negotiable:

::: {.smaller}
1. **Self-contained** — runs in a fresh session. Every `library()` included.
2. **Minimal** — block 2's procedure. Not your script.
3. **Data included** — `dput(head(x))`, or a tiny made-up frame. Never
   "my data looks like this".
4. **The actual output** — the real error text, copied, not described.
:::

. . .

```r
library(dplyr)

d <- data.frame(id = c("P001", "P002"), sbp = c("148", "152"))

d |> summarise(mean_sbp = mean(sbp))
#> Warning: argument is not numeric or logical: returning NA
#>   mean_sbp
#> 1       NA
```

. . .

::: {.smaller}
`reprex::reprex()` runs your clipboard in a clean session and formats it exactly
like this. **Most reprexes never get posted** — you find the bug while building
one.
:::


## Summary {.smaller}

- Reproduce the failure before editing the code.
- Reduce the data and code until the first broken assumption is visible.
- Use the call stack, browser, and breakpoints to inspect state deliberately.
- Assertions and small tests preserve a repair after today.
- Finite-precision arithmetic requires tolerance-aware reasoning.

. . .

**Next connection:** Session 13 asks how to repeat a correct operation clearly and efficiently.

