---
title: "Session 5 lab: data import"
date: now
date-format: "YYYY-MM-DD HH:mm:ss Z"
format: html
---

The first chunk creates synthetic practice files. Run it unchanged; writing this
setup is not a lab task. Work only on the copy in your personal repository.

```{r}
source_table <- data.frame(
  id = c("P01", "P02", "P03", "P04", "P05"),
  visit_date = c("2026-01-03", "01/04/2026", "2026-01-05",
                 "2026-02-30", "2026-01-07"),
  marker = c("4.1", ".", "6.2", "NM", "6.5")
)
csv_path <- tempfile(fileext = ".csv")
write.csv(source_table, csv_path, row.names = FALSE, quote = FALSE)
# Output paths supplied for Section 3 (export and round-trip check).
csv_out <- tempfile(fileext = ".csv")
rds_out <- tempfile(fileext = ".rds")

site_table <- data.frame(
  id = c("P11", "P12", "P13"),
  marker = c(4.2, 5.1, 6.3)
)
site_path <- tempfile(fileext = ".csv")
write.table(site_table, site_path, sep = ";", dec = ",",
            row.names = FALSE)
```

Documented rules: `.` and `NM` mean missing marker. Dates are intended
to be ISO `YYYY-MM-DD`; the non-ISO and impossible values require review.

## Input-file record

Include this table in your rendered report before importing the CSV.

```{r}
input_md5 <- tools::md5sum(csv_path)
stopifnot(!is.na(input_md5))
data.frame(
  file = basename(csv_path),
  md5 = unname(input_md5)
)
```

This exercise creates a temporary file; a real analysis should record its actual
input filename. Keep the rendered report to preserve the checksum from this run.
A different checksum means different file contents, not necessarily different
analytical values or an error in the data.

## 1. Import contract and audit

```{r}
# Preview: readLines(csv_path, n = 6)
# Import raw with colClasses = "character".
# Import analysis_import with the documented na.strings and named colClasses.
# Check dim(), names(), str(), table(id), and colSums(is.na(...)).
# Expected IDs: c("P01", "P02", "P03", "P04", "P05"); use %in%.
```

## 2. Diagnose and clean

```{r}
# Show raw id/marker rows where is.na(analysis_import$marker).
# Parse visit_date with as.Date(..., format = "%Y-%m-%d").
# Show raw id/visit_date rows for failed dates; do not guess corrections.
# Copy analysis_import to clean; add marker_raw and visit_date_raw using $.
# Replace clean$visit_date with the parsed dates.
```

## 3. Export and round-trip check

```{r}
# write.csv(clean, csv_out, row.names = FALSE)
# saveRDS(clean, rds_out)
# Read back with read.csv() and readRDS(); compare str() output.
```

## 4. Interpretation

List the records requiring source clarification and explain why.

## 5. File-layout checkpoint

Run the following file-generation chunk unchanged.

```{r}
layout_path <- tempfile(fileext = ".csv")
writeLines(c(
  "Synthetic clinic export",
  "Prepared 2026-09-26",
  "export_row,export_row,id,weight_kg",
  "1,1,001,62.4",
  "2,2,002,71.8",
  "3,3,003,68.0",
  "TOTAL,,,202.2"
), layout_path)
```

Both `export_row` columns are export counters, not participant identifiers.
Read the text first, then import the three observations and keep the IDs intact.

```{r}
# Inspect with readLines(); specify skip, nrows, check.names, and colClasses.
# Remove only the documented export-counter columns, then check names and types.
```

If a workbook used yellow highlighting to mean "exclude from analysis", what
explicit columns should replace that visual convention?

## 6. Additional practice (optional)

Choose any tasks from the lab page: delimiter/decimal settings, incomplete missing
codes, all-character import, mixed or ambiguous dates, and CSV/RDS comparison.
The optional readr task needs only the import example shown in class, `str()`,
and missing-value counts. No optional-reference-slide content is required.

```{r}
# Add your chosen additional-practice code here.
```
