---
title: "Session 5 lab: data import"
date: now
date-format: "YYYY-MM-DD HH:mm:ss Z"
format: html
---

```{r}
source_table <- data.frame(
  id = c("P01", "P02", "P03", "P04", "P05"),
  visit_date = c("2026-01-03", "01/04/2026", "2026-01-05",
                 "2026-02-30", "2026-01-07"),
  marker = c("4.1", ".", "6.2", "not measured", "6.5")
)
csv_path <- tempfile(fileext = ".csv")
write.csv(source_table, csv_path, row.names = FALSE)

site_table <- data.frame(
  id = c("P11", "P12", "P13"),
  marker = c(4.2, 5.1, 6.3)
)
site_path <- tempfile(fileext = ".csv")
write.table(site_table, site_path, sep = ";", dec = ",",
            row.names = FALSE)
```

Documented rules: `.` and `not measured` mean missing marker. Dates are intended
to be ISO `YYYY-MM-DD`; the non-ISO and impossible values require review.

## 1. Import contract and audit

```{r}
# Create a raw character import and a typed analysis import, then audit both.
```

## 2. Diagnose and clean

```{r}
# Retain raw text beside the parsed marker and date columns.
```

## 3. Export and round-trip check

```{r}
# Write temporary CSV and RDS files, read them back, and compare classes.
```

## 4. Interpretation

List the records requiring source clarification and explain why.
## 5. File-layout checkpoint

```{r}
layout_path <- tempfile(fileext = ".csv")
writeLines(c(
  "Synthetic clinic export",
  "Prepared 2026-09-26",
  "export_row,export_row,id,weight_kg",
  "1,1,001,62.4",
  "2,2,002,71.8",
  "3,3,003,68.0",
  "TOTAL,,,202.2"
), layout_path)
```

Both `export_row` columns are export counters, not participant identifiers.
Read the text first, then import the three observations and keep the IDs intact.

```{r}
# Inspect with readLines(); specify skip, nrows, check.names, and colClasses.
# Remove only the documented export-counter columns, then check names and types.
```

If a workbook used yellow highlighting to mean "exclude from analysis", what
explicit columns should replace that visual convention?
