---
title: "Session 6 lab: tidy data and dplyr"
author: "Your name"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S %Z')`"
output:
  html_document:
    self_contained: true
---

Complete the lab page's installation check in the Console first.
Keep `library(dplyr)` below so the report also works in a fresh R session.
Do not add `install.packages()` to this report.

The data-creation code is supplied. One row is one participant at one visit.
Follow tasks 1–7 on the lab page; write explanations below the relevant heading.

```{r setup}
library(dplyr)

visit <- data.frame(
  id = rep(sprintf("P%02d", 1:6), each = 2),
  arm = rep(rep(c("control", "treated"), each = 3), each = 2),
  week = rep(c(0, 4), 6),
  marker = c(4.1, 4.4, 5.0, NA, 4.6, 4.8,
             5.2, 4.6, 6.0, 5.1, 5.6, NA)
)
```

## 1. Validate the row unit and key

Explain the row unit and proposed key.

```{r key-check}
# Use count(id, week) and filter() to check for repeated keys.
```

## 2. Keep the analysis population

Write the inclusion rule in one sentence.

```{r analysis-population}
# Save all week-4 rows as analysis, including rows with missing marker.
```

## 3. Create the derived variables

```{r derived-variables}
# In analysis, create observed = !is.na(marker).
# Use case_when() to create marker_band:
# missing for NA; high for marker >= 5; lower otherwise.
# Handle missingness first.
```

## 4. Summarize by arm

```{r arm-summary}
# Use group_by() and summarise(): n, n_observed, mean_marker, sd_marker,
# and prop_observed = n_observed / n. Use .groups = "drop".
# Keep missing markers out of the mean and SD, not out of the total n.
```

## 5. Center within arm without losing rows

```{r grouped-transformation}
# Use grouped mutate(): centered_marker = marker - mean(marker, na.rm = TRUE).
# Save as analysis_centered, ungroup(), and compare nrow() before and after.
```

## 6. Arrange and select a review table

```{r review-table}
# Start with analysis_centered. Arrange by arm, then desc(marker).
# Select id, arm, week, marker, observed, marker_band, centered_marker.
```

## 7. Verify with base R

```{r base-verification}
# Use tapply() to check observed counts by arm.
```

Explain the row unit in analysis and in the arm summary.

## Additional practice (optional)

### 1. Recreate the week-4 table with base R

```{r practice-base}
# Use indexing, assignment, and ifelse(). Sort both versions by id to compare.
```

### 2. Missingness by arm and week

```{r practice-missingness}
# Report n, n_observed, n_missing, and prop_observed. Drop grouping.
```

### 3. Category counts within arm

```{r practice-categories}
# Use count(arm, marker_band) on analysis. Check the missing category.
```

### 4. Select helpers and rename

```{r practice-selection}
# Use select(id, arm, starts_with("marker"), observed) on analysis.
# Then rename marker to marker_value.
# Alternatively, use all_of() with a character vector of column names.
```

### 5. Count missing values across columns

```{r practice-across}
# In summarise(), use across(c(week, marker), function(x) sum(is.na(x))).
```

### 6. Summarize by two groups

```{r practice-two-groups}
# Use group_by(arm, week), then summarise() with .groups = "drop".
# Explain the row unit and why you drop grouping.
```

### 7. Diagnose and remove an identical row

```{r practice-duplicate}
# This line is supplied to create a duplicate:
visit_with_duplicate <- rbind(visit, visit[1, ])
# Inspect count(id, week). Confirm the whole row is duplicated.
# Then use distinct() and compare row counts.
# Do not remove legitimate visits just because they share an id.
```
