---
title: "Selected Practical Topics in R"
subtitle: "Current focus: Strings and Regular Expressions · BIOS 6301"
author: "Department of Biostatistics, Vanderbilt University Medical Center"
date: "Session 16"
date-format: "[Session 16]"
format:
  revealjs:
    theme: simple
    css: ../../slides.css
    slide-number: c/t
    show-slide-number: all
    incremental: false
    width: 1280
    height: 720
    fig-align: center
    code-line-numbers: false
    highlight-style: github
    footer: "BIOS 6301 · Session 16 · Selected Practical Topics in R"
    navigation-mode: linear
execute:
  enabled: true
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(knitr)
options(width = 76)

# Synthetic teaching data for BIOS 6301. Not real patients.
set.seed(6301)
n <- 60
trial <- data.frame(
  id        = sprintf("P%03d", 1:n),
  arm       = rep(c("Placebo", "Treatment"), each = n / 2),
  site      = sample(1:4, n, replace = TRUE),
  age       = round(rnorm(n, 61, 9)),
  sex       = sample(c("F", "M"), n, replace = TRUE),
  weight_kg = round(rnorm(n, 82, 15), 1),
  sbp_0     = round(rnorm(n, 148, 14)),
  stringsAsFactors = FALSE
)
trial$sbp_12 <- round(trial$sbp_0 -
  ifelse(trial$arm == "Treatment", rnorm(n, 14, 8), rnorm(n, 3, 8)))

# ---------------------------------------------------------------------------
# Added for this session: the four free-text columns as they arrive from the
# EHR extract. Nothing here is a number yet. What is deliberately broken:
#   dose_raw    drug name, dose, unit and frequency in one field; inconsistent
#               spacing and case; one row with no number at all
#   icd_raw     ICD-10 codes with and without the decimal point, mixed case,
#               leading whitespace, and one row that is a word not a code
#   mrn_raw     the same MRN written four ways, with and without zero padding
#   comorb_raw  free-text comorbidities, mixed separators, synonyms, one
#               negation ("no HTN") and one substring trap ("mild asthma")
# ---------------------------------------------------------------------------
dose_pool <- c(
  "10 mg BID", "2.5mg daily", " 20 MG bid ", "lisinopril 10 mg PO QD",
  "5mg q.d.", "Amlodipine 2.5 mg BID", "40 mg  daily", "one tablet daily",
  "12.5 mg QHS", "100mcg daily", "10 mg  bid", "HCTZ 25 mg daily"
)
icd_pool <- c("I10", "E11.9", "i10", "E119", " I10 ", "Z79.899", "N18.3",
              "HTN", "I10.", "e11.9")
mrn_pool <- c("MRN 0012345", "12345", "mrn-0012345", "0012345", "MRN#12345",
              "MRN 0098761", "98761", "0098761")
comorb_pool <- c("HTN; T2DM", "hypertension, CKD stage 3", "HTN | DM2 | CKD",
                 "mild asthma; HTN", "NIDDM; htn", "no HTN; T2DM",
                 "prior MI; HTN", "none", "HTN;T2DM;CKD")

trial$dose_raw   <- sample(dose_pool,   n, replace = TRUE)
trial$icd_raw    <- sample(icd_pool,    n, replace = TRUE)
trial$mrn_raw    <- sample(mrn_pool,    n, replace = TRUE)
trial$comorb_raw <- sample(comorb_pool, n, replace = TRUE)

# Short vectors for the live demos, so slide output stays readable.
dose   <- c("10 mg BID", "2.5mg daily", " 20 MG bid ",
            "lisinopril 10 mg PO QD", "5mg q.d.", "one tablet daily")
icd    <- c("I10", "E11.9", "i10", "E119", " I10 ", "Z79.899", "HTN")
mrn    <- c("MRN 0012345", "12345", "mrn-0012345", "0012345", "MRN#12345")
comorb <- c("HTN; T2DM", "hypertension, CKD stage 3", "HTN | DM2 | CKD",
            "mild asthma; HTN", "prior MI; HTN", "no HTN; T2DM", "none")
```

## Where we are {.smaller}

| Course arc | Sessions | |
|---|---:|---|
| Setup and workflow | 0–2 |  |
| Data foundations | 3–8 |  |
| Modern tools | 9–10 |  |
| R programming | 11–13 |  |
| **Statistical computing** | **14–16** | **← today** |

Today's deck develops one part of the current arc; later-session tools are named only when they clarify where this idea will be used.


## What you will be able to do {.smaller}

::: {.smaller}
[1]{.objective-num} **Use** the seven `stringr` verbs, and recognise their base R
equivalents in inherited code.

[2]{.objective-num} **Explain** the double-backslash problem and write patterns
that survive it.

[3]{.objective-num} **Build** a regex from character classes, quantifiers,
anchors, groups, and alternation.

[4]{.objective-num} **Predict** greedy versus lazy matching, and pull values out
with capture groups.

[5]{.objective-num} **Extract** structured values — dose, ICD-10, MRN — from
clinical free text.

[6]{.objective-num} **Test** a pattern against known-bad cases before trusting
it.

[7]{.objective-num} **Package** a cleaning rule as a validated function applied
across columns.
:::


## Roadmap for today

| | Block | Minutes |
|---|---|---|
| 1 | Text is data, and it arrives broken | 8 |
| 2 | Seven verbs, two dialects | 18 |
| 3 | Escaping and the double backslash | 8 |
| 4 | Regex proper | 28 |
| 5 | Clinical free text, end to end | 16 |
| 6 | Testing a pattern, and making it a function | 8 |
| 7 | Practice and summary | 4 |

: {tbl-colwidths="[6,64,30]"}

. . .

**This is the most practice-dependent session in the course.** Nobody has ever
learned regex from slides. Today is vocabulary; the lab is where it becomes a
skill.


# 1 · Text is data, and it arrives broken {.section-divider}

## The column that will not be a number {.smaller}

Six rows of a real-shaped `dose` field from an EHR extract:

```{r}
#| label: dose-show
dose
```

. . .

```{r}
#| label: dose-numeric
#| warning: true
as.numeric(dose)
```

. . .

::: {.callout-important appearance="simple"}
The dose is in there. It is just wrapped in a drug name, a unit, a frequency, and
somebody's spacebar.
:::


## Two dialects, one job {.smaller}

::: {style="font-size:0.84em"}
| Task | `stringr` | Base R |
|---|---|---|
| Is the pattern there? | `str_detect(x, p)` | `grepl(p, x)` |
| Which elements? | `str_which(x, p)` | `grep(p, x)` |
| Keep the matches | `str_subset(x, p)` | `grep(p, x, value = TRUE)` |
| Pull the match out | `str_extract(x, p)` | `regmatches(x, regexpr(p, x))` |
| Replace the first | `str_replace(x, p, r)` | `sub(p, r, x)` |
| Replace all | `str_replace_all(x, p, r)` | `gsub(p, r, x)` |
| Split on a pattern | `str_split(x, p)` | `strsplit(x, p)` |
| Characters by position | `str_sub(x, 1, 3)` | `substr(x, 1, 3)` |
| Strip outer whitespace | `str_trim(x)` | `trimws(x)` |
| Pad to a fixed width | `str_pad(x, 7, pad = "0")` | `formatC(x, width = 7, flag = "0")` |
| How long is it? | `str_length(x)` | `nchar(x)` |
| Glue pieces together | `str_c(a, b)` | `paste0(a, b)` |
:::


# 2 · Seven verbs, two dialects {.section-divider}

## Detect: is the pattern there? {.smaller}

```r
# stringr
str_detect(dose, "mg")
str_subset(dose, "BID")
```

```{r}
#| label: detect
# the base R way
grepl("mg", dose)
grep("BID", dose, ignore.case = TRUE, value = TRUE)
```

. . .

::: {.smaller}
`grepl()` returns a **logical vector the same length as the input** — so it goes
straight into `subset()`, `filter()`, or `[`. `grep()` returns positions, or the
matching values with `value = TRUE`.
:::


## Extract: pull the match out {.smaller}

```r
# stringr
str_extract(dose, "[0-9.]+")
```

```{r}
#| label: extract-bad
# the base R way, and it has a trap
m <- regexpr("[0-9.]+", dose)
regmatches(dose, m)
```

. . .

::: {.callout-warning appearance="simple"}
Six strings in, **five** out. `regmatches()` drops non-matches instead of
returning `NA`, so the result no longer lines up with your data frame.
:::

```{r}
#| label: extract-good
out <- rep(NA_character_, length(dose))
out[m > 0] <- regmatches(dose, m)
as.numeric(out)
```


## Replace: first, or all {.smaller}

```r
# stringr
str_replace(dose, " +", " ")       # first run of spaces only
str_replace_all(dose, " +", " ")   # every run
```

```{r}
#| label: replace
# the base R way
sub(" +", " ", dose)
gsub(" +", " ", dose)
```

. . .

::: {.smaller}
`sub()` is `str_replace()`. `gsub()` is `str_replace_all()`. The **`g` is for
global**, and forgetting it is the most common string bug in R.
:::


## Split: one field into many {.smaller}

```r
# stringr
str_split(comorb, " *[;,|] *")
```

```{r}
#| label: split
# the base R way — note what comes back
parts <- strsplit(comorb, " *[;,|] *")
parts[1:3]
lengths(parts)
```


## Slice, trim, pad {.smaller}

```r
# stringr
str_sub(icd, 1, 1)                  # the ICD-10 chapter letter
str_trim(icd)
str_pad(c("12345", "7"), 7, pad = "0")
```

```{r}
#| label: slice
# the base R way
substr(icd, 1, 1)
trimws(icd)
sprintf("%07d", as.integer(c("12345", "7")))
```

. . .

::: {.callout-note appearance="simple"}
`formatC(x, width = 7, flag = "0")` pads with **spaces** when `x` is character.
`sprintf("%07d", ...)` on an integer is the reliable base idiom.
:::


# 3 · Escaping and the double backslash {.section-divider}

## The double-backslash problem {.smaller}

The regex for "a digit" is `\d`. But R will not let you type that.

```{r}
#| label: backslash
#| error: true
p <- "\d+"
```

. . .

```{r}
#| label: backslash2
p <- "\\d+"
p            # what R stores, shown as source
cat(p)       # what the regex engine actually receives
nchar(p)
```


## What has to be escaped {.smaller}

::: {style="font-size:0.86em"}
| You want to match | Write in R | Because |
|---|---|---|
| any digit | `"\\d"` | `\d` is a regex class |
| whitespace | `"\\s"` | `\s` is a regex class |
| a word character | `"\\w"` | letters, digits, `_` |
| a literal dot | `"\\."` | bare `.` means *any character* |
| a literal `(` | `"\\("` | bare `(` opens a group |
| a literal `[` | `"\\["` | bare `[` opens a class |
| a literal backslash | `"\\\\"` | two layers, two doublings |
:::

. . .

```{r}
#| label: dot
c(grepl(".", "no dot here"),
  grepl("\\.", "no dot here"),
  grepl(".", "no dot here", fixed = TRUE))
```


## Raw strings {.smaller}

R 4.0 added a syntax that turns off the backslash doubling entirely.

```{r}
#| label: raw
r"(\d+\s*mg)"          # a raw string: what you see is what R stores
grepl(r"(\d+)", dose)
```

. . .

::: {.smaller}
`r"(...)"` — the delimiter is `r"(` and `)"`. If your pattern contains `)"`, use
`r"[...]"` or `r"{...}"` instead.
:::

. . .

::: {.callout-tip appearance="simple"}
Copy a pattern from regex101 into `r"( )"` and it works unchanged. That is the
whole reason to know this exists.
:::


# 4 · Regex proper {.section-divider}

## A pattern is a description of a set of strings {.smaller}

Read this out loud, left to right:

```r
"^[A-Z][0-9]{2}(\\.[0-9]{1,4})?$"
```

::: {style="font-size:0.9em"}
| Piece | Means |
|---|---|
| `^` | start of the string — nothing before this |
| `[A-Z]` | exactly one capital letter |
| `[0-9]{2}` | exactly two digits |
| `(...)` | a group, so the `?` applies to the whole thing |
| `\\.` | a literal dot |
| `[0-9]{1,4}` | one to four digits |
| `?` | the group is optional |
| `$` | end of the string — nothing after this |
:::

. . .

That is a valid ICD-10 code: **`I10`**, **`E11.9`**, **`Z79.899`**.


## Character classes {.smaller}

::: {style="font-size:0.86em"}
| Class | Matches | Note |
|---|---|---|
| `[abc]` | one of `a`, `b`, `c` | |
| `[a-z]` `[A-Z]` `[0-9]` | a range | |
| `[^0-9]` | anything **except** a digit | `^` inside `[ ]` means "not" |
| `\\d` `\\w` `\\s` | digit · word character · whitespace | |
| `\\D` `\\W` `\\S` | the negations | capital = "not" |
| `.` | any single character | except newline |
| `[[:alpha:]]` `[[:punct:]]` | POSIX classes | portable, verbose |
:::

. . .

```{r}
#| label: classes
gsub("[^0-9]", "", mrn)      # keep only the digits
```


## Quantifiers {.smaller}

::: {style="font-size:0.86em"}
| Quantifier | Means | Example |
|---|---|---|
| `*` | zero or more | `"mg *"` |
| `+` | one or more | `"[0-9]+"` |
| `?` | zero or one | `"\\.?"` |
| `{n}` | exactly *n* | `"[0-9]{2}"` |
| `{n,}` | *n* or more | `"[0-9]{3,}"` |
| `{n,m}` | between *n* and *m* | `"[0-9]{1,4}"` |
:::

. . .

```{r}
#| label: quant
v <- c("E11", "E11.9", "E11.90", "E1", "")
rbind(strict  = grepl("^[A-Z][0-9]{2}$", v),
      relaxed = grepl("^[A-Z][0-9]{2}\\.?[0-9]*$", v))
```


## Anchors and boundaries {.smaller}

```{r}
#| label: anchors
# unanchored: "is it in there anywhere?"
grepl("[0-9]{3}", c("P001", "site 1004 visit", "P01"))
# anchored: "is that the whole string?"
grepl("^P[0-9]{3}$", c("P001", "site 1004 visit", "P01"))
```

. . .

`\\b` is a **word boundary** — the edge between a word character and anything
else.

```{r}
#| label: boundary
rbind(loose = grepl("MI", comorb, ignore.case = TRUE),
      tight = grepl("\\bMI\\b", comorb, ignore.case = TRUE))
```


## Alternation and groups {.smaller}

`|` means **or**. Parentheses control what it applies to.

```{r}
#| label: alternation
grepl("BID|TID|daily", dose, ignore.case = TRUE)
```

. . .

```{r}
#| label: grouping
# the parentheses matter
v <- c("HTN", "HTN x", "x hypertension")
rbind(grouped   = grepl("^(HTN|hypertension)$", v),
      ungrouped = grepl("^HTN|hypertension$",   v))
```

. . .

::: {.callout-warning appearance="simple"}
`^HTN|hypertension$` means *"starts with HTN"* **or** *"ends with
hypertension"* — almost never what was intended.
:::


## Capture groups {.smaller}

Parentheses do a second job: they **remember** what they matched.

```{r}
#| label: capture
m <- regexec("([0-9.]+) *(mg|mcg)\\b", dose, ignore.case = TRUE)
regmatches(dose, m)[1:3]
```

. . .

::: {.smaller}
Element 1 of each result is the whole match; elements 2 and 3 are the two
groups. The `stringr` equivalent is `str_match(dose, ...)`, which returns a
matrix — one column per group — and keeps non-matches as `NA`.
:::


## The card to keep {.smaller}

::: {style="font-size:0.78em"}
| | | | |
|---|---|---|---|
| `^` | start | `\\d` | digit |
| `$` | end | `\\w` | word character |
| `.` | any character | `\\s` | whitespace |
| `*` | 0 or more | `\\b` | word boundary |
| `+` | 1 or more | `[abc]` | one of these |
| `?` | 0 or 1 | `[^abc]` | none of these |
| `{n,m}` | between n and m | `(...)` | group and capture |
| `*?` `+?` | lazy versions | `(?:...)` | group only |
| <code>&#124;</code> | or | `\\1` | backreference |
:::

. . .

::: {.callout-tip appearance="simple"}
Nine rows. That is the whole language for everything you will do this year.
:::


# 5 · Clinical free text, end to end {.section-divider}

## Five steps, every time {.smaller}

```{mermaid}
%%| echo: false
%%| fig-width: 10
flowchart LR
  A["<b>1 · Look</b><br/>table() the<br/>raw values"] --> B["<b>2 · Normalise</b><br/>case, whitespace,<br/>separators"]
  B --> C["<b>3 · Extract</b><br/>pattern with<br/>capture groups"]
  C --> D["<b>4 · Convert</b><br/>to number,<br/>factor, date"]
  D --> E["<b>5 · Validate</b><br/>count the NAs<br/>you created"]
  style A fill:#f4efe4,stroke:#CFAE70
  style B fill:#f4efe4,stroke:#CFAE70
  style D fill:#f4efe4,stroke:#CFAE70
  style C fill:#CFAE70,stroke:#8a7448,color:#fff
  style E fill:#CFAE70,stroke:#8a7448,color:#fff
```

. . .

::: {.smaller}
Step 1 is the one people skip, and step 5 is the one that catches the mistake
step 1 would have prevented.
:::


## Dose strings {.smaller}

```{r}
#| label: dose-look
sort(table(trial$dose_raw), decreasing = TRUE)[1:6]
```

. . .

```{r}
#| label: dose-extract
d <- tolower(trimws(gsub(" +", " ", trial$dose_raw)))     # normalise
m <- regexec("([0-9.]+) ?mg\\b", d)                       # extract
mg <- vapply(regmatches(d, m),
             function(p) if (length(p)) as.numeric(p[2]) else NA_real_,
             numeric(1))
c(n = length(mg), missing = sum(is.na(mg)), median = median(mg, na.rm = TRUE))
```


## ICD-10 codes {.smaller}

```{r}
#| label: icd
icd
norm  <- sub("\\.$", "", toupper(trimws(icd)))    # normalise
valid <- grepl("^[A-Z][0-9][0-9A-Z](\\.[0-9A-Z]{1,4})?$", norm)
data.frame(raw = icd, norm, valid)
```

. . .

::: {.smaller}
`E119` fails — and it should. Some extracts drop the decimal point, and
**`E119` is not a code**; it is `E11.9` with information removed.
:::


## What a regex is bad at {.smaller}

::: {.smaller}
| Problem | Why regex is wrong | Use instead |
|---|---|---|
| Nested or paired structure | It cannot count depth | A parser |
| Dates in mixed formats | The hard part is *meaning*, not shape | `lubridate` — session 5 |
| "Not X" in a clinical note | Negation is language, not syntax | A human, or clinical NLP |
| A fixed, known string | Slower and can misfire | `==`, `%in%`, `fixed = TRUE` |
| A fixed-width field | Position is simpler and safer | `substr()` |
:::

. . .

::: {.callout-note appearance="simple"}
Every row here is a case where the regex **appears to work** on your data and
fails on somebody else's.
:::


# 6 · Testing a pattern, and making it a function {.section-divider}

## Never trust a pattern you have not tried to break {.smaller}

```r
# stringr: highlights every match, in the Viewer
str_view(dose, "[0-9.]+ ?mg")
```

. . .

::: {.smaller}
- **regex101.com** — paste the pattern and the strings; it explains every token
  and shows what each group captured. Set the flavour to **PCRE**.
- Build the pattern against a vector that contains **the cases that should
  fail**, not only the ones that should pass.
:::

. . .

::: {.callout-important appearance="simple"}
A pattern tested only on strings that match is not tested.
:::


## The test-case habit {.smaller}

```{r}
#| label: fn
# Return the milligram dose from a free-text prescription string, or NA.
extract_dose_mg <- function(x) {
  stopifnot(is.character(x))
  x <- tolower(trimws(gsub(" +", " ", x)))
  m <- regexec("([0-9.]+) ?mg\\b", x)
  vapply(regmatches(x, m),
         function(p) if (length(p)) as.numeric(p[2]) else NA_real_,
         numeric(1))
}

stopifnot(
  identical(extract_dose_mg("10 mg BID"),        10),
  identical(extract_dose_mg("2.5mg daily"),      2.5),
  identical(extract_dose_mg("one tablet daily"), NA_real_),
  identical(extract_dose_mg("100mcg daily"),     NA_real_)
)
```


## Flexible practical block {.smaller}

Reserve the final 15–20 minutes for one recurring need identified during the course—for example, a difficult factor, date, import, help-page, or package-use problem.

::: {.question}
What small R problem has cost several students time, and can we reduce it to one reproducible example?
:::


## Summary {.smaller}

- Start with literal string operations; add regex only when the rule is patterned.
- R string escaping and regex escaping are separate layers.
- Anchors distinguish finding a pattern from validating a complete value.
- Test near misses and missing values, not only examples that should match.
- Turn a validated pattern into reusable, documented code.

. . .

**Next connection:** Use the remaining class time for the practical R issue selected from student needs.

