---
title: "Iteration and Vectorization"
subtitle: "BIOS 6301 · Introduction to Statistical Computing"
author: "Department of Biostatistics, Vanderbilt University Medical Center"
date: "Session 13"
date-format: "[Session 13]"
format:
  revealjs:
    theme: simple
    css: ../../slides.css
    slide-number: c/t
    show-slide-number: all
    incremental: false
    width: 1280
    height: 720
    fig-align: center
    code-line-numbers: false
    highlight-style: github
    footer: "BIOS 6301 · Session 13 · Iteration and Vectorization"
    navigation-mode: linear
execute:
  enabled: true
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(ggplot2)
library(knitr)

# Synthetic teaching data for BIOS 6301. Not real patients.
set.seed(6301)
n <- 60
trial <- data.frame(
  id        = sprintf("P%03d", 1:n),
  arm       = rep(c("Placebo", "Treatment"), each = n / 2),
  site      = sample(1:4, n, replace = TRUE),
  age       = round(rnorm(n, 61, 9)),
  sex       = sample(c("F", "M"), n, replace = TRUE),
  weight_kg = round(rnorm(n, 82, 15), 1),
  sbp_0     = round(rnorm(n, 148, 14)),
  stringsAsFactors = FALSE
)
trial$sbp_12 <- round(trial$sbp_0 -
  ifelse(trial$arm == "Treatment", rnorm(n, 14, 8), rnorm(n, 3, 8)))

# Added for this session: a 6-week visit, so there are three outcome columns
# to iterate over and a three-column matrix for the apply() margins demo.
trial$sbp_6 <- round(trial$sbp_0 -
  ifelse(trial$arm == "Treatment", rnorm(n, 9, 7), rnorm(n, 2, 7)))

outcomes <- c("sbp_0", "sbp_6", "sbp_12")
sbp_mat  <- as.matrix(trial[, outcomes])
rownames(sbp_mat) <- trial$id

# The same extract, re-run a month later. Nothing about the query changed;
# site 3's most recent enrolments simply happen to all be female. This is the
# only difference between trial and trial_jun, and it is the whole sapply demo.
trial_jun <- trial
trial_jun$sex[trial_jun$site == 3] <- "F"

theme_set(theme_minimal(base_size = 15))
```

## Where we are {.smaller}

| Course arc | Sessions | |
|---|---:|---|
| Setup and workflow | 0–2 |  |
| Data foundations | 3–8 |  |
| Modern tools | 9–10 |  |
| **R programming** | **11–13** | **← today** |
| Statistical computing | 14–16 |  |

Today's deck develops one part of the current arc; later-session tools are named only when they clarify where this idea will be used.


## What you will be able to do {.smaller}

::: {.smaller}
[1]{.objective-num} **Write** a `for` loop that pre-allocates its result, and
say what it costs when you do not.

[2]{.objective-num} **Use** `seq_along()` and explain what `1:length(x)` does
when `x` is empty.

[3]{.objective-num} **Explain** what "vectorized" means, and why it is fast.

[4]{.objective-num} **Choose** between `lapply()`, `sapply()`, `vapply()`,
`apply()`, `mapply()`, and `tapply()`.

[5]{.objective-num} **Say** why `sapply()` is unsafe in code that matters, and
write the `vapply()` that fixes it.

[6]{.objective-num} **Read** `purrr::map()` code and translate it to base R,
and back.

[7]{.objective-num} **Recognise** a problem that genuinely needs a loop.
:::


## Roadmap for today

| | Block | Minutes |
|---|---|---|
| 1 | Doing the same thing many times | 6 |
| 2 | `for`, `while`, and pre-allocation | 18 |
| 3 | What "vectorized" actually means | 12 |
| 4 | The apply family | 25 |
| 5 | Head to head: base and `purrr` | 15 |
| 6 | Choosing, and when a loop is right | 9 |
| 7 | Practice and summary | 5 |

: {tbl-colwidths="[6,64,30]"}

. . .

**In lab this week** you turn one function into a table of results across every
outcome in the dataset.


# 1 · Doing the same thing many times {.section-divider}

## earlier in the course's function, twelve times {.smaller}

You finished session 11 with this:

```r
summarise_by_arm(trial, sbp_0)
summarise_by_arm(trial, sbp_6)
summarise_by_arm(trial, sbp_12)
summarise_by_arm(trial, weight_kg)
summarise_by_arm(trial, age)
# ... and eight more outcomes in the final project
```

. . .

::: {.smaller}
The duplication did not disappear. It **moved up one level** — from the body of
the block to the list of calls.
:::

. . .

::: {.callout-important appearance="simple"}
A function names *what* you are doing. Iteration names *what you are doing it
to*.
:::


## Three ways to say "for each" {.smaller}

::: {style="font-size:0.9em"}
| | Looks like | Gives you |
|---|---|---|
| **Loop** | `for (v in outcomes) { ... }` | Nothing. You manage the result yourself |
| **Apply family** | `lapply(outcomes, f)` | A list, one element per input |
| **`purrr`** | `map(outcomes, f)` | A list, with a stricter contract |
| **Vectorized** | `sqrt(x)`, `x > 140` | The answer, with no iteration written at all |
:::

. . .

::: {.smaller}
The first three are the same idea with different bookkeeping. The fourth is a
different idea, and where it applies it beats all of them.
:::


# 2 · `for`, `while`, and pre-allocation {.section-divider}

## Anatomy of a `for` loop {.smaller}

```{r}
#| label: for-anatomy
for (v in outcomes) {
  cat(v, ":", round(mean(trial[[v]]), 1), "mmHg\n")
}
```

. . .

::: {.smaller}
- `v` takes each element of `outcomes` **in turn**. It is a real variable and it
  survives the loop.
- `trial[[v]]` — double brackets, because `v` holds a *name*, not a column.
- This loop **prints**. It does not produce an object. That is the next slide,
  and it is where the trouble is.
:::


## Pre-allocate, do not grow {.smaller}

Same answer, two ways. Watch the clock.

```{r}
#| label: prealloc-bench
grow <- function(k) {
  out <- c()                                  # grows on every iteration
  for (i in seq_len(k)) out <- c(out, sqrt(i))
  out
}

prealloc <- function(k) {
  out <- numeric(k)                           # the right size, once
  for (i in seq_len(k)) out[i] <- sqrt(i)
  out
}

system.time(grow(50000))["elapsed"]
system.time(prealloc(50000))["elapsed"]
```


## Why growing is quadratic {.smaller}

`out <- c(out, x)` does not append. It **allocates a new vector and copies the
old one into it**, every single iteration.

```{r}
#| label: quadratic
#| echo: false
grow2 <- function(k) {
  out <- c()
  for (i in seq_len(k)) out <- c(out, sqrt(i))
  out
}
elapsed <- function(e) round(system.time(e)["elapsed"], 2)
data.frame(
  k       = c(20000, 40000, 80000),
  seconds = c(elapsed(grow2(20000)),
              elapsed(grow2(40000)),
              elapsed(grow2(80000)))
) |> kable()
```

. . .

::: {.smaller}
Doubling `k` costs roughly **four times** as much. Building a vector of length
`k` this way copies about `k²/2` values.
:::

. . .

::: {.callout-warning appearance="simple"}
`rbind()`-ing rows onto a data frame inside a loop is the same bug, and worse —
a data frame copy is far more expensive than a vector copy.
:::


## `seq_along()`, and the empty-vector trap {.smaller}

```{r}
#| label: empty-trap
sbp <- numeric(0)          # a filter that matched nobody
1:length(sbp)
```

. . .

```{r}
#| label: empty-trap-loop
for (i in 1:length(sbp)) {
  cat("iteration", i, "gives value:", sbp[i], "\n")
}
```

. . .

```{r}
#| label: empty-trap-fix
seq_along(sbp)             # the fix: length zero means zero iterations
```


## `while`: when you do not know how many {.smaller}

```{r}
#| label: while-newton
# Newton's method for sqrt(2): iterate until the step is small enough.
x    <- 1
iter <- 0

while (abs(x^2 - 2) > 1e-10 && iter < 100) {
  x    <- x - (x^2 - 2) / (2 * x)
  iter <- iter + 1
}

c(root = x, iterations = iter)
```

. . .

::: {.callout-warning appearance="simple"}
**Always bound the loop.** `iter < 100` is not decoration — without it, a
condition that never becomes false hangs your session with no error and no
output.
:::


# 3 · What "vectorized" actually means {.section-divider}

## The loop is still there. It is in C {.smaller}

```{mermaid}
%%| echo: false
%%| fig-width: 10.5
flowchart LR
  A["<b>sum(x)</b>"] --> B["<b>one</b> call<br/>into C"]
  B --> C["1,000,000<br/>additions,<br/>compiled"]
  D["<b>for (i in ...)</b><br/>s &lt;- s + x[i]"] --> E["<b>1,000,000</b> trips<br/>through the<br/>R interpreter"]
  E --> F["look up s,<br/>look up x, index,<br/>add, assign —<br/>every time"]
  style A fill:#CFAE70,stroke:#8a7448,color:#fff
  style D fill:#f4efe4,stroke:#CFAE70
  style C fill:#f4efe4,stroke:#CFAE70
  style F fill:#f4efe4,stroke:#CFAE70
```

. . .

::: {.smaller}
"Vectorized" does not mean *no loop*. It means **the loop is not written in R**.
:::


## `for` loops are not slow. Growing is {.smaller}

```{r}
#| label: bench-three
y  <- rnorm(5e5)
tt <- \(expr) system.time(expr)[["elapsed"]]      # seconds, unnamed

c(`for loop`   = tt({
                   out <- numeric(length(y))
                   for (i in seq_along(y)) out[i] <- y[i]^2
                 }),
  `vapply()`   = tt(vapply(y, \(v) v^2, numeric(1))),
  `vectorized` = tt(y^2))
```

. . .

::: {.smaller}
The functional is **not faster than the loop** — it is often slower. Only the
last row is a different kind of thing.
:::


## The vectorized answers you already have {.smaller}

::: {style="font-size:0.84em"}
| Instead of looping to... | Write |
|---|---|
| build a new column from an old one | `trial$high <- trial$sbp_0 > 140` |
| choose per row | `ifelse()`, `if_else()`, `case_when()` |
| total or average a column | `sum()`, `mean()`, `cumsum()` |
| average across visits, per patient | `rowMeans()`, `rowSums()` |
| average one visit, across patients | `colMeans()`, `colSums()` |
| element-wise max of two vectors | `pmax()`, `pmin()` |
| count each value | `table()` |
| test membership | `x %in% y` |
| find where a condition holds | `which()` |
| running total | `cumsum()`, `cumprod()`, `cummax()` |
:::

. . .

::: {.callout-tip appearance="simple"}
Before writing a loop, ask whether a function already does it. Ten minutes of
searching beats an hour of debugging an index.
:::


# 4 · The apply family {.section-divider}

## `lapply()`: a list in, a list out {.smaller}

```{r}
#| label: lapply-demo
summary_of <- function(x) {
  round(c(n = sum(!is.na(x)), mean = mean(x), sd = sd(x)), 1)
}

results <- lapply(trial[outcomes], summary_of)
results
```


## `sapply()`: convenience, and what it costs {.smaller}

`sapply()` is `lapply()` plus a guess about what shape you wanted.

```{r}
#| label: sapply-good
sapply(trial[outcomes], summary_of)
```

. . .

```{r}
#| label: sapply-sex
# For each site, which sexes are enrolled?
sapply(split(trial$sex, trial$site), unique)
```


## The day the data changed {.smaller}

Same code. Same query. The extract re-run a month later:

```{r}
#| label: sapply-bad
sapply(split(trial_jun$sex, trial_jun$site), unique)
```

. . .

```{r}
#| label: sapply-bad-class
class(sapply(split(trial$sex,     trial$site), unique))
class(sapply(split(trial_jun$sex, trial_jun$site), unique))
```

. . .

::: {.callout-important appearance="simple"}
A matrix on Tuesday, a list on Wednesday. Nothing downstream was written to
handle both.
:::


## `vapply()`: declare the shape {.smaller}

`FUN.VALUE` is a template. It says what one answer looks like.

```{r}
#| label: vapply-good
vapply(trial[outcomes], summary_of, numeric(3))
```

. . .

```{r}
#| label: vapply-error
#| error: true
vapply(split(trial_jun$sex, trial_jun$site), unique, character(2))
```

. . .

::: {.smaller}
The error names the element that broke the contract: `FUN(X[[3]])`. That is
site 3.
:::


## `apply()`: margins on a matrix {.smaller}

::: columns
::: {.column width="49%"}
```{r}
#| label: apply-rows
# MARGIN = 1: one answer per ROW
head(apply(sbp_mat, 1, max), 4)
```
:::

::: {.column width="49%"}
```{r}
#| label: apply-cols
# MARGIN = 2: one answer per COLUMN
round(apply(sbp_mat, 2, mean), 1)
```
:::
:::

. . .

::: {.smaller}
`sbp_mat` is 60 × 3: one row per patient, one column per visit. **`MARGIN` is
the dimension you keep.**
:::

. . .

```r
# and when the function is mean or sum, do not use apply() at all
rowMeans(sbp_mat)   colMeans(sbp_mat)   # written in C, much faster
```


## `apply()` on a data frame: the coercion trap {.smaller}

```{r}
#| label: apply-coerce
apply(trial[, c("id", "sbp_0", "sbp_12")], 1, max) |> head(4)
```

. . .

::: {.smaller}
Those are patient **IDs**, not blood pressures. `apply()` needs a matrix, so it
called `as.matrix()` — and a matrix holds one type, so every number became
`"P001"`-style text.
:::

. . .

::: {.callout-warning appearance="simple"}
Session 3's coercion rules, arriving through a new door. `apply()` on a data
frame with any character column silently converts **everything** to character.
:::


## `mapply()` and `Map()`: two vectors at once {.smaller}

When you need to walk **two** things in parallel:

```{r}
#| label: mapply-demo
labels <- c("Baseline", "6 weeks", "12 weeks")

mapply(
  \(col, lab) sprintf("%s: %.1f mmHg", lab, mean(trial[[col]])),
  outcomes, labels
)
```

. . .

::: {.smaller}
`Map()` is the same thing without simplification — it is to `mapply()` what
`lapply()` is to `sapply()`, and it is the safer default.
:::


## `tapply()`: split, apply, combine {.smaller}

You met this in session 4. Look at it again as a **functional**:

```{r}
#| label: tapply-demo
round(tapply(trial$sbp_12, trial$arm, mean), 1)
```

. . .

```{r}
#| label: tapply-two
round(tapply(trial$sbp_12, list(trial$arm, trial$site), mean), 1)
```

. . .

::: {.smaller}
```r
# the tidyverse way
trial |> group_by(arm) |> summarise(mean = mean(sbp_12))
```
:::


## The family on one slide {.smaller}

::: {style="font-size:0.82em"}
| Function | Input | Output | Use it |
|---|---|---|---|
| `lapply(x, f)` | vector or list | **always a list** | The safe default |
| `sapply(x, f)` | vector or list | vector, matrix, *or list* | Console only |
| `vapply(x, f, FUN.VALUE)` | vector or list | the shape you declared | **In scripts** |
| `apply(m, MARGIN, f)` | matrix | vector or matrix | Rows or columns of a matrix |
| `mapply(f, x, y)` | two or more vectors | simplified, like `sapply()` | Parallel iteration |
| `Map(f, x, y)` | two or more vectors | always a list | Parallel, safely |
| `tapply(x, g, f)` | vector + grouping | vector or matrix | Grouped summary |
| `Reduce(f, x)` | vector or list | one value | Combining, cumulatively |
| `Filter(f, x)` | vector or list | a subset | Keeping what matches |
:::


# 5 · Head to head: base and `purrr` {.section-divider}

## `map()` is `lapply()` with better manners {.smaller}

::: columns
::: {.column width="49%"}
### base R

```{r}
#| label: hh-lapply
res <- lapply(trial[outcomes],
              summary_of)
res$sbp_12
```
:::

::: {.column width="49%"}
### `purrr`

```r
library(purrr)

res <- map(trial[outcomes],
           summary_of)
res$sbp_12
```
:::
:::

. . .

::: {.smaller}
Same idea, same result. `map()` puts the data first — so it pipes — and is the
only member of a family whose names all mean something.
:::

::: {.callout-note appearance="simple"}
`purrr` is not installed in this container, so the right-hand column is shown,
not run. Everything on the left is live.
:::


## Typed variants: `map_dbl()` and friends {.smaller}

::: columns
::: {.column width="49%"}
### base R

```{r}
#| label: hh-vapply
vapply(trial[outcomes],
       mean, numeric(1))
```
:::

::: {.column width="49%"}
### `purrr`

```r
map_dbl(trial[outcomes], mean)
```
:::
:::

. . .

::: {style="font-size:0.86em"}
| Want back | base R | `purrr` |
|---|---|---|
| a list | `lapply(x, f)` | `map(x, f)` |
| a double vector | `vapply(x, f, numeric(1))` | `map_dbl(x, f)` |
| a character vector | `vapply(x, f, character(1))` | `map_chr(x, f)` |
| a logical vector | `vapply(x, f, logical(1))` | `map_lgl(x, f)` |
| an integer vector | `vapply(x, f, integer(1))` | `map_int(x, f)` |
| a data frame of rows | `do.call(rbind, lapply(...))` | `map_dfr(x, f)` |
:::


## The `\(x)` shorthand {.smaller}

`\(x)` is base R 4.1, not tidyverse. It works everywhere a function does.

```{r}
#| label: lambda-base
vapply(trial[outcomes], \(x) round(mean(x), 1), numeric(1))
```

. . .

```{r}
#| label: lambda-pvals
# One t-test per outcome, in one line.
vapply(outcomes, \(v) t.test(trial[[v]] ~ trial$arm)$p.value, numeric(1)) |>
  signif(3)
```

. . .

::: {.smaller}
```r
# purrr, with the same shorthand and its own ~ .x form
map_dbl(outcomes, \(v) t.test(trial[[v]] ~ trial$arm)$p.value)
map_dbl(outcomes, ~ t.test(trial[[.x]] ~ trial$arm)$p.value)
```
:::


## `map2()`, `pmap()`, and the rest {.smaller}

::: {style="font-size:0.86em"}
| Job | base R | `purrr` |
|---|---|---|
| Two vectors in parallel | `mapply(f, x, y)` / `Map(f, x, y)` | `map2(x, y, f)` |
| Many vectors in parallel | `mapply(f, x, y, z)` | `pmap(list(x, y, z), f)` |
| Side effects only — plots, files | `for` / `invisible(lapply())` | `walk(x, f)` |
| Element and its name | `mapply(f, x, names(x))` | `imap(x, f)` |
| Keep what matches | `Filter(f, x)` | `keep(x, f)` |
| Combine cumulatively | `Reduce(f, x)` | `reduce(x, f)` |
| Fail softly, keep going | `try()` inside the function | `possibly(f, NA)` |
:::

. . .

::: {.smaller}
Note the argument order: `purrr` is **data first, function second**, always.
Base R is not.
:::


## The same job, twice {.smaller}

**Task.** For every outcome, the treatment-minus-placebo difference in means.

::: columns
::: {.column width="49%"}
```{r}
#| label: hh-base-final
vapply(outcomes, \(v) {
  m <- tapply(trial[[v]],
              trial$arm, mean)
  round(m[["Treatment"]] -
        m[["Placebo"]], 1)
}, numeric(1))
```
:::

::: {.column width="49%"}
```r
map_dbl(outcomes, \(v) {
  trial |>
    group_by(arm) |>
    summarise(m = mean(.data[[v]])) |>
    pull(m) |>
    diff() |>
    round(1)
}) |> set_names(outcomes)
```
:::
:::

. . .

::: {.smaller}
Same numbers. `vapply()` names its result for free; `map_dbl()` needs
`set_names()`. And `v` holds a **string**, so it is `.data[[v]]` — not last
week's curly-curly, which wants a bare column name.
:::


# 6 · Choosing, and when a loop is right {.section-divider}

## Which one do I want? {.smaller}

::: {style="font-size:0.88em"}
| If... | Use |
|---|---|
| A vectorized function already exists | **That function.** Nothing else is close |
| One value per element, type known | `vapply()` / `map_dbl()` |
| Results have different shapes | `lapply()` / `map()` |
| You are looping for a side effect | `for` loop, or `walk()` |
| Step *i* depends on step *i−1* | **`for` or `while`.** No functional can help |
| You are exploring at the console | Whatever is quickest to type |
:::

. . .

::: {.callout-note appearance="simple"}
Readability is the tie-breaker, not speed. Slide 16 showed that the timing
difference between a loop and a functional is usually noise.
:::


## The loop you cannot remove {.smaller}

**Iterative dependence**: step *i* needs the result of step *i−1*.

```{r}
#| label: newton-mle
# Newton-Raphson. Each step is computed FROM the previous estimate.
x     <- 1
trace <- numeric(4)

for (i in seq_len(4)) {
  x        <- x - (x^2 - 2) / (2 * x)
  trace[i] <- x
}

round(trace, 8)
```

. . .

::: {.smaller}
There is nothing to hand to `lapply()`. The input to iteration 3 does not exist
until iteration 2 has run. **MCMC, EM, gradient descent, and every iterative
optimiser have this shape.**
:::


## Accumulating results without growing {.smaller}

The pattern for "one row of results per iteration":

```{r}
#| label: accumulate
#| results: hold
res <- vector("list", length(outcomes))     # pre-allocated LIST
names(res) <- outcomes

for (v in outcomes) {
  res[[v]] <- data.frame(
    outcome = v,
    mean    = round(mean(trial[[v]]), 1),
    sd      = round(sd(trial[[v]]), 1)
  )
}

do.call(rbind, res)                          # combine ONCE, at the end
```


## Summary {.smaller}

- Vectorize when the vector expression states the operation directly.
- Use a loop when the algorithm has state or the steps need to remain explicit.
- Preallocate output; repeated growth is the expensive pattern.
- Prefer functionals with a stable output contract for repeated independent work.
- Verify equivalence before benchmarking or optimizing.

. . .

**Next connection:** Session 14 uses these functions and iteration patterns to build simulation studies.

