---
title: "RStudio and Reproducible Documents"
subtitle: "BIOS 6301 · Introduction to Statistical Computing"
author: "Department of Biostatistics, Vanderbilt University Medical Center"
date: "Session 1"
date-format: "[Session 1]"
format:
  revealjs:
    theme: simple
    css: ../../slides.css
    slide-number: c/t
    show-slide-number: all
    incremental: false
    width: 1280
    height: 720
    fig-align: center
    code-line-numbers: false
    highlight-style: github
    footer: "BIOS 6301 · Session 1 · RStudio and Reproducible Documents"
    navigation-mode: linear
execute:
  enabled: true
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false
set.seed(6301)
# Small synthetic trial dataset used for the live examples in this deck.
trial <- data.frame(
  id     = sprintf("P%03d", 1:60),
  arm    = rep(c("Placebo", "Treatment"), each = 30),
  age    = round(rnorm(60, 61, 9)),
  sex    = sample(c("F", "M"), 60, replace = TRUE),
  sbp_0  = round(rnorm(60, 148, 14)),
  sbp_12 = NA_real_
)
trial$sbp_12 <- round(trial$sbp_0 -
  ifelse(trial$arm == "Treatment", rnorm(60, 14, 8), rnorm(60, 3, 8)))
```

## Where we are {.smaller}

| Course arc | Sessions | |
|---|---:|---|
| **Setup and workflow** | **0–2** | **← today** |
| Data foundations | 3–8 |  |
| Modern tools | 9–10 |  |
| R programming | 11–13 |  |
| Statistical computing | 14–16 |  |

Today's deck develops one part of the current arc; later-session tools are named only when they clarify where this idea will be used.


## What you will be able to do {.smaller}

::: {.smaller}
[1]{.objective-num} **Navigate** the four RStudio panes and distinguish R from
the RStudio interface.

[2]{.objective-num} **Write and run** saved commands in an `.R` script rather
than relying on console history.

[3]{.objective-num} **Build** an `.Rmd` document containing YAML, Markdown,
R chunks, inline R, and appropriate knitr chunk options.

[4]{.objective-num} **Apply** that document workflow in `.qmd`, using Quarto's
`format:` key and `#|` cell-option syntax.

[5]{.objective-num} **Choose** a deliberate path strategy for project data and
for large or protected data stored elsewhere.

[6]{.objective-num} **Explain** why reproducibility is part of statistical
quality rather than end-of-project clerical work.

[7]{.objective-num} **Verify** that a document renders from a clean R session.
:::


## Roadmap for today

| | Block | Minutes |
|---|---|---|
| 1 | RStudio workbench and `.R` scripts | 20 |
| 2 | R Markdown documents and knitr options | 25 |
| 3 | What Quarto adds or changes | 15 |
| 4 | Reproducible projects, paths, and the Duke case | 20 |
| 5 | Clean-session rendering and summary | 10 |

: {tbl-colwidths="[6,64,30]"}

. . .

**In lab this session** you build and render your first report. Today is the
material you need in order to do that.


## Two useful R styles, one analytical standard {.smaller}

- Sessions 3–5 emphasize **base R** so everyone can read inherited R code.
- Sessions 6–7 teach the **tidyverse** and compare equivalent operations.
- When an exercise does not require a particular tool, **students may choose** the clearer approach.
- Correctness, statistical meaning, verification, and reproducibility matter more than dialect.


# 1 · RStudio and the workbench {.section-divider}

## R and RStudio are not the same {.smaller}

::: columns
::: {.column width="48%"}
### R

- The statistical programming language
- Performs calculations
- Stores objects in a running R session
- Can run without RStudio
:::

::: {.column width="48%"}
### RStudio

- An integrated development environment, or IDE
- Provides an editor, Console, plots, Help, projects, and Git tools
- Sends instructions to R
- Does not replace R
:::
:::

. . .

::: {.callout-note appearance="simple"}
If RStudio opens but R is not installed or configured, the interface exists but
there is no R engine available to execute your code.
:::


## The four primary panes {.smaller}

::: {style="font-size:0.88em"}
| Pane | What it contains | Use it for |
|---|---|---|
| **Source** | Saved `.R`, `.Rmd`, and `.qmd` files | Writing work that should persist |
| **Console** | The current interactive R session | Short experiments and immediate feedback |
| **Environment** | Objects currently held in memory; History and Git tabs | Inspecting temporary session state |
| **Output** | Files, Plots, Packages, Help, Viewer | Inspecting files, results, and documentation |
:::

. . .

Pane locations can be customized. Learn their **functions**, not merely their
default positions.


## The Console is a conversation with the current session {.smaller}

```r
2 + 2

x <- c(4, 7, 9)
mean(x)
```

::: columns
::: {.column width="48%"}
### Useful for

- Testing one expression
- Asking for help: `?mean`
- Inspecting an object: `str(x)`
:::

::: {.column width="48%"}
### Not a durable record

- Commands may be incomplete or out of order
- Objects disappear when the session ends
- Console history does not explain analytical decisions
:::
:::


## Your first saved `.R` script {.smaller}

**File → New File → R Script**, then save as `analysis.R`.

```r
# Summarize the synthetic clinic data
clinic <- read.csv("data/clinic_summary.csv")

n_records <- nrow(clinic)
mean_age <- mean(clinic$age, na.rm = TRUE)

n_records
mean_age
```

An `.R` file contains R instructions and comments. It does not automatically
contain formatted narrative or the results produced when the code runs.


## Run a line, a selection, or the entire script {.smaller}

::: {style="font-size:0.88em"}
| Action | RStudio command | Appropriate use |
|---|---|---|
| Current line or selection | {{< kbd Ctrl-Enter >}} / {{< kbd Cmd-Enter >}} | Develop and inspect small steps |
| Entire script | **Source** button or `source("analysis.R")` | Test whether the saved script is complete |
| Restart R | {{< kbd Ctrl-Shift-F10 >}} / {{< kbd Cmd-Shift-0 >}} | Remove hidden session state |
:::

. . .

::: {.callout-important appearance="simple"}
Running selected lines answers “does this step work now?” Sourcing after a
restart answers “does the saved script contain everything it needs?”
:::


## Console versus `.R` script {.smaller}

| | Console | `.R` script |
|---|---|---|
| Immediate feedback | Yes | Yes, when sent to R |
| Saved deliberately | No | Yes |
| Easy to rerun in order | Fragile | Yes |
| Contains formatted report text | No | No |
| Best role | Exploration | Reusable analysis instructions |

. . .

The next step is to connect saved instructions with the explanation and results
they produce.


# 2 · R Markdown documents {.section-divider}

## R Markdown is the practical starting point {.smaller}

You will meet `.Rmd` constantly—in existing analyses, package vignettes,
collaborators' reports, and online examples. It remains a supported and widely
used format for R-based reporting.

. . .

::: {.callout-note appearance="simple"}
An R Markdown document combines narrative, executable R code, and computed
results in one inspectable source file.
:::

. . .


## Anatomy of an `.Rmd` {.smaller}

````markdown
---
title: "Baseline summary"
output: html_document
---

The report contains **prose** and computed results.

```{r summary, echo=FALSE}
clinic <- read.csv("data/clinic_summary.csv")
mean(clinic$age)
```
````

::: {.smaller}
Three layers: **YAML** says what to create; **Markdown** explains the analysis;
an **R chunk** performs the computation. `rmarkdown`, `knitr`, and Pandoc turn
the source into the report.
:::


## What `.Rmd` adds to an `.R` script {.smaller}

| `.R` script | `.Rmd` document |
|---|---|
| R code and comments | Explanatory Markdown plus executable R chunks |
| Results appear in Console or plot device | Results are placed into a rendered report |
| Usually one stream of commands | Organized into named analytical sections |
| Excellent for reusable functions and pipelines | Excellent for an analysis narrative and report |

. . .

These formats complement one another. A report may `source()` functions from an
`.R` file while keeping the statistical explanation in `.Rmd`.


## R Markdown YAML: what should be produced? {.smaller}

```yaml
---
title: "Synthetic clinic summary"
author: "Student name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
---
```

::: {.smaller}
- YAML is between the opening and closing `---` lines.
- R Markdown uses the `output:` key.
- Indentation expresses nesting; spaces after colons matter.
- Start with HTML because it requires no LaTeX installation.
:::


## Markdown supplies the narrative structure {.smaller}

```markdown
# Main heading

## Data check

The analysis uses **synthetic data**.

- first item
- second item

[R documentation](https://www.r-project.org/)
```

. . .

Markdown is plain text. It remains readable even before the document is knitted.


## Anatomy of an R Markdown code chunk {.smaller}

````markdown
```{r age-summary, echo=FALSE, warning=FALSE}
mean_age <- mean(clinic$age, na.rm = TRUE)
mean_age
```
````

::: {style="font-size:0.88em"}
| Part | Meaning |
|---|---|
| `r` | Execute the chunk with R |
| `age-summary` | A unique, informative chunk label |
| `echo=FALSE` | Run the code but hide the source in the report |
| `warning=FALSE` | Do not print warnings in the report |
| Lines inside the fences | Ordinary R code, executed from top to bottom |
:::


## Three options students commonly confuse {.smaller}

::: {style="font-size:0.88em"}
| R Markdown option | Does the code run? | Is code shown? | Are results shown? |
|---|:---:|:---:|:---:|
| default | Yes | Yes | Yes |
| `echo=FALSE` | Yes | **No** | Yes |
| `eval=FALSE` | **No** | Yes | No |
| `include=FALSE` | Yes | No | **No** |
:::

. . .

::: {.callout-note appearance="simple"}
`include=FALSE` is ideal for a setup chunk: packages and data are loaded, but the
setup code and all its output are omitted from the report.
:::


## Useful R Markdown chunk options {.smaller}

::: {style="font-size:0.82em"}
| Option | Example | Effect |
|---|---|---|
| Show source | `echo=FALSE` | Hide code but retain its results |
| Execute | `eval=FALSE` | Display code without running it |
| Include anything | `include=FALSE` | Run but omit code, results, plots, messages, and warnings |
| Text output | `results='hide'` | Run but hide printed text output |
| Messages | `message=FALSE` | Hide package and function messages |
| Warnings | `warning=FALSE` | Hide warnings after you have investigated them |
| Figure size | `fig.width=7, fig.height=4` | Set figure dimensions in inches |
| Figure caption | `fig.cap='Marker by group.'` | Add a caption to the generated figure |
:::

. . .

::: {.callout-warning appearance="simple"}
Never suppress a warning merely to make the report look clean. Read it,
understand it, and then decide whether suppression is appropriate.
:::


## A setup chunk establishes document-wide defaults {.smaller}

````markdown
```{r rmd-setup-example, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4
)

clinic <- read.csv("data/clinic_summary.csv")
```
````

Individual chunks can override these defaults in their own headers.


## Inline R keeps numbers connected to calculations {.smaller}

Write this in the R Markdown prose:

```markdown
The dataset contains `r knitr::inline_expr("nrow(clinic)")` records, with a mean
age of `r knitr::inline_expr("round(mean(clinic$age), 1)")` years.
```

. . .

It renders as a sentence containing the current calculated values.

::: {.callout-important appearance="simple"}
If a number comes from the data, calculate it in code or inline R. Do not type a
copy of the result into the narrative.
:::


## Knit means process the complete document {.smaller}

In RStudio, use the **Knit** button or run:

```r
rmarkdown::render("report.Rmd")
```

::: {.incremental}
1. Read the YAML.
2. Execute R chunks in document order.
3. Insert code, results, figures, and inline values as requested.
4. Use Pandoc to produce the selected output format.
5. Write a generated `.html`, `.docx`, or `.pdf` file.
:::


# 3 · Quarto {.section-divider}

## Apply the same document model in `.qmd` {.smaller}

These ideas do **not** change:

::: {.incremental}
- Markdown provides headings, prose, lists, and links.
- Fenced `{r}` chunks contain ordinary R code.
- `knitr` executes R chunks from top to bottom.
- Inline R inserts calculated values into prose.
- Source remains separate from generated output.
- A complete render should not depend on objects left in the Environment pane.
:::

. . .

Quarto changes selected authoring and publishing syntax; it does not change the
R language or the logic of the analysis.


## The Quarto-specific changes for today {.smaller}

::: {style="font-size:0.86em"}
| Feature | In the `.Rmd` block | In a `.qmd` |
|---|---|---|
| File extension | `.Rmd` | `.qmd` |
| Output key | `output: html_document` | `format: html` |
| Button in RStudio | **Knit** | **Render** |
| Local option style | `{r fig1, echo=FALSE}` | `#\| label: fig1` and `#\| echo: false` |
| Figure dimensions | `fig.width=7` | `fig-width: 7` |
| Document defaults | `knitr::opts_chunk$set()` | `execute:` in YAML, or knitr defaults |
:::

. . .

The course will normally use Quarto for new reports, while expecting students to
read and maintain R Markdown correctly.


## What Quarto does {.smaller}

```{mermaid}
%%| echo: false
%%| fig-width: 10
flowchart LR
  QMD["<b>report.qmd</b><br/>prose + code"] --> K["<b>knitr</b><br/>runs the R,<br/>captures output"]
  K --> MD["markdown<br/>+ results"]
  MD --> P["<b>pandoc</b>"]
  P --> H["HTML"]
  P --> PDF["PDF"]
  P --> W["Word"]
  P --> S["slides"]
  style QMD fill:#f4efe4,stroke:#CFAE70
  style K fill:#CFAE70,stroke:#8a7448,color:#fff
  style P fill:#CFAE70,stroke:#8a7448,color:#fff
```

. . .

::: {.smaller}
Two things follow from this picture, and both matter:

- **knitr executes the document in order during rendering.** Anything the report
  needs must be created or loaded by saved source.
- **The source is one file; the outputs are many.** You never maintain an HTML and
  a Word version by hand.
:::


## Anatomy of a `.qmd` {.smaller}

Three kinds of content, and nothing else:

````markdown
---                                    ← 1. YAML header: how to render
title: "Blood pressure at 12 weeks"
format: html
---

## Baseline characteristics            ← 2. Prose: markdown

Mean baseline SBP was `r round(mean(trial$sbp_0), 1)` mmHg.

```{{r}}                               ← 3. Code chunk: executed
#| label: fig-sbp
#| fig-cap: "SBP by arm at 12 weeks."
boxplot(sbp_12 ~ arm, data = trial)
```
````

. . .

::: {.callout-note appearance="simple"}
The YAML header is the only part with strict syntax. **Indentation and the space
after each colon both matter.**
:::


## The YAML header {.smaller}

```yaml
---
title: "Blood pressure at 12 weeks"
author: "Shilin Zhao"
date: today
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  echo: false
  warning: false
---
```

::: {.smaller}
| Key | Does |
|---|---|
| `format:` | The output type, and its options. **`output:` in R Markdown** |
| `toc: true` | Table of contents |
| `code-fold: true` | Code hidden behind a toggle — useful for a clinical audience |
| `embed-resources: true` | One self-contained HTML file you can email |
| `execute:` | Document-wide chunk defaults |
:::


## Code chunks {.smaller}

```{{r}}
#| label: fig-sbp-arm
#| fig-cap: "Systolic BP at 12 weeks by treatment arm."
#| fig-width: 7
#| fig-height: 3.5
#| echo: true

boxplot(
  sbp_12 ~ arm,
  data = trial,
  xlab = "Treatment arm",
  ylab = "SBP at 12 weeks (mmHg)"
)
```

. . .

::: {.smaller}
- Options go **inside** the chunk, on `#|` lines, at the top. Not in the header.
- One option per line, `key: value`, YAML syntax.
- `label` is optional but makes error messages and cross-references usable.
:::

::: {.callout-note appearance="simple"}
This is the biggest visible change from R Markdown, where options lived in the
chunk header: ```` ```{r fig-sbp, echo=TRUE} ````.
:::


## The chunk options that matter {.smaller}

::: {style="font-size:0.86em"}
| Option | Effect | Use when |
|---|---|---|
| `echo: false` | Hide the code, show the output | Writing for a non-technical reader |
| `eval: false` | Show the code, don't run it | Teaching, or documenting a slow step |
| `include: false` | Run it, show nothing | Setup chunks — `library()` calls, data loading |
| `warning: false`<br>`message: false` | Suppress those streams | Tidy output — **after** you have read them once |
| `error: true` | Keep rendering despite an error | Deliberately demonstrating a failure |
| `fig-width`, `fig-height` | Figure size in inches | Always, for anything published |
| `cache: true` | Reuse results unless the chunk changes | An expensive step |
:::

. . .

::: {.callout-warning appearance="simple"}
`warning: false` is for a *finished* document. Suppressing warnings you have never
read is how you ship a bug.
:::


## Inline code {.smaller}

Write this in your prose:

```markdown
Mean baseline SBP was `r knitr::inline_expr("round(mean(trial$sbp_0), 1)")` mmHg
in `r knitr::inline_expr("nrow(trial)")` patients.
```

. . .

And it renders as:

> Mean baseline SBP was `r round(mean(trial$sbp_0), 1)` mmHg in `r nrow(trial)`
> patients.

. . .

::: {.callout-important}
This is the feature that removes an entire class of error. A number typed into
prose by hand becomes wrong the moment the data changes — and **nothing will tell
you**. A number computed inline cannot.
:::


## Figures come from code {.smaller}

```{r}
#| label: fig-sbp-live
#| fig-cap: "Systolic BP at 12 weeks by treatment arm."
#| fig-width: 8
#| fig-height: 3.4
#| echo: true
boxplot(
  sbp_12 ~ arm,
  data = trial,
  xlab = "Treatment arm",
  ylab = "SBP at 12 weeks (mmHg)",
  col = c("grey85", "grey55")
)
```

::: {.smaller}
Cross-reference it in prose as `@fig-sbp-live` and Quarto numbers it for you.
:::


## Tables come from code too {.smaller}

```{r}
#| label: tbl-baseline
#| tbl-cap: "Baseline characteristics by arm."
#| echo: true
baseline <- do.call(rbind, lapply(split(trial, trial$arm), function(d) {
  data.frame(
    Arm = unique(d$arm),
    n = nrow(d),
    `Age (mean)` = round(mean(d$age), 1),
    `Female (%)` = round(100 * mean(d$sex == "F")),
    `SBP (mean)` = round(mean(d$sbp_0), 1),
    check.names = FALSE
  )
}))
row.names(baseline) <- NULL
knitr::kable(baseline)
```

::: {.smaller}
`knitr::kable()` is enough for this course. `gt` and `flextable` do more when you
need it.
:::


## Rendering {.smaller}

::: {style="font-size:0.9em"}
| Target | Command | Needs |
|---|---|---|
| **HTML** | `quarto render report.qmd` | Nothing extra. Start here, always |
| **PDF** | `--to pdf` | A LaTeX installation — `quarto install tinytex` |
| **Word** | `--to docx` | Nothing extra; useful when a co-author needs track changes |
| **Slides** | `format: revealjs` | Nothing extra — these slides are a `.qmd` |
:::

. . .

In RStudio: the **Render** button, or {{< kbd Ctrl-Shift-K >}} /
{{< kbd Cmd-Shift-K >}}.

. . .

::: {.callout-tip appearance="simple"}
Debug in HTML. It is fast and forgiving. Switch to PDF only once the content is
right — LaTeX errors are far harder to read than anything else you will meet
today.
:::


# 4 · Why reproducibility matters {.section-divider}

## A case worth knowing {.smaller}

**Duke University, 2006–2011.** Papers claimed that gene-expression signatures
could predict which chemotherapy a patient would respond to. **Three clinical
trials** enrolled patients on the strength of the work.

. . .

When **Keith Baggerly and Kevin Coombes** tried to reproduce the analyses, they
reported problems including shifted gene indices, reversed sensitive/resistant
labels, and analyses that could not be reconstructed from the stated methods.

. . .

The trials were suspended and stopped; multiple papers were retracted. The case
contributed to a major review of how omics-based tests enter clinical use.

<small>Detailed analysis: [Baggerly & Coombes (2009)](https://doi.org/10.1214/09-AOAS291). Timeline and policy review: [Institute of Medicine (2012)](https://doi.org/10.17226/13297).</small>


## Three words that get confused {.smaller}

::: {style="font-size:0.88em"}
| Term | Means | Needs |
|---|---|---|
| **Reproducible** | Same data, same code → same result | Organized source and documented inputs |
| **Replicable** | New data, same question → same conclusion | A new study |
| **Generalizable** | The conclusion holds beyond the sample | Theory and external validity |
:::

. . .

Only the first is directly under the analyst's control, and it is the minimum
needed for someone else to evaluate the computation.


## The two ways analyses fail {.smaller}

::: columns
::: {.column width="48%"}
### It will not run

Missing files, unavailable data mounts, uninstalled packages, or undocumented
manual steps.

**Annoying, but visible.**
:::

::: {.column width="48%"}
### It runs and is wrong

Wrong file silently read, stale object left in memory, or a result edited by
hand.

**Dangerous, because it looks finished.**
:::
:::

. . .

::: {.callout-warning appearance="simple"}
A saved script and a document rendered from its source make the second class of
failure easier to detect.
:::


# 5 · Reproducible project workflow {.section-divider}

## Projects give the analysis a stable home {.smaller}

Open the supplied `.Rproj` file, or use
**File → New Project → Existing Directory**.

```
bios6301-session01/
├── bios6301-session01.Rproj
├── analysis.R
├── report.Rmd
├── report.qmd
└── data/
    └── clinic_summary.csv
```

::: {.smaller}
- Opening the project establishes a predictable project root.
- Relative paths are interpreted from that project context.
- The project folder becomes a natural Git repository in Session 2.
- Source files and input data remain distinct from generated output.
:::


## Paths are a design choice {.smaller}

::: columns
::: {.column width="48%"}
### Data inside the project

```r
clinic <- read.csv(
  "data/clinic_summary.csv"
)
```

Relative paths are convenient for course-sized, shareable data.
:::

::: {.column width="48%"}
### Large or protected data elsewhere

```r
data_root <- "D:/shared/study123"
clinic <- readRDS(file.path(
  data_root, "derived/cohort.rds"
))
```

An absolute root is often correct on a server, cluster, or shared drive.
:::
:::

. . .

`setwd()` is legitimate when the working directory is chosen deliberately. For
collaboration, define a machine-specific location once and document the required
drive, mount, or environment.


## Restarting R tests saved source {.smaller}

```r
rm(list = ls())
```

This removes objects, but loaded packages, changed options, working-directory
changes, and other process state can remain.

. . .

::: {.callout-tip}
Use **Session → Restart R** for a genuinely fresh process, then source the `.R`
script or render the complete document from the project.
:::


## The standing rule of this course {.smaller}

::: {.callout-important}
## Every submission must render from a clean session in its documented computing environment.
:::

. . .

Before you submit, every time:

::: {.incremental}
1. **Restart R** — {{< kbd Ctrl-Shift-F10 >}}
2. **Render the whole document** from scratch
3. Confirm the output is what you expect — not just that it succeeded
4. Confirm every external data location is defined once and documented
5. Confirm there is no `install.packages()` or hidden manual step in the report
:::

. . .

::: {.smaller}
Rendering already uses a fresh session, so a document that renders is *close* to
verified. Restarting first catches the remaining case: a file that quietly depends
on something you did in the console.
:::


## Summary {.smaller}

- Reproducibility is a statistical concern, not end-of-project clerical work.
- R Markdown and Quarto both keep prose, code, and computed results together.
- Use relative paths for project data; use documented absolute roots for large or
  protected data stored elsewhere.
- This course begins from `.Rmd` literacy and focuses new work on `.qmd`.
- Inline values and generated tables prevent copy-and-paste drift.
- A clean-session render tests whether the source is complete.

. . .

**Next connection:** Session 2 adds Git and GitHub so analytical changes become inspectable and recoverable.

