---
title: "R Objects: Vectors, Types, and Coercion"
subtitle: "BIOS 6301 · Introduction to Statistical Computing"
author: "Department of Biostatistics, Vanderbilt University Medical Center"
date: "Session 3"
date-format: "[Session 3]"
format:
  revealjs:
    theme: simple
    css: ../../slides.css
    slide-number: c/t
    show-slide-number: all
    incremental: false
    width: 1280
    height: 720
    fig-align: center
    code-line-numbers: false
    highlight-style: github
    footer: "BIOS 6301 · Session 3 · R Objects"
    navigation-mode: linear
execute:
  enabled: true
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false
set.seed(6301)
# Synthetic teaching data. A 60-patient two-arm trial, held as parallel
# vectors -- a data frame is a list of vectors, and that container arrives
# in session 4. Two columns are deliberately messy, exactly as they would
# arrive from a CRF export.
n_pat  <- 60
id     <- sprintf("P%03d", 1:n_pat)
arm    <- rep(c("Placebo", "Treatment"), each = 30)
age    <- round(rnorm(n_pat, 61, 9))
sex    <- sample(c("F", "M"), n_pat, replace = TRUE)
sbp_0  <- round(rnorm(n_pat, 148, 14))
sbp_12 <- round(sbp_0 -
  ifelse(arm == "Treatment", rnorm(n_pat, 14, 8), rnorm(n_pat, 3, 8)))

# Creatinine as exported: a character column, because three cells hold the
# text "NA" and two hold a below-limit-of-detection marker.
creat_raw <- as.character(round(rnorm(n_pat, 0.95, 0.25), 2))
creat_raw[c(7, 23, 41)] <- "NA"
creat_raw[c(12, 55)]    <- "<0.30"

# Weight as exported: the source system codes "not recorded" as -99.
weight_kg <- round(rnorm(n_pat, 82, 15), 1)
weight_kg[c(3, 19, 44, 58)] <- -99
```

## Where we are {.smaller}

| Course arc | Sessions | |
|---|---:|---|
| Setup and workflow | 0–2 |  |
| **Data foundations** | **3–8** | **← today** |
| Modern tools | 9–10 |  |
| R programming | 11–13 |  |
| Statistical computing | 14–16 |  |

Today's deck develops one part of the current arc; later-session tools are named only when they clarify where this idea will be used.


## What you will be able to do {.smaller}

::: {.smaller}
[1]{.objective-num} **Name** the four atomic vector types and inspect any object
with `class()`, `typeof()`, and `str()`.

[2]{.objective-num} **Distinguish** `NA`, `NaN`, `Inf`, and `NULL`, and predict
how each propagates.

[3]{.objective-num} **Predict** the type of a coerced vector, and explain how one
stray `"NA"` turns a numeric column into text.

[4]{.objective-num} **Subset** a vector by position, negation, logical condition,
and name — and say what `x[NA]` returns.

[5]{.objective-num} **Recognise** recycling, including the case where it produces
a wrong answer silently.

[6]{.objective-num} **Write** a vectorized conditional with `ifelse()`, and
explain why it is not `if`.

[7]{.objective-num} **Diagnose** a sentinel value such as `-99` before it reaches
a mean.
:::


## Roadmap for today

| | Block | Minutes |
|---|---|---|
| 1 | Everything is a vector | 15 |
| 2 | `NA`, `NaN`, `Inf` — three kinds of nothing | 10 |
| 3 | **Coercion**, and the silent wrong answer | 20 |
| 4 | Names and indexing | 20 |
| 5 | **Recycling** | 10 |
| 6 | Vectorized conditionals: `ifelse()` | 15 |

: {tbl-colwidths="[6,64,30]"}

. . .

**In lab this session** you clean two genuinely broken columns. Blocks 3 and 5 are
the ones that make that possible.


# 1 · Everything is a vector {.section-divider}

## There are no scalars {.smaller}

R has no single-number type. `42` is a **vector of length one**.

```{r}
#| label: no-scalars
length(42)
length("hello")
x <- c(4, 9, 16)
sqrt(x)
```

. . .

::: {.callout-important appearance="simple"}
`sqrt(x)` did not loop. **The function applied to the whole vector at once.**
That is what "vectorized" means, and it is why R code for data analysis contains
so few loops.
:::


## Four types you will actually use {.smaller}

::: {style="font-size:0.9em"}
| Type | Looks like | `typeof()` | Holds |
|---|---|---|---|
| **logical** | `TRUE`, `FALSE`, `NA` | `"logical"` | Conditions, flags, filters |
| **integer** | `1L`, `1:10` | `"integer"` | Counts, indices, IDs |
| **double** | `1`, `3.14`, `1e-8` | `"double"` | Everything measured |
| **character** | `"F"`, `"P001"` | `"character"` | Labels, IDs, free text |
:::

. . .

```{r}
#| label: four-types
c(typeof(TRUE), typeof(1L), typeof(1), typeof("F"))
```

. . .

::: {.smaller}
There are two more — `complex` and `raw`. You will not meet them this term.
:::


## Integer or double? {.smaller}

```{r}
#| label: int-dbl
typeof(3)
typeof(3L)
typeof(1:3)
```

. . .

```{r}
#| label: int-overflow
#| warning: true
.Machine$integer.max + 1L
```

. . .

::: {.callout-note appearance="simple"}
**Use doubles.** Write `3`, not `3L`. Integers matter in three places: memory for
very large data, `1:n` sequences, and reading someone else's code.
:::


## Three questions, three functions {.smaller}

::: {style="font-size:0.92em"}
| Question | Function | Answers with |
|---|---|---|
| What do functions think this is? | `class()` | `"numeric"`, `"character"`, `"data.frame"` |
| What is it *made of*? | `typeof()` | `"double"`, `"integer"`, `"list"` |
| What is actually in it? | `str()` | Type, length, and the first few values |
:::

. . .

```{r}
#| label: three-q
class(sbp_0); typeof(sbp_0)
str(creat_raw)
```


# 2 · Three kinds of nothing {.section-divider}

## `NA`, `NaN`, `Inf`, `NULL` {.smaller}

::: {style="font-size:0.88em"}
| | Means | Comes from | Test with |
|---|---|---|---|
| `NA` | **Not available** — a value exists, we do not know it | Missing data | `is.na()` |
| `NaN` | **Not a number** — the arithmetic is undefined | `0/0`, `log(-1)` | `is.nan()` |
| `Inf` | **Infinite** — too big, or divided by zero | `1/0`, `exp(1000)` | `is.infinite()` |
| `NULL` | **Nothing at all** — length zero, not a value | An empty result | `is.null()` |
:::

. . .

```{r}
#| label: nothings
c(1 / 0, -1 / 0, 0 / 0)
length(NA); length(NULL)
```


## `NA` is contagious {.smaller}

```{r}
#| label: na-contagious
creat <- as.numeric(creat_raw)
mean(creat)
```

. . .

```{r}
#| label: na-rm
mean(creat, na.rm = TRUE)
sum(is.na(creat))
```

. . .

::: {.callout-warning appearance="simple"}
`na.rm = TRUE` is not a fix. It is a **decision** — to analyse a different set of
patients than the one you enrolled. Report how many you dropped.
:::


## `NA == NA` is not `TRUE` {.smaller}

```{r}
#| label: na-eq
NA == NA
creat[7] == NA
```

. . .

Two unknown values. Are they equal? **Unknowable.** So R returns `NA`.

```{r}
#| label: is-na
is.na(creat[7])
```

. . .

```{r}
#| label: na-logic
NA & FALSE
NA | TRUE
```


# 3 · Coercion {.section-divider}

## One vector, one type {.smaller}

An atomic vector holds **one type**. Combine types and R does not complain — it
picks the type that can hold everything.

```{=html}
<style>
.reveal .coercion-arrow pre.mermaid { width: 100%; }
.reveal .coercion-arrow svg {
  max-width: none !important;
  width: 100% !important;
  height: auto !important;
}
</style>
```

::: {.coercion-arrow}
```{mermaid}
%%| echo: false
%%| fig-width: 10
flowchart LR
  L["<b>logical</b><br/>TRUE FALSE"] --> I["<b>integer</b><br/>1L 2L"]
  I --> D["<b>double</b><br/>1.5 3.14"]
  D --> C["<b>character</b><br/>&quot;a&quot; &quot;1.5&quot;"]
  style L fill:#f4efe4,stroke:#CFAE70
  style I fill:#f4efe4,stroke:#CFAE70
  style D fill:#CFAE70,stroke:#8a7448,color:#fff
  style C fill:#CFAE70,stroke:#8a7448,color:#fff
```
:::

. . .

::: {.smaller}
Coercion runs **left to right only**. One character element and the whole vector
is character. There is no way back that R will find for you.
:::


## `c(1, 2, "3")` {.smaller}

```{r}
#| label: coerce-c
c(1, 2, "3")
```

. . .

```{r}
#| label: coerce-c2
typeof(c(TRUE, 1L))
typeof(c(1L, 1.5))
typeof(c(1.5, "a"))
```

. . .

::: {.callout-important appearance="simple"}
**No error. No warning. No message.** The quotes in the output are the only
evidence anything happened.
:::


## Where this actually bites {.smaller}

Three cells in a creatinine column hold the text `"NA"`. Two hold `"<0.30"`.

```{r}
#| label: bites
str(creat_raw)
creat_raw[c(7, 12)]
```

. . .

```{r}
#| label: bites-mean
#| warning: true
mean(creat_raw)
```

. . .

```{r}
#| label: bites-cmp
sum(creat_raw > 1.2)
```

. . .

::: {.smaller}
Nine patients have a creatinine above 1.2. The count says **fourteen**: as text,
both `"NA"` and `"<0.30"` sort above `"1.2"`.
:::


## `as.numeric()` and what it costs {.smaller}

```{r}
#| label: as-num
#| warning: true
creat <- as.numeric(creat_raw)
sum(is.na(creat))
```

. . .

::: {.smaller}
R warned. But look at the count: **five** `NA`s, from three `"NA"` cells and two
`"<0.30"` cells. The warning does not say which.
:::

. . .

```{r}
#| label: as-num-check
creat_raw[is.na(creat)]
```

::: {.callout-tip appearance="simple"}
After every coercion: **print the values that became `NA`.** One line, and it
turns a warning into information.
:::


## The one coercion that is a gift {.smaller}

`TRUE` is 1 and `FALSE` is 0, so arithmetic on logicals is counting.

```{r}
#| label: logical-arith
sum(arm == "Treatment")
mean(sex == "F")
```

. . .

```{r}
#| label: logical-arith2
sum(is.na(creat))
round(100 * mean(sbp_0 > 140), 1)
```

. . .

::: {.smaller}
`sum()` of a logical is **how many**. `mean()` of a logical is **what
proportion**. You will use both constantly.
:::


## Sentinel values {.smaller}

The source system codes "not recorded" as `-99`.

```{r}
#| label: sentinel
mean(weight_kg)
```

. . .

```{r}
#| label: sentinel-fix
range(weight_kg)
weight_kg[weight_kg == -99] <- NA
mean(weight_kg, na.rm = TRUE)
```

. . .

::: {.callout-important appearance="simple"}
The first number is a plausible adult weight. **It is wrong, and nothing warned
you.** No coercion, no error, no `NA` — a valid mean of invalid numbers.
:::


# 4 · Names and indexing {.section-divider}

## Names {.smaller}

A vector's elements can carry names, and names are a way to index.

```{r}
#| label: names
sbp <- c(P001 = 152, P002 = 138, P003 = 161, P004 = 144)
names(sbp)
sbp["P003"]
```

. . .

```{r}
#| label: names2
sbp[c("P001", "P004")]
sbp["P999"]
```

. . .

::: {.smaller}
A name that does not exist gives `NA`, not an error. **Silent again.**
:::


## Four ways to subset {.smaller}

::: {style="font-size:0.88em"}
| By | Looks like | Returns |
|---|---|---|
| **Position** | `x[c(1, 3)]` | Elements 1 and 3 |
| **Negation** | `x[-1]` | Everything except element 1 |
| **Logical** | `x[x > 140]` | Elements where the condition is `TRUE` |
| **Name** | `x["P003"]` | The element called `P003` |
:::

. . .

::: {.smaller}
One operator, `[`, and four meanings, decided by **what type you put inside it**.
:::


## Position, and negation {.smaller}

```{r}
#| label: idx-pos
sbp_0[1:3]
sbp_0[c(1, 60)]
```

. . .

```{r}
#| label: idx-neg
length(sbp_0[-1])
length(sbp_0[-(1:10)])
```

. . .

```{r}
#| label: idx-neg-bad
#| error: true
sbp_0[c(-1, 2)]
```


## The workhorse: logical subsetting {.smaller}

```{r}
#| label: idx-logi
high <- sbp_0 > 160
head(high, 8)
length(sbp_0[high])
```

. . .

```{r}
#| label: idx-logi2
mean(sbp_12[arm == "Treatment"])
mean(sbp_12[arm == "Placebo"])
```

. . .

::: {.smaller}
`arm == "Treatment"` is a **logical vector of length 60**, used as a mask over a
different vector of length 60. This one line is the base-R equivalent of
`filter()` plus `summarise()`.
:::


## `x[NA]`, `x[FALSE]`, `x[0]` {.smaller}

```{r}
#| label: idx-weird
x <- c(a = 10, b = 20, c = 30)
x[NA]
x[FALSE]
```

. . .

```{r}
#| label: idx-weird2
x[0]
x[10]
```

. . .

::: {.callout-warning appearance="simple"}
`x[NA]` returns **three** `NA`s — the logical `NA` is *recycled* to length 3.
`x[FALSE]` returns nothing at all. These are different failures with very
different symptoms.
:::


## Subsetting on the left of `<-` {.smaller}

```{r}
#| label: subassign
w <- c(78.2, -99, 91.5, -99, 84.0)
w[w == -99] <- NA
w
```

. . .

```{r}
#| label: subassign2
sbp_flag <- rep("normal", 5)
sbp_flag[c(152, 138, 161, 144, 130) > 140] <- "high"
sbp_flag
```

. . .

::: {.smaller}
The same four ways of subsetting work on the left-hand side. This is how you
repair a column **in place**.
:::


## `which()` {.smaller}

`which()` converts a logical vector into **positions**.

```{r}
#| label: which
which(sbp_0 > 175)
id[which(sbp_0 > 175)]
```

. . .

```{r}
#| label: which2
which.max(sbp_0)
id[which.max(sbp_0)]
```

. . .

::: {.smaller}
`which()` silently **drops** `NA`s. That is convenient, and it is a difference
from logical subsetting you must know about.
:::


## `%in%`: the one that never returns `NA` {.smaller}

```{r}
#| label: in-op
c(1, NA, 3) == 1
c(1, NA, 3) %in% 1
```

. . .

```{r}
#| label: in-op2
sites <- c("VUMC", "MCV", "typo", NA)
sites %in% c("VUMC", "MCV")
```

. . .

::: {.callout-tip appearance="simple"}
Use `%in%` for set membership, `==` for a single comparison. `%in%` always
returns `TRUE` or `FALSE` — never `NA` — which makes it safe inside `[`.
:::


# 5 · Recycling {.section-divider}

## The rule {.smaller}

When two vectors have different lengths, R **repeats the shorter one** to match.

```{r}
#| label: recycle
c(10, 20, 30) + 1
sbp_0[1:4] * 2
```

. . .

::: {.smaller}
You have been relying on this since your first line of R. `x * 2` is a length-60
vector plus a length-**one** vector, recycled sixty times.
:::

. . .

```{r}
#| label: recycle2
c(1, 2, 3, 4) + c(10, 20)
```


## When it divides evenly, R says nothing {.smaller}

Two arms, sixty patients. You want a per-arm baseline correction.

```{r}
#| label: recycle-bad
correction <- c(2, 5)             # Placebo, Treatment
adjusted <- sbp_0 - correction
adjusted[1:6]
```

. . .

::: {.smaller}
Intended: subtract 2 from the 30 placebo patients, 5 from the 30 treatment
patients. Actual: **alternating** 2, 5, 2, 5 down all sixty rows.
:::

. . .

::: {.callout-important appearance="simple"}
60 is a multiple of 2, so **no warning**. Every patient got a number. Half of
them got the wrong one.
:::


## When it does not, R warns {.smaller}

```{r}
#| label: recycle-warn
#| warning: true
c(1, 2, 3) + c(10, 20)
```

. . .

::: {.smaller}
`longer object length is not a multiple of shorter object length` — a warning,
not an error. **The result is still returned and assigned.**
:::

. . .

::: {.callout-warning appearance="simple"}
This warning is one of the most valuable messages R produces. If you have
`warning: false` in your YAML and have never read your warnings once, you will
never see it.
:::


# 6 · Vectorized conditionals {.section-divider}

## `ifelse()` is a function. `if` is not. {.smaller}

::: {style="font-size:0.9em"}
| | `ifelse(test, yes, no)` | `if (cond) ... else ...` |
|---|---|---|
| Is a | **vectorized function** | **control-flow keyword** |
| Test length | Any — one answer per element | **Exactly one** `TRUE` or `FALSE` |
| Returns | A vector as long as `test` | Whatever the chosen branch returns |
| Used for | Deriving a variable from a column | Choosing what a *script* does |
| Taught in | **Today** | **Session 11** |
:::

. . .

::: {.callout-important appearance="simple"}
They look alike, share four letters, and do unrelated jobs. Today we use
`ifelse()`. `if` does not appear on any slide in this session.
:::


## `ifelse()` {.smaller}

```{r}
#| label: ifelse-1
bp_group <- ifelse(sbp_0 > 140, "high", "normal")
head(bp_group, 6)
table(bp_group)
```

. . .

```{r}
#| label: ifelse-2
adjusted <- sbp_0 - ifelse(arm == "Treatment", 5, 2)
adjusted[c(1, 2, 31, 32)]
```

. . .

::: {.smaller}
The second line is the recycling bug from block 5, **fixed** — because the
condition is length 60, so the answer is length 60.
:::


## Three ways `ifelse()` surprises you {.smaller}

```{r}
#| label: ifelse-gotcha1
ifelse(c(TRUE, FALSE, NA), "yes", "no")
```

. . .

```{r}
#| label: ifelse-gotcha2
ifelse(c(TRUE, FALSE), c(1, 2, 3), 0)
```

. . .

```{r}
#| label: ifelse-gotcha3
ifelse(c(FALSE, TRUE), 1, "x")
```

. . .

::: {.smaller}
`NA` in the test gives `NA` out · the result is as long as the **test**, not the
branches · and the branches are **coerced to one type**.
:::


## Borrowed — session 6: `if_else()` {.smaller}

```r
library(dplyr)

# Errors instead of coercing:
if_else(c(TRUE, FALSE), 1, "x")
#> Error: Can't combine `true` <double> and `false` <character>.

# Errors instead of truncating:
if_else(c(TRUE, FALSE), c(1, 2, 3), 0)
#> Error: `true` must have size 2, not size 3.

# And gives you somewhere to put the missing values:
if_else(c(TRUE, FALSE, NA), 1, 0, missing = -1)
#> [1]  1  0 -1
```

::: {.callout-note appearance="simple"}
`dplyr::if_else()` is `ifelse()` with the three surprises turned into errors.
You meet dplyr properly in **session 6** — this is here so you recognise it.
:::


## Borrowed — session 6: `case_when()` {.smaller}

```r
# Conditions are checked top to bottom; first match wins.
bp_group <- case_when(
  is.na(sbp_0)  ~ "missing",
  sbp_0 >= 160  ~ "stage 2",
  sbp_0 >= 140  ~ "stage 1",
  sbp_0 >= 130  ~ "elevated",
  .default      = "normal"
)
```

. . .

::: {.smaller}
**The base R way**

```r
bp_group <- rep("normal", length(sbp_0))
bp_group[sbp_0 >= 130] <- "elevated"
bp_group[sbp_0 >= 140] <- "stage 1"
bp_group[sbp_0 >= 160] <- "stage 2"
bp_group[is.na(sbp_0)] <- "missing"
```
:::


# 7 · Where this leaves you {.section-divider}

## Everything today was a silent wrong answer {.smaller}

Session 1: **the two ways an analysis fails.** It won't run — annoying, honest. It
runs and is wrong — dangerous.

::: {style="font-size:0.86em"}
| Today's bug | What R said | What you got |
|---|---|---|
| `c(1, 2, "3")` | Nothing | A character vector |
| `creat_raw > 1.2` | Nothing | 14 patients above 1.2. There are 9 |
| `"<0.30"` coerced | A warning you will skim | Two patients silently missing |
| `mean(weight_kg)` with `-99` | Nothing | A plausible, wrong mean |
| `sbp_0 - c(2, 5)` | Nothing | Half the cohort mis-adjusted |
| `x["P999"]` | Nothing | `NA` where you expected a value |
| `ifelse(cond, 1, "x")` | Nothing | Numbers turned into text |
:::

. . .

::: {.callout-important appearance="simple"}
Seven bugs. **One warning between them.** This is why `class()`, `str()`,
`range()`, and `length()` are habits and not commands.
:::


## Summary {.smaller}

- Atomic vectors have one type and R coerces toward a common type.
- `NA` propagates; test it with `is.na()` rather than equality.
- Logical indexing expresses inclusion rules directly.
- Recycling can return plausible results without an error.
- `ifelse()` is vectorized classification; scalar control flow arrives in Session 11.

. . .

**Next connection:** Session 4 combines vectors into matrices, lists, factors, and data frames.

