---
title: "Session 7 lab: joining, reshaping, and cleaning"
author: "Your name"
date: now
date-format: "YYYY-MM-DD HH:mm:ss Z"
format:
  html:
    embed-resources: true
execute:
  enabled: true
---

Complete the lab page's package check in the Console first. Keep the library
calls below so this report works in a fresh R session. Replace the author with
your name. All data are supplied and synthetic.

Tasks 1–8 follow the main lecture. The additional exercises are optional.
Only the last two practice exercises depend on optional lecture slides.

```{r setup}
library(dplyr)
library(tidyr)
library(stringr)
library(forcats)
library(lubridate)

# Synthetic teaching data, not real patients.
patient_lookup <- data.frame(
  id = c("P01", "P02", "P03", "P04"),
  arm = c("control", "treated", "treated", "control")
)
marker_wide <- data.frame(
  id = c("P01", "P02", "P03", "P05"),
  marker_w0 = c(4.1, 5.2, 6.0, 4.8),
  marker_w4 = c(4.4, 4.7, 5.1, 4.9)
)
raw_patients <- data.frame(
  id = c(" p01 ", "p02", " P03", "p04 "),
  arm = c(" Control ", "TREATED", " treated ", "control"),
  enrolled = c("2026-09-01", "2026-09-03", "2026-10-01", "2026-10-02"),
  note = c(" first   visit ", " repeat sample", "repeat  visit ", "first visit")
)
```

The lookup is a clean export of the participant IDs and treatment labels.
The raw export below lets you practice recreating that cleaning and adding
dates. P05 has measurements without metadata; P04 has metadata without
measurements.

## 1. Row units and keys

State what one row represents in `patient_lookup` and `marker_wide`.
Use `count(id) |> filter(n > 1)` to check both tables.
Identify the participant appearing only in each table by inspecting the data.

```{r task-1}
# Your code here.
```

Your explanation: **YOUR ANSWER**

## 2. Join and inspect unmatched participants

Save `joined_wide` with a `left_join()` of `marker_wide` and `patient_lookup`
by `id`. Compare row counts before and after. Use `anti_join()` in both
directions. Explain the missing arm value. How many rows would an `inner_join()` keep?

```{r task-2}
# Your code here.
```

Your explanation: **YOUR ANSWER**

## 3. Wide to long

Reshape `marker_wide` and save `marker_long`. Use
`cols = starts_with("marker_w")`, `names_to = "week"`,
`names_prefix = "marker_w"`, and `values_to = "marker"`.
Then use `mutate(week = as.integer(week))`.
Predict the row count and explain what one row represents now.

```{r task-3}
# Your code here.
```

Your explanation: **YOUR ANSWER**

## 4. Return to wide

Save `marker_back` using `pivot_wider()` with `id_cols = id`,
`names_from = week`, `values_from = marker`, and `names_prefix = "marker_w"`.
Compare IDs, column names, and values with `marker_wide` after sorting by ID.

```{r task-4}
# Your code here.
```

Your explanation: **YOUR ANSWER**

## 5. Clean and search text with stringr

The documented ID rule is uppercase with no outer spaces. Save `patients_text`
from `raw_patients` using `mutate()`: clean `id` with `str_trim()` and
`str_to_upper()`; clean `arm` and `note` with `str_squish()` and
`str_to_lower()`. Keep the raw table unchanged.
Show the IDs and notes where `str_detect(note, fixed("repeat"))` is true.

```{r task-5}
# Your code here.
```

Your explanation: **YOUR ANSWER**

## 6. Factor labels and order with forcats

Save `patients_factor` from `patients_text`. Use `fct_recode()` to map
`control` to `Control` and `treated` to `Treatment`. Then use
`fct_relevel()` to put `Control` first. Print the levels and counts.
Explain which step changes labels and which step changes only their order.

```{r task-6}
# Your code here.
```

Your explanation: **YOUR ANSWER**

## 7. Parse dates with lubridate

Save `patients_clean` from `patients_factor`. Convert `enrolled` with
`ymd()` into `enrolled_date`. Add `enrolled_year` and `enrolled_month`
using `year()` and `month()`. Inspect the result and check for missing parsed
dates with `filter(is.na(enrolled_date))`.
Explain why `"03/04/2026"` needs a stated month/day or day/month convention.

```{r task-7}
# Your code here.
```

Your explanation: **YOUR ANSWER**

## 8. Combine the cleaned data

Save `analysis` by left-joining `marker_long` to the `id`, `arm`, and
`enrolled_date` columns of `patients_clean`. Check the row count.
Keep week 4 and summarize mean marker by arm using Session 6's verbs.
Explain the group with missing arm instead of silently removing it.

```{r task-8}
# Your code here.
```

Your explanation: **YOUR ANSWER**

## Additional practice

Choose exercises for extra practice; these are not additional required tasks.
Practices 1–4 use the main lecture. Practices 5–6 require the named optional
slides. Only practice 5 permits an intentional error.

### Practice 1. Left versus inner

On `marker_long` and `patient_lookup`, compare left and inner joins by ID.
Explain why their row counts differ by two rather than one.

```{r practice-1}
# Your code here.
```

Your explanation: **YOUR ANSWER**

### Practice 2. Spaces inside a note

Compare `str_trim("  repeat   visit  ")` and
`str_squish("  repeat   visit  ")`. Explain which preserves the extra internal spaces.

```{r practice-2}
# Your code here.
```

Your explanation: **YOUR ANSWER**

### Practice 3. A complete factor order

For `factor(c("a", "b", "c", "a"))`, use `fct_relevel()` to set the order
to `c, b, a`. Show the levels and the character values to confirm that
reordering did not relabel observations.

```{r practice-3}
# Your code here.
```

Your explanation: **YOUR ANSWER**

### Practice 4. Three date formats

Use `ymd()`, `mdy()`, and `dmy()` to read `"2026-10-18"`,
`"10/18/2026"`, and `"18/10/2026"`. Show that they represent the same day.

```{r practice-4}
# Your code here.
```

Your explanation: **YOUR ANSWER**

### Practice 5. Optional join detail: relationship check

Only after the optional join slides: the supplied `conflicting_lookup` has two
different arm records for P01. Join `marker_long` to it without a relationship
assertion and inspect the row count. Then request `relationship = "many-to-one"`
and explain the expected error. Use the supplied error-display chunk only for
this deliberate failure.

```{r practice-5, error=TRUE}
conflicting_lookup <- data.frame(
  id = c("P01", "P01", "P02", "P03", "P04"),
  arm = c("control", "treated", "treated", "treated", "control")
)
# Your code here.
```

Your explanation: **YOUR ANSWER**

### Practice 6. Optional window example: previous visit

Only after the optional window slides: arrange `marker_long` by ID/week,
group by ID, and compute `change = marker - lag(marker)`. Ungroup afterward.
Explain the missing first change for each participant.

```{r practice-6}
# Your code here.
```

Your explanation: **YOUR ANSWER**
