---
title: "Resampling Methods"
subtitle: "BIOS 6301 · Introduction to Statistical Computing"
author: "Department of Biostatistics, Vanderbilt University Medical Center"
date: "Session 15"
date-format: "[Session 15]"
format:
  revealjs:
    theme: simple
    css: ../../slides.css
    slide-number: c/t
    show-slide-number: all
    incremental: false
    width: 1280
    height: 720
    fig-align: center
    code-line-numbers: false
    highlight-style: github
    footer: "BIOS 6301 · Session 15 · Resampling Methods"
    navigation-mode: linear
execute:
  enabled: true
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(knitr)

# Synthetic teaching data for BIOS 6301. Not real patients.
set.seed(6301)
n <- 60
trial <- data.frame(
  id        = sprintf("P%03d", 1:n),
  arm       = rep(c("Placebo", "Treatment"), each = n / 2),
  site      = sample(1:4, n, replace = TRUE),
  age       = round(rnorm(n, 61, 9)),
  sex       = sample(c("F", "M"), n, replace = TRUE),
  weight_kg = round(rnorm(n, 82, 15), 1),
  sbp_0     = round(rnorm(n, 148, 14)),
  stringsAsFactors = FALSE
)
trial$sbp_12 <- round(trial$sbp_0 -
  ifelse(trial$arm == "Treatment", rnorm(n, 14, 8), rnorm(n, 3, 8)))

# Repeated-measures version of the same cohort, for the resampling-unit demo:
# three visits per patient, a large patient-level random intercept (SD 12) and
# small within-patient noise (SD 4). Nothing is "broken" here — the data are
# simply clustered, which is what makes the row bootstrap wrong.
set.seed(2024)
patient_effect <- rnorm(n, 0, 12)
trial_long <- data.frame(
  id   = rep(trial$id, each = 3),
  arm  = rep(trial$arm, each = 3),
  week = rep(c(4, 8, 12), times = n),
  stringsAsFactors = FALSE
)
trial_long$sbp <- round(140 + rep(patient_effect, each = 3) +
  ifelse(trial_long$arm == "Treatment", -10, 0) + rnorm(n * 3, 0, 4))

theme_set(theme_minimal(base_size = 15))
gold <- "#CFAE70"
```

## Where we are {.smaller}

| Course arc | Sessions | |
|---|---:|---|
| Setup and workflow | 0–2 |  |
| Data foundations | 3–8 |  |
| Modern tools | 9–10 |  |
| R programming | 11–13 |  |
| **Statistical computing** | **14–16** | **← today** |

Today's deck develops one part of the current arc; later-session tools are named only when they clarify where this idea will be used.


## What you will be able to do {.smaller}

::: {.smaller}
[1]{.objective-num} **Explain** what sampling *with* and *without* replacement
each estimate, and why one line of code separates them.

[2]{.objective-num} **Write** a bootstrap by hand — resample, recompute, collect
— and report a bootstrap standard error.

[3]{.objective-num} **Construct** percentile and basic bootstrap intervals, and
say what each assumes.

[4]{.objective-num} **Name** four situations where the bootstrap fails, and
recognise one from its output.

[5]{.objective-num} **Build** a permutation null distribution, and compute a
*p*-value that is never exactly zero.

[6]{.objective-num} **Choose** the correct resampling unit for clustered or
longitudinal data — and show what goes wrong when you do not.
:::


## Roadmap for today {.smaller}

| | Block | Minutes |
|---|---|---|
| 1 | One argument apart | 10 |
| 2 | The bootstrap, by hand | 25 |
| 3 | When the bootstrap fails | 10 |
| 4 | Permutation and exchangeability | 20 |
| 5 | **Choosing the resampling unit** | 20 |
| 6 | Cross-validation, and closing the course | 15 |

: {tbl-colwidths="[6,64,30]"}

. . .

**Block 5 is the one to protect.** If the room is running slow, compress block 3
and the `boot` pointers — not block 5.


# 1 · One argument apart {.section-divider}

## The whole session, in one table {.smaller}

| | Resample | Target |
|---|---|---|
| **Bootstrap** | *with* replacement, from the observed data | the **sampling distribution** of a statistic |
| **Permutation** | *without* replacement — a reshuffle of labels | the **null distribution** under exchangeability |

: {tbl-colwidths="[18,44,38]"}

. . .

Which decides, for the rest of your career, which one you reach for:

::: {.smaller}
| You want | Use | You get |
|---|---|---|
| A standard error | Bootstrap | `sd()` of the resampled statistics |
| A confidence interval | Bootstrap | quantiles, or a reflection of them |
| A *p*-value | Permutation | how often chance beats what you saw |
:::

. . .

::: {.callout-important appearance="simple"}
The bootstrap answers **"how variable is my estimate?"**
Permutation answers **"could chance alone have produced this?"**
:::


## The same line of code, twice {.smaller}

```{r}
#| label: two-samples
sbp <- c(150, 142, 168, 131, 155)

set.seed(1)
sample(sbp, replace = TRUE)   # bootstrap: some values twice, some not at all
sample(sbp)                   # permutation: same values, new order
```

. . .

::: {.smaller}
- **With replacement** the multiset changes — that is what generates variability
  between resamples.
- **Without replacement** the multiset is identical every time. Only the *pairing*
  with something else — a label, an arm, a time — changes.
:::

. . .

::: {.callout-note appearance="simple"}
`sample(x)` with no other arguments is a permutation of `x`. You have been
writing permutations since Session 14 without calling them that.
:::


# 2 · The bootstrap, by hand {.section-divider}

## Why it works at all {.smaller}

The thing you want is a property of the **population**: the standard deviation of
your estimator across repeated samples from it.

. . .

You do not have the population. You have one sample — which is your best available
estimate of it.

. . .

::: {.callout-important appearance="simple"}
**The plug-in principle.** Wherever the formula says *population*, substitute the
*empirical distribution of your sample*, and compute.
:::

. . .

::: {.smaller}
So: resample from your data exactly as nature sampled from the population — same
size *n*, drawn with replacement — and watch how the statistic moves.
:::


## Resample, recompute, collect {.smaller}

```{mermaid}
%%| echo: false
%%| fig-width: 10
flowchart LR
  D["<b>your data</b><br/>n = 60"] --> R["<b>resample</b><br/>n = 60,<br/>with replacement"]
  R --> S["<b>recompute</b><br/>the statistic"]
  S --> C["<b>collect</b><br/>one number"]
  C -.->|"B = 2000 times"| R
  C --> H["<b>bootstrap<br/>distribution</b>"]
  style D fill:#f4efe4,stroke:#CFAE70
  style R fill:#CFAE70,stroke:#8a7448,color:#fff
  style S fill:#CFAE70,stroke:#8a7448,color:#fff
```

. . .

::: {.smaller}
Three lines of code, and none of them are new. `sample()` is Session 14, indexing
a data frame by a vector of row numbers is Session 4, and `replicate()` is a
functional from Session 13.
:::


## The statistic, as a function {.smaller}

We want the **difference in median** 12-week SBP between arms. There is no
convenient formula for its standard error — which is exactly when the bootstrap
earns its keep.

```{r}
#| label: med-diff
# Difference in median sbp_12, Treatment minus Placebo.
# Takes a data frame with `arm` and `sbp_12`; returns one number.
med_diff <- function(data) {
  stopifnot(is.data.frame(data), all(c("arm", "sbp_12") %in% names(data)))
  median(data$sbp_12[data$arm == "Treatment"]) -
    median(data$sbp_12[data$arm == "Placebo"])
}

obs_med <- med_diff(trial)
obs_med
```

. . .

::: {.smaller}
Statistic first, resampling second. If the function is wrong, 2000 resamples make
it wrong 2000 times.
:::


## The bootstrap, in three lines {.smaller}

```{r}
#| label: boot-med
set.seed(14)
B <- 2000

boot_med <- replicate(B, {
  idx <- sample(nrow(trial), replace = TRUE)   # 60 row numbers, with repeats
  med_diff(trial[idx, ])
})

str(boot_med)
head(round(boot_med, 1))
```

. . .

::: {.smaller}
::: {.callout-note appearance="simple" title="The tidyverse way — same thing"}
```r
purrr::map_dbl(1:B, \(i) med_diff(dplyr::slice_sample(trial, prop = 1,
                                                      replace = TRUE)))
```
:::
:::


## What came back {.smaller}

```{r}
#| label: fig-boot-med
#| fig-cap: "Bootstrap distribution of the difference in median SBP at 12 weeks."
#| fig-width: 9
#| fig-height: 3.2
#| echo: false
ggplot(data.frame(stat = boot_med), aes(stat)) +
  geom_histogram(binwidth = 1, fill = gold, colour = "white") +
  geom_vline(xintercept = obs_med, linewidth = 1) +
  annotate("text", x = obs_med, y = Inf, label = "  observed", hjust = 0,
           vjust = 1.6, size = 4.5) +
  labs(x = "Difference in median SBP (mmHg)", y = "Resamples")
```

::: {.smaller}
Centred near the observed value. Ragged, because a median of 30 integers can only
land on so many values.
:::


## The bootstrap standard error {.smaller}

The standard deviation of the resampled statistics **is** the estimated standard
error. That is the whole definition.

```{r}
#| label: boot-se
se_boot <- sd(boot_med)
se_boot
```

. . .

```{r}
#| label: boot-bias
mean(boot_med) - obs_med    # bootstrap estimate of bias
```

. . .

::: {.smaller}
A small bias estimate is reassuring. A large one is a signal to stop and think —
not a licence to subtract it, which usually inflates variance more than it
removes bias.
:::


## Percentile and basic intervals {.smaller}

```{r}
#| label: boot-ci
perc <- quantile(boot_med, c(0.025, 0.975))              # percentile
basic <- 2 * obs_med - quantile(boot_med, c(0.975, 0.025))  # basic
rbind(percentile = perc, basic = unname(basic))
```

. . .

::: {.smaller}
| Interval | Rule | Assumes |
|---|---|---|
| **Percentile** | the 2.5th and 97.5th percentiles of the resamples | the bootstrap distribution is roughly the sampling distribution |
| **Basic** | reflect those percentiles through the observed value | the *shape* is right, the *centre* may be off |
:::

. . .

::: {.callout-warning appearance="simple"}
Here they disagree — the basic interval reaches across zero and the percentile
interval does not. That disagreement is **information**, not a nuisance.
:::


## Check it where you know the answer {.smaller}

The bootstrap is most trustworthy when you have first watched it reproduce a
result you can verify.

```{r}
#| label: boot-vs-t
mean_diff <- function(data) {
  mean(data$sbp_12[data$arm == "Treatment"]) -
    mean(data$sbp_12[data$arm == "Placebo"])
}

set.seed(17)
boot_mean <- replicate(B, mean_diff(trial[sample(nrow(trial), replace = TRUE), ]))

c(boot_se = sd(boot_mean),
  welch_se = t.test(sbp_12 ~ arm, data = trial)$stderr)
```

. . .

::: {.smaller}
Two decimal places apart. Now go back and believe the median result, which has no
formula to check against.
:::


## How many resamples? {.smaller}

More resamples do **not** improve the estimate. They reduce the *Monte Carlo
noise* in your report of it.

::: {.smaller}
| Purpose | *B* | Why |
|---|---|---|
| Standard error | 1,000–2,000 | The SD stabilises quickly |
| Confidence interval | ≥ 10,000 | Tail quantiles are estimated from the tail |
| A number going in a paper | 10,000+ | It must not change on re-render |
:::

. . .

```{r}
#| label: mc-check
boot_se_once <- function(b) {
  sd(replicate(b, med_diff(trial[sample(nrow(trial), replace = TRUE), ])))
}

set.seed(101)
round(replicate(6, boot_se_once(500)), 3)   # six runs, same data, B = 500
```


# 3 · When the bootstrap fails {.section-divider}

## Four ways it breaks {.smaller}

::: {style="font-size:0.9em"}
| Situation | What goes wrong | Symptom |
|---|---|---|
| **Small *n*** | The empirical distribution is a poor stand-in; only *n^n^* distinct resamples exist | Chunky, lumpy bootstrap distribution |
| **Extremes** — max, min, range | The extreme of a resample is drawn from a handful of values | A distribution with visible spikes |
| **Heavy tails** — infinite variance | There is no sampling distribution to estimate | Intervals that grow with *B*; wild resamples |
| **Dependence** — clustered, longitudinal, time series | Resampling rows destroys the dependence you needed to keep | Standard errors that are **too small**, confidently |
:::

. . .

::: {.callout-warning appearance="simple"}
Only the last one is silent. The other three announce themselves in a histogram
you should have plotted anyway.
:::


## The maximum, bootstrapped {.smaller}

```{r}
#| label: boot-max
set.seed(18)
boot_max <- replicate(B, max(trial$sbp_0[sample(nrow(trial), replace = TRUE)]))

length(unique(boot_max))                # distinct values across 2000 resamples
mean(boot_max == max(trial$sbp_0))      # how often the resample max is THE max
```

. . .

::: {.smaller}
The resampled maximum can never exceed the observed maximum, and it equals it
whenever the largest patient is drawn at all — probability
1 − (1 − 1/*n*)^*n*^ ≈ 0.63.
:::

. . .

::: {.callout-important appearance="simple"}
The bootstrap distribution of a maximum is a spike at the observed value with a
short left tail. It is not a sampling distribution. **No amount of *B* fixes it.**
:::


## Two packages you should know exist {.smaller}

Not installed here — this is a pointer, not a demonstration.

```r
# boot: the classical implementation. Statistic takes (data, indices).
library(boot)
b <- boot(trial, statistic = function(data, i) med_diff(data[i, ]), R = 10000)
boot.ci(b, type = c("perc", "basic", "bca"))

# rsample: tidymodels. Resamples as a tibble of splits.
library(rsample)
bootstraps(trial, times = 1000)
```

. . .

::: {.smaller}
`boot.ci()` gives you **BCa** intervals — bias-corrected and accelerated — which
adjust for both the bias and the skewness we saw two slides ago. That is the
single best reason to use the package rather than your own loop.
:::


# 4 · Permutation and exchangeability {.section-divider}

## The logic {.smaller}

**Null hypothesis:** the treatment label carries no information about the outcome.

. . .

If that is true, then each patient's outcome would have been the same whichever
label they received. The labels are **exchangeable** — any reassignment of the 30
Treatment labels among the 60 patients was equally likely.

. . .

::: {.smaller}
So: reassign them, thousands of times, and each time compute the statistic. That
set of numbers is the **null distribution** — not assumed, not looked up in a
table, but constructed from your own data.
:::

. . .

::: {.callout-important appearance="simple"}
In a randomised trial this is not a modelling assumption. It is a description of
what the randomisation actually did.
:::


## Shuffle the labels {.smaller}

```{r}
#| label: perm-run
obs_mean <- mean_diff(trial)

set.seed(19)
P <- 999
perm_stat <- replicate(P, {
  shuffled <- sample(trial$arm)                 # permutation, not a bootstrap
  mean(trial$sbp_12[shuffled == "Treatment"]) -
    mean(trial$sbp_12[shuffled == "Placebo"])
})

c(observed = obs_mean, perm_mean = mean(perm_stat), perm_sd = sd(perm_stat))
```

. . .

::: {.smaller}
Note what is **not** shuffled: `sbp_12` never moves. Only its label does. Shuffling
both, or shuffling the outcome instead, gives the same answer here — and stops
being equivalent the moment you have a covariate.
:::


## The null distribution {.smaller}

```{r}
#| label: fig-perm
#| fig-cap: "Permutation null distribution of the mean difference, with the observed value."
#| fig-width: 9
#| fig-height: 3.2
#| echo: false
ggplot(data.frame(stat = perm_stat), aes(stat)) +
  geom_histogram(bins = 40, fill = "grey78", colour = "white") +
  geom_vline(xintercept = obs_mean, colour = "#8a4b1f", linewidth = 1.2) +
  annotate("text", x = obs_mean, y = Inf, label = "observed  ", hjust = 1,
           vjust = 1.6, size = 4.5, colour = "#8a4b1f") +
  labs(x = "Mean difference in SBP at 12 weeks (mmHg)", y = "Permutations")
```

::: {.smaller}
Centred at zero **by construction**. The *p*-value is the proportion of this
histogram at least as extreme as the line.
:::


## Exact, or Monte Carlo? {.smaller}

With 4 patients per arm there are only `choose(8, 4)` = **70** possible label
assignments. You can enumerate **all** of them:

```{r}
#| label: exact-perm
small <- trial[c(1:4, 31:34), c("id", "arm", "sbp_12")]
obs_small <- mean_diff(small)

combos <- combn(8, 4)                    # every way to choose 4 "Treatment"
exact <- apply(combos, 2, function(k) mean(small$sbp_12[k]) -
                                      mean(small$sbp_12[-k]))

ncol(combos)                             # the entire reference set
round(mean(abs(exact) >= abs(obs_small)), 4)    # exact two-sided p-value
```

. . .

::: {.smaller}
With 30 per arm there are `choose(60, 30)` ≈ 1.2 × 10^17^. So you
sample from them instead — that is all "Monte Carlo permutation" means.
:::


## A permutation *p*-value is never zero {.smaller}

```{r}
#| label: perm-p
n_extreme <- sum(abs(perm_stat) >= abs(obs_mean))
n_extreme                                        # out of 999
```

. . .

```{r}
#| label: perm-p2
c(wrong = n_extreme / P,
  right = (n_extreme + 1) / (P + 1))
```

. . .

::: {.smaller}
The observed arrangement **is** one of the possible arrangements. Counting it in
both numerator and denominator is not a fudge to avoid zero — it is the correct
count.
:::

. . .

::: {.callout-warning appearance="simple"}
Report `p < 0.001`, or increase *P*. Never `p = 0`.
:::


## What permutation does not give you {.smaller}

::: columns
::: {.column width="49%"}
### It gives

- A valid *p*-value with no distributional assumption
- Exact validity when the randomisation was real
- Freedom to use any statistic you like
:::

::: {.column width="49%"}
### It does not give

- A confidence interval
- An effect size
- Protection against confounding
- A test of "equal means" specifically
:::
:::

. . .

::: {.smaller}
The last one is subtle. The null being tested is that the two label groups have
the **same distribution**. A significant result may be a difference in spread or
shape, not in centre.
:::

. . .

::: {.callout-note appearance="simple"}
Which is why the two halves of today go together: permutation for the *p*-value,
bootstrap for the interval.
:::


# 5 · Choosing the resampling unit {.section-divider}

## The same cohort, three visits each {.smaller}

```{r}
#| label: long-head
head(trial_long, 7)
c(rows = nrow(trial_long), patients = length(unique(trial_long$id)))
```

. . .

```{r}
#| label: long-obs
diff_long <- function(data) {
  mean(data$sbp[data$arm == "Treatment"]) - mean(data$sbp[data$arm == "Placebo"])
}
obs_long <- diff_long(trial_long)
obs_long
```

::: {.smaller}
180 rows. **60 patients.** Hold on to that distinction — the rest of this block is
about which number is the sample size.
:::


## The naive row bootstrap {.smaller}

```{r}
#| label: boot-rows
set.seed(21)
boot_row <- replicate(B, {
  idx <- sample(nrow(trial_long), replace = TRUE)   # 180 rows, independently
  diff_long(trial_long[idx, ])
})

c(estimate = obs_long, se = sd(boot_row))
round(quantile(boot_row, c(0.025, 0.975)), 2)
```

. . .

::: {.smaller}
Perfectly reasonable code. Runs without complaint. Produces a clean, symmetric
histogram. The interval comfortably excludes zero.
:::

. . .

::: {.callout-warning appearance="simple"}
It is wrong, and nothing on this slide tells you so.
:::


## The cluster bootstrap {.smaller}

Resample **patients**, and take all of each patient's rows with them.

```{r}
#| label: boot-cluster
ids <- unique(trial_long$id)
rows_by_id <- split(seq_len(nrow(trial_long)), trial_long$id)   # id -> row numbers

set.seed(22)
boot_clu <- replicate(B, {
  picked <- sample(ids, replace = TRUE)              # 60 patients, with repeats
  diff_long(trial_long[unlist(rows_by_id[picked]), ])
})

c(row_se = sd(boot_row), cluster_se = sd(boot_clu),
  ratio = sd(boot_clu) / sd(boot_row))
```

. . .

::: {.smaller}
The honest standard error is **1.7 times** the naive one. Every interval built on
the naive number is 40% too narrow.
:::


## Side by side {.smaller}

```{r}
#| label: fig-unit
#| fig-cap: "Bootstrap distributions of the same estimate under two resampling schemes."
#| fig-width: 9
#| fig-height: 3.3
#| echo: false
cmp <- rbind(
  data.frame(stat = boot_row, scheme = "Naive: resample rows (wrong)"),
  data.frame(stat = boot_clu, scheme = "Cluster: resample patients (right)")
)
cmp$scheme <- factor(cmp$scheme, levels = unique(cmp$scheme))
ggplot(cmp, aes(stat, fill = scheme)) +
  geom_histogram(bins = 45, colour = "white", alpha = 0.85) +
  geom_vline(xintercept = 0, linetype = 2) +
  facet_wrap(~scheme) +
  scale_fill_manual(values = c("grey72", gold)) +
  labs(x = "Difference in mean SBP (mmHg)", y = "Resamples") +
  theme(legend.position = "none")
```

::: {.smaller}
Same data, same estimate, same statistic function. Only the **unit** changed.
:::


## Why: how much is a visit worth? {.smaller}

```{r}
#| label: icc
between <- var(tapply(trial_long$sbp, trial_long$id, mean))
within  <- mean(tapply(trial_long$sbp, trial_long$id, var))
icc <- between / (between + within)
c(icc = icc, design_effect = 1 + (3 - 1) * icc, inflation = sqrt(1 + 2 * icc))
```

. . .

::: {.smaller}
With an intra-class correlation near 0.9, three visits carry barely more
information than one. The **design effect** 1 + (*m* − 1)ρ says how many
independent observations your *n* × *m* rows are actually worth — and its square
root is exactly the factor the cluster bootstrap recovered.
:::

. . .

::: {.callout-note appearance="simple"}
The bootstrap found that number without being told the formula. That is the point
of the method.
:::


## Permuting at the wrong level, too {.smaller}

The same mistake, in the other half of the session:

```{r}
#| label: perm-unit
arm_by_id <- trial_long$arm[match(ids, trial_long$id)]

set.seed(32)
p_rows <- replicate(999, {
  s <- sample(trial_long$arm)                    # shuffles visits, not patients
  mean(trial_long$sbp[s == "Treatment"]) - mean(trial_long$sbp[s == "Placebo"])
})
p_pats <- replicate(999, {
  s <- rep(sample(arm_by_id), each = 3)          # shuffles patients (rows sorted by id)
  mean(trial_long$sbp[s == "Treatment"]) - mean(trial_long$sbp[s == "Placebo"])
})

c(p_row_unit = (sum(abs(p_rows) >= abs(obs_long)) + 1) / 1000,
  p_patient_unit = (sum(abs(p_pats) >= abs(obs_long)) + 1) / 1000)
```

::: {.smaller}
An order of magnitude apart. **Randomisation happened to patients**, so the
permutation must too.
:::


## The rule {.smaller}

::: {.callout-important}
## Resample the unit that was independently sampled — the unit that was randomised.
:::

. . .

::: {style="font-size:0.92em"}
| Design | Rows | Resample |
|---|---|---|
| Simple random sample | Patients | Patients |
| Repeated measures | Patient-visits | **Patients** |
| Cluster-randomised trial | Patients | **Clinics** |
| Multi-site trial, site-stratified | Patients | Patients, **within site** |
| Matched pairs / crossover | Observations | **Pairs** |
| Time series | Time points | **Blocks** of consecutive points |
:::

. . .

::: {.smaller}
Before you write `sample()`, answer one question in prose: *what did this study
draw at random?*
:::


# 6 · Cross-validation, and closing the course {.section-divider}

## The same idea, aimed at prediction {.smaller}

Fitting and evaluating on the same data is the statistical equivalent of marking
your own homework.

```{r}
#| label: cv
set.seed(55)
noise <- as.data.frame(matrix(rnorm(n * 12), nrow = n))   # 12 pure-noise predictors
names(noise) <- sprintf("z%02d", 1:12)
d <- cbind(trial[, c("sbp_12", "sbp_0", "age", "weight_kg")], noise)

fit <- lm(sbp_12 ~ ., data = d)
rmse_in <- sqrt(mean(residuals(fit)^2))

set.seed(56)
fold <- sample(rep(1:5, length.out = n))
pred <- numeric(n)
for (k in 1:5) {
  m <- lm(sbp_12 ~ ., data = d[fold != k, ])
  pred[fold == k] <- predict(m, newdata = d[fold == k, ])
}

round(c(in_sample_rmse = rmse_in, cv_rmse = sqrt(mean((d$sbp_12 - pred)^2))), 2)
```


## Summary {.smaller}

- Bootstrap, permutation, and cross-validation answer different questions.
- The resampling unit must preserve the design's dependence structure.
- Permutation validity rests on exchangeability under the null.
- Preprocessing learned outside a training fold leaks information.
- Finite resampling adds Monte Carlo variability that should be recorded.

. . .

**Next connection:** Session 16 reserves time for strings, regular expressions, and practical needs observed during the course.

