---
title: "Data Structures and Base-R Data Handling"
subtitle: "BIOS 6301 · Introduction to Statistical Computing"
author: "Department of Biostatistics, Vanderbilt University Medical Center"
date: "Session 4"
date-format: "[Session 4]"
format:
  revealjs:
    theme: simple
    css: ../../slides.css
    slide-number: c/t
    show-slide-number: all
    incremental: false
    width: 1280
    height: 720
    fig-align: center
    code-line-numbers: false
    highlight-style: github
    footer: "BIOS 6301 · Session 4 · Data structures and base-R data handling"
    navigation-mode: linear
execute:
  enabled: true
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false
library(knitr)
set.seed(6301)

# Synthetic teaching data. A two-arm hypertension trial, 60 patients, invented
# for this deck — no real patients, no real site names in the clinical sense.
# Deliberately imperfect: `dose_mg` arrives from the CRF as *text*, `grade` has
# a level nobody was assigned, and three 12-week measurements are missing.
n <- 60
trial <- data.frame(
  id    = sprintf("P%03d", 1:n),
  arm   = rep(c("Placebo", "Treatment"), each = n / 2),
  site  = sample(c("VUMC", "Meharry", "Regional"), n, replace = TRUE),
  sex   = sample(c("F", "M"), n, replace = TRUE),
  age   = round(rnorm(n, 61, 9)),
  sbp_0 = round(rnorm(n, 148, 14)),
  stringsAsFactors = FALSE
)
trial$sbp_6 <- round(trial$sbp_0 -
  ifelse(trial$arm == "Treatment", rnorm(n, 9, 7), rnorm(n, 2, 7)))
trial$sbp_12 <- round(trial$sbp_0 -
  ifelse(trial$arm == "Treatment", rnorm(n, 14, 8), rnorm(n, 3, 8)))
trial$sbp_12[c(7, 23, 41)] <- NA
trial$dose_mg <- ifelse(trial$arm == "Treatment",
  sample(c("5", "10", "20", "40"), n, replace = TRUE), "0")
trial$grade <- factor(
  sample(c("mild", "moderate", "severe"), n, replace = TRUE,
         prob = c(0.5, 0.35, 0.15)),
  levels = c("mild", "moderate", "severe", "life-threatening")
)

# Repeated SBP measurements as a matrix, for the matrix block.
sbp <- as.matrix(trial[, c("sbp_0", "sbp_6", "sbp_12")])
rownames(sbp) <- trial$id
```

## Where we are {.smaller}

| Course arc | Sessions | |
|---|---:|---|
| Setup and workflow | 0–2 |  |
| **Data foundations** | **3–8** | **← today** |
| Modern tools | 9–10 |  |
| R programming | 11–13 |  |
| Statistical computing | 14–16 |  |

Today's deck develops one part of the current arc; later-session tools are named only when they clarify where this idea will be used.


## What you will be able to do {.smaller}

::: {.smaller}
[1]{.objective-num} **Distinguish** the four containers — vector, matrix, list,
data frame — and say which one a given object is.

[2]{.objective-num} **Index** a data frame by row, by column, and by both, and
predict the class of what comes back.

[3]{.objective-num} **Choose** correctly between `[`, `[[`, and `$`.

[4]{.objective-num} **Explain** what a factor actually is, and never again write
`as.numeric(as.factor(x))`.

[5]{.objective-num} **Use** `subset()`, `transform()`, and `order()` to filter,
derive, and sort.

[6]{.objective-num} **Compute** grouped summaries with `tapply()` and
`aggregate()`.

[7]{.objective-num} **Inspect** any new data frame with `str()` before trusting
a single number from it.
:::


## Roadmap for today

| | Block | Minutes |
|---|---|---|
| 1 | Four containers: matrix, list, data frame | 20 |
| 2 | Factors, and the catastrophe | 15 |
| 3 | Indexing a data frame: `[`, `[[`, `$` | 20 |
| 4 | Three base verbs: `subset`, `transform`, `order` | 15 |
| 5 | Split–apply–combine: `tapply`, `aggregate` | 15 |
| 6 | Inspection discipline | 5 |

: {tbl-colwidths="[6,64,30]"}

. . .

Every example uses one 60-patient synthetic trial. **It is deliberately
imperfect** — that is where the teaching is.


# 1 · Four containers {.section-divider}

## Two questions, four answers {.smaller}

Every R data structure answers two questions: **how many dimensions**, and
**must everything be the same type?**

::: {style="font-size:0.92em"}
| | One type only | Mixed types allowed |
|---|---|---|
| **1 dimension** | `vector` | `list` |
| **2 dimensions** | `matrix` | `data.frame` |
:::

. . .

::: {.smaller}
- A **matrix** is a vector with a `dim` attribute. Nothing more.
- A **data frame** is a list of equal-length columns. Nothing more.
:::

. . .

::: {.callout-important appearance="simple"}
Those two sentences are the whole session. Everything else follows from them.
:::


## A matrix is a vector with a shape {.smaller}

```{r}
#| label: matrix-build
sbp <- as.matrix(trial[, c("sbp_0", "sbp_6", "sbp_12")])
rownames(sbp) <- trial$id
dim(sbp)
sbp[1:3, ]
```

. . .

```{r}
#| label: matrix-summaries
round(colMeans(sbp, na.rm = TRUE), 1)
```


## Matrices are strict about type {.smaller}

Ask for a character column and a numeric column in the same matrix:

```{r}
#| label: matrix-coerce
as.matrix(trial[, c("id", "age")])[1:3, ]
```

. . .

The ages are now `"57"`, `"63"`, `"55"` — **text**. Coercion, exactly as in
session 3, silently.

. . .

::: {.callout-warning appearance="simple"}
This is why clinical data lives in data frames. One identifier column would turn
your entire dataset into strings.
:::


## Lists hold anything {.smaller}

```{r}
#| label: list-build
result <- list(
  study = "SBP-12",
  n     = nrow(trial),
  arms  = c("Placebo", "Treatment"),
  sbp_0 = trial$sbp_0
)
str(result)
```

. . .

::: {.smaller}
Different types, different lengths, all in one object. Every model object in R —
everything `lm()` or `t.test()` gives you back — **is a list**.
:::


## `[` and `[[` on a list {.smaller}

```{r}
#| label: list-bracket
class(result["n"])
class(result[["n"]])
```

. . .

::: {.smaller}
| Notation | Gives you | Think |
|---|---|---|
| `result["n"]` | a **list** of length 1 | a smaller version of the same thing |
| `result[["n"]]` | the **value** inside | reach in and take it out |
| `result$n` | the value inside | `[[` with less typing |
:::

. . .

::: {.callout-note appearance="simple"}
The pigeonholes: `[` gives you a pigeonhole, `[[` gives you what is *in* the
pigeonhole. `[` on a list always returns a list.
:::


## A data frame is a list of columns {.smaller}

```{r}
#| label: df-is-list
length(trial)
nrow(trial)
class(trial[[2]])
```

. . .

::: {.smaller}
`length()` on a data frame counts **columns**, because a data frame *is* a list
and its elements are its columns. It just happens to enforce that all of them
have the same length.
:::

. . .

::: {.callout-important appearance="simple"}
Consequence: everything you just learned about `[` versus `[[` on lists applies
unchanged to data frame columns.
:::


# 2 · Factors {.section-divider}

## A factor is integers wearing labels {.smaller}

```{r}
#| label: factor-guts
head(trial$grade, 4)
levels(trial$grade)
head(as.integer(trial$grade), 4)
```

. . .

::: {.smaller}
Stored as `3 3 1 3`. Displayed as `severe severe mild severe`. The labels live in
the `levels` attribute, and R looks them up when it prints.
:::


## The levels are the point {.smaller}

```{r}
#| label: factor-levels
table(trial$grade)
```

. . .

Nobody was `life-threatening`. The level is still there — **and that is
correct**. The study protocol defined four grades; your sample happened to
contain three.

. . .

```{r}
#| label: factor-drop
table(droplevels(trial$grade))
```


## Level order is metadata {.smaller}

```{r}
#| label: factor-relevel
arm <- factor(trial$arm, levels = c("Placebo", "Treatment"))
arm_treatment_ref <- relevel(arm, ref = "Treatment")

levels(arm)
levels(arm_treatment_ref)
show_rows <- c(1, 31)  # one patient from each arm
data.frame(
  before = as.character(arm[show_rows]),
  after = as.character(arm_treatment_ref[show_rows]),
  before_code = as.integer(arm[show_rows]),
  after_code = as.integer(arm_treatment_ref[show_rows])
)
```

. . .

::: {.callout-tip appearance="simple"}
Changing level order changes the factor's metadata and integer codes—**not the
observed labels**. With R's default treatment contrasts, the first level is the
model reference.
:::


## Safely set the complete level order {.smaller}

```{r}
#| label: factor-complete-order
new_order <- c("life-threatening", "severe", "moderate", "mild")
grade_reordered <- factor(
  as.character(trial$grade),
  levels = new_order
)

levels(trial$grade)
levels(grade_reordered)
identical(as.character(trial$grade), as.character(grade_reordered))
```

. . .

::: {.callout-tip appearance="simple"}
`factor(as.character(x), levels = new_order)` rebuilds the factor from its labels.
The observations stay in the same categories; only the complete level order and
internal codes change.
:::


## `levels(x) <- ...` relabels the data {.smaller}

```{r}
#| label: factor-wrong-reorder
arm1 <- arm
levels(arm1) <- c("Treatment", "Placebo")

data.frame(
  id = trial$id[show_rows],
  before = as.character(arm[show_rows]),
  after = as.character(arm1[show_rows])
)
```

. . .

::: {.callout-warning appearance="simple"}
This did **not** reorder the level dictionary while preserving the data. It
renamed code 1 as `Treatment` and code 2 as `Placebo`, so the displayed group for
every patient changed.
:::

. . .

Use `relevel()` to change the reference. Use `levels(x) <- ...` only when you
intentionally mean to **rename** categories.


## First inspect `dose_mg` {.smaller}

```{r}
#| label: dose-before-factor
head(trial[trial$arm == "Treatment", c("id", "arm", "dose_mg")], 8)
str(trial$dose_mg)
sort(unique(trial$dose_mg))
```

. . .

::: {.callout-note appearance="simple"}
The values *look* numeric, but the vector is character text. Its sorted order is
lexicographic: `"5"` comes after `"40"`.
:::


## The catastrophe {.smaller}

`dose_mg` came off the CRF as character text. Somebody converts it to a factor
first and later needs it numeric.

```{r}
#| label: catastrophe
wrong <- as.numeric(as.factor(trial$dose_mg))
data.frame(text = trial$dose_mg, wrong = wrong)[31:36, ]
```

. . .

::: {.callout-important appearance="simple"}
40 mg became **4**. 20 mg became **3**. 5 mg became **5**, which is right by
accident and makes it harder to spot.
:::


## How bad, quantitatively {.smaller}

```{r}
#| label: catastrophe-mean
mean(as.numeric(as.factor(trial$dose_mg)))   # wrong
mean(as.numeric(as.character(trial$dose_mg))) # right
```

. . .

::: {.callout-tip appearance="simple"}
## The fix
If `x` is character, use `as.numeric(x)`. If `x` is already a factor, use
`as.numeric(as.character(x))`: recover the labels, *then* convert them.
:::

. . .

::: {.smaller}
And check afterwards. `as.numeric()` on genuinely non-numeric text gives `NA`
with a warning — so `sum(is.na(x))` before and after tells you whether the
conversion lost anything.
:::


## Where factors come from now {.smaller}

::: {style="font-size:0.9em"}
| Source | Factors? |
|---|---|
| `data.frame(...)` in R ≥ 4.0 | **No.** Character stays character |
| `read.csv()` in R ≥ 4.0 | **No** |
| Anything written before 2020 | **Probably yes** — `stringsAsFactors` defaulted to `TRUE` |
| `haven::read_dta()`, `read_sav()` | Labelled columns; factor-adjacent |
| `factor()`, `cut()`, `table()` | Yes, by construction |
:::

. . .

::: {.smaller}
R 4.0.0 (April 2020) changed the default. **Inherited scripts predate it** — and
they were written with the old behaviour assumed.
:::


# 3 · Indexing a data frame {.section-divider}

## Rows first, then columns {.smaller}

```{r}
#| label: df-ij
trial[1:3, c("id", "arm", "sbp_12")]
```

. . .

```{r}
#| label: df-ij-blank
trial[trial$age > 80, c("id", "age")]
```

. . .

::: {.smaller}
`df[i, j]` — **i is rows, j is columns**, and the comma is not optional. A blank
means "all of them".
:::


## Four ways to ask for a column {.smaller}

```{r}
#| label: four-ways
class(trial$age)
class(trial[["age"]])
class(trial[, "age"])
class(trial["age"])
```

. . .

::: {.smaller}
Three hand you a **vector**. The fourth hands you a **one-column data frame** —
`[` with a single argument is list indexing, and `[` on a list returns a list.
:::

. . .

```{r}
#| label: four-ways-cost
#| warning: true
mean(trial["age"])
```


## The `drop` surprise {.smaller}

```{r}
#| label: drop-surprise
class(trial[, c("age", "sbp_0")])
class(trial[, "age"])
```

. . .

Same notation, **different class**, because the result had one column.

. . .

```{r}
#| label: drop-fix
class(trial[, "age", drop = FALSE])
```


## `$` is convenient, and slightly dangerous {.smaller}

```{r}
#| label: dollar-partial
head(trial$ag, 3)
```

. . .

There is no column called `ag`. **`$` partially matched it to `age`.**

. . .

```{r}
#| label: dollar-null
trial$sbp
```

::: {.smaller}
Ambiguous — `sbp_0`, `sbp_6`, `sbp_12` — so you get `NULL`, silently. And `NULL`
propagates: `mean(trial$sbp)` is `NA`, not an error.
:::


## Selecting rows by condition {.smaller}

```{r}
#| label: rows-cond
low <- trial[trial$sbp_12 < 120, c("id", "arm", "sbp_12")]
nrow(low)
head(low, 4)
```

. . .

::: {.smaller}
The condition is a **logical vector of length 60**, and `[` keeps the rows where
it is `TRUE`. Exactly session 3's logical indexing, applied to the row position.
:::


## The `NA` row trap {.smaller}

```{r}
#| label: na-rows
trial[trial$sbp_12 < 120, c("id", "sbp_12")][1:4, ]
```

. . .

Three patients have a missing `sbp_12`. `NA < 120` is `NA`, not `FALSE` — and
**`[` returns a row of `NA`s for every `NA` in the index.**

. . .

```{r}
#| label: na-rows-fix
nrow(trial[trial$sbp_12 < 120, ])
nrow(trial[which(trial$sbp_12 < 120), ])
```


# 4 · Three base verbs {.section-divider}

## `subset()` {.smaller}

```{r}
#| label: subset-basic
subset(trial, arm == "Treatment" & age > 70,
       select = c(id, age, sbp_12))
```

. . .

::: {.smaller}
Columns unquoted, rows as a plain condition, and **`NA` rows dropped for you**.
Compare the alternative:
:::

```r
trial[which(trial$arm == "Treatment" & trial$age > 70), c("id", "age", "sbp_12")]
```


## `subset()` hazard 1: `=` is not `==` {.smaller}

```{r}
#| label: subset-eq
#| warning: true
nrow(subset(trial, arm = "Treatment"))
```

. . .

One `=` instead of `==`. A warning, and **all 60 rows back**.

. . .

::: {.callout-warning appearance="simple"}
The call runs and returns a data frame. Without checking the warning or row count,
this can look like a successful filter.
:::


## `subset()` hazard 2: bare names have two homes {.smaller}

```{r}
#| label: subset-name-lookup
age_cutoff <- 70
nrow(subset(trial, age > age_cutoff))

trial_with_cutoff <- trial
trial_with_cutoff$age_cutoff <- 100
nrow(subset(trial_with_cutoff, age > age_cutoff))
```

. . .

The expression is identical, but `subset()` looks inside the data frame first.
After a column named `age_cutoff` appears, it wins over the external object.

. . .

::: {.callout-warning appearance="simple"}
`subset()` is convenient for interactive work. In a script that matters, use `[`
with explicit names such as `trial$age` so the source of every value is visible.
:::


## Adding columns with `transform()` {.smaller}

```{r}
#| label: transform
trial <- transform(trial,
  change = sbp_12 - sbp_0,
  responder = sbp_12 < 130
)
head(trial[, c("id", "sbp_0", "sbp_12", "change", "responder")], 3)
```

. . .

::: {.smaller}
The equivalent, and what you will see far more often:
:::

```r
trial$change <- trial$sbp_12 - trial$sbp_0
trial$responder <- trial$sbp_12 < 130
```


## Sorting with `order()` {.smaller}

```{r}
#| label: order-basic
head(trial[order(trial$sbp_12), c("id", "arm", "sbp_12")], 4)
```

. . .

```{r}
#| label: order-multi
head(trial[order(trial$arm, -trial$sbp_0), c("id", "arm", "sbp_0")], 3)
```

::: {.smaller}
`order()` returns **positions**, which you then use to index. Multiple keys sort
in the order given; negate a numeric key to reverse it.
:::


## `sort`, `order`, `rank` {.smaller}

```{r}
#| label: sort-order-rank
x <- c(30, 10, 20)
sort(x)
order(x)
rank(x)
```

. . .

::: {style="font-size:0.9em"}
| Function | Returns | Read as |
|---|---|---|
| `sort(x)` | the values, rearranged | "give me the sorted values" |
| `order(x)` | positions, in sorted order | "where do I find them" |
| `rank(x)` | each value's position when sorted | "what place did each come in" |
:::


# 5 · Split–apply–combine {.section-divider}

## The pattern {.smaller}

```{mermaid}
%%| echo: false
%%| fig-width: 10
flowchart LR
  D["<b>trial</b><br/>60 rows"] --> S1["Placebo<br/>30 rows"]
  D --> S2["Treatment<br/>30 rows"]
  S1 --> A1["mean(sbp_12)"]
  S2 --> A2["mean(sbp_12)"]
  A1 --> C["<b>result</b><br/>2 rows"]
  A2 --> C
  style D fill:#f4efe4,stroke:#CFAE70
  style C fill:#CFAE70,stroke:#8a7448,color:#fff
```

. . .

::: {.smaller}
**Split** the data by a grouping variable · **apply** a summary to each piece ·
**combine** the answers into one table. Almost every table in almost every paper
is this.
:::


## Split: form one piece per group {.smaller}

```{r}
#| label: split-one-variable
bp_by_arm <- split(trial$sbp_12, trial$arm)

names(bp_by_arm)
length(bp_by_arm$Placebo)
length(bp_by_arm$Treatment)
head(bp_by_arm$Placebo, 4)
head(bp_by_arm$Treatment, 4)
```

. . .

::: {.smaller}
The grouping vector, `trial$arm`, has one label for every value in
`trial$sbp_12`. `split()` returns a **named list**: one vector per distinct arm.
Values stay in their original within-group order, and missing values stay present.
:::


## Apply, then combine {.smaller}

```{r}
#| label: apply-combine-manual
placebo_mean <- mean(bp_by_arm$Placebo, na.rm = TRUE)
treatment_mean <- mean(bp_by_arm$Treatment, na.rm = TRUE)

arm_means <- c(
  Placebo = placebo_mean,
  Treatment = treatment_mean
)
arm_means
```

. . .

::: {.smaller}
**Apply:** ask the same question—`mean(..., na.rm = TRUE)`—of every group.

**Combine:** place the separate answers into one named result. The result has one
value per group, not one value per patient.
:::


## `tapply()`: all three steps in one call {.smaller}

```{r}
#| label: tapply-correct
sum(is.na(trial$sbp_12))
tapply(trial$sbp_12, trial$arm, mean, na.rm = TRUE)
```

. . .

First document the three missing outcomes; then calculate the observed-value mean
for each arm. What happens if `na.rm = TRUE` is omitted?

```{r}
#| label: tapply-na
tapply(trial$sbp_12, trial$arm, mean)
```

. . .

Each arm contains at least one missing value, so its mean is `NA`. This is a
diagnostic result, not a software failure.

. . .

::: {.smaller}
`tapply(X, INDEX, FUN, ...)`

- `X`: the values to split
- `INDEX`: one group label per value
- `FUN`: the function applied to each group
- `...`: extra arguments passed to `FUN`, such as `na.rm = TRUE`
:::


## `aggregate()`: the formula interface {.smaller}

```{r}
#| label: agg-basic
aggregate(sbp_12 ~ arm, data = trial, FUN = mean)
```

. . .

```{r}
#| label: agg-multi
aggregate(cbind(sbp_0, sbp_12) ~ arm + sex, data = trial, FUN = mean)
```

::: {.smaller}
Read `~` as **"by"**. Left of it: what to summarise. Right of it: what to group
by. `+` on the right adds grouping variables; `cbind()` on the left adds
outcomes.
:::


## `aggregate()` drops rows quietly {.smaller}

```{r}
#| label: agg-count
aggregate(sbp_12 ~ arm, data = trial, FUN = length)
```

. . .

**28 and 29, not 30 and 30.** The formula interface defaults to
`na.action = na.omit` — rows with a missing outcome vanish before `FUN` runs.

. . .

```{r}
#| label: agg-napass
aggregate(sbp_12 ~ arm, data = trial, FUN = mean,
          na.rm = TRUE, na.action = na.pass)
```


## Counting: `table()` and friends {.smaller}

```{r}
#| label: table-two
table(trial$arm, trial$grade)
```

. . .

```{r}
#| label: prop-table
round(prop.table(table(trial$arm, trial$sex), margin = 1), 2)
```

::: {.smaller}
`margin = 1` gives row proportions, `2` gives columns, and omitting it divides by
the grand total. **Getting the margin wrong is the most common table error in
submitted work.**
:::


## You are learning the pattern, not the syntax {.smaller}

```{r}
#| label: base-way
aggregate(sbp_12 ~ arm, data = trial, FUN = mean, na.rm = TRUE)
```

. . .

The same question in session 6 — **not executed here, you have not been taught
it**:

```r
trial |>
  group_by(arm) |>
  summarise(sbp_12 = mean(sbp_12, na.rm = TRUE))
```

. . .

::: {.callout-important appearance="simple"}
Split, apply, combine. The syntax changes in session 6. **The idea does not.**
:::


# 6 · Inspection discipline {.section-divider}

## `str()` first, always {.smaller}

```{r}
#| label: str-trial
str(trial[, c("id", "arm", "age", "sbp_12", "dose_mg", "grade")])
```

::: {.smaller}
Six columns, six lines. Name, **type**, and the first few values — everything you
need to know whether your next line will work.
:::


## Five checks on any new data frame {.smaller}

```{r}
#| label: five-checks
dim(trial)
colSums(is.na(trial[, c("age", "sbp_0", "sbp_12", "dose_mg")]))
```

. . .

::: {.smaller}
1. `str()` — types, and are they what you expected
2. `dim()` — is the row count the number of patients you were promised
3. `summary()` — impossible minima and maxima
4. `table()` on every categorical column, with `useNA = "ifany"`
5. `colSums(is.na(...))` — where the holes are, before anyone asks
:::


## When it goes wrong {.smaller}

::: {style="font-size:0.82em"}
| Symptom | Cause | Fix |
|---|---|---|
| `argument is not numeric or logical: returning NA` | `df["x"]` where you wanted `df[["x"]]` | Use `$` or `[[` |
| Rows of `NA` with rownames like `NA.1` | `NA` in a logical row index | `which()`, or `subset()` |
| `undefined columns selected` | Row/column arguments swapped | Rows first, then columns |
| `incorrect number of dimensions` | `[i, j]` on a vector or a list | Check `str()`; it is not 2-d |
| Numbers become small integers | `as.numeric(as.factor(x))` | `as.numeric(as.character(x))` |
| Filter returns 0 rows for no reason | Column name typo → `$` gave `NULL` | `names(df)`, then `[[` |
| Grouped `n` smaller than expected | `aggregate()` dropped `NA` rows | Add a `FUN = length` check |
:::


## Summary {.smaller}

- Choose a container whose contract matches the operation.
- A data frame is a list of equal-length columns; a matrix has one type.
- `[` preserves a container while `[[` extracts one component.
- State row meaning and keys before subsetting or summarizing.
- Every grouped summary needs denominators and checks.

. . .

**Next connection:** Session 5 begins with files and asks whether the imported data frame matches the study.

