---
title: "Conditions and Functions"
subtitle: "BIOS 6301 · Introduction to Statistical Computing"
author: "Department of Biostatistics, Vanderbilt University Medical Center"
date: "Session 11"
date-format: "[Session 11]"
format:
  revealjs:
    theme: simple
    css: ../../slides.css
    slide-number: c/t
    show-slide-number: all
    incremental: false
    width: 1280
    height: 720
    fig-align: center
    code-line-numbers: false
    highlight-style: github
    footer: "BIOS 6301 · Session 11 · Conditions and Functions"
    navigation-mode: linear
execute:
  enabled: true
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(ggplot2)
library(knitr)

# Synthetic teaching data for BIOS 6301. Not real patients.
set.seed(6301)
n <- 60
trial <- data.frame(
  id        = sprintf("P%03d", 1:n),
  arm       = rep(c("Placebo", "Treatment"), each = n / 2),
  site      = sample(1:4, n, replace = TRUE),
  age       = round(rnorm(n, 61, 9)),
  sex       = sample(c("F", "M"), n, replace = TRUE),
  weight_kg = round(rnorm(n, 82, 15), 1),
  sbp_0     = round(rnorm(n, 148, 14)),
  stringsAsFactors = FALSE
)
trial$sbp_12 <- round(trial$sbp_0 -
  ifelse(trial$arm == "Treatment", rnorm(n, 14, 8), rnorm(n, 3, 8)))

# Added for this session: the derived outcome the refactoring demo summarises.
trial$change <- trial$sbp_12 - trial$sbp_0

# A copy carrying the two defects a real extract arrives with, used for the
# validation demos. What is broken, deliberately:
#   sbp_12    three missing 12-week outcomes (P007, P023, P048)
#   weight_kg one weight recorded as the sentinel -99 (P012)
trial_raw <- trial
trial_raw$sbp_12[c(7, 23, 48)] <- NA
trial_raw$weight_kg[12] <- -99
trial_raw$change <- trial_raw$sbp_12 - trial_raw$sbp_0

theme_set(theme_minimal(base_size = 15))
```

## Where we are {.smaller}

| Course arc | Sessions | |
|---|---:|---|
| Setup and workflow | 0–2 |  |
| Data foundations | 3–8 |  |
| Modern tools | 9–10 |  |
| **R programming** | **11–13** | **← today** |
| Statistical computing | 14–16 |  |

Today's deck develops one part of the current arc; later-session tools are named only when they clarify where this idea will be used.


## What you will be able to do {.smaller}

::: {.smaller}
[1]{.objective-num} **Distinguish** `if` from `ifelse()` and say which one a
given problem needs.

[2]{.objective-num} **Write** a function with arguments, defaults, and a return
value.

[3]{.objective-num} **Explain** where R looks for a name, and why a function that
reads a global variable is not reproducible.

[4]{.objective-num} **Validate** inputs with `stopifnot()`, `stop()`, and guard
clauses.

[5]{.objective-num} **Write** a function that takes a data frame and a column
name.

[6]{.objective-num} **Refactor** a copy-pasted script into functions and verify
the output is unchanged.
:::


## Roadmap for today

| | Block | Minutes |
|---|---|---|
| 1 | The block you have written four times | 8 |
| 2 | `if` is control flow, not `ifelse()` | 12 |
| 3 | Writing functions | 25 |
| 4 | Lexical scoping, and globals | 12 |
| 5 | Input validation | 12 |
| 6 | Functions that take data frames | 10 |
| 7 | Refactoring a script, and house style | 12 |

: {tbl-colwidths="[6,64,30]"}

. . .

**In lab this week** you refactor a working but repetitive script. Today is the
material you need for it.


# 1 · The block you have written four times {.section-divider}

## Look at what you have been doing {.smaller}

```r
# session 6 — baseline SBP by arm
trial |> group_by(arm) |>
  summarise(n = n(), mean = mean(sbp_0), sd = sd(sbp_0))

# session 7 — the same, after the join
trial |> group_by(arm) |>
  summarise(n = n(), mean = mean(sbp_12), sd = sd(sbp_12))

# session 8 — weight, to annotate the figure
trial |> group_by(arm) |>
  summarise(n = n(), mean = mean(weight_kg), sd = sd(weight_kg))

# coursework — age, for the baseline table
trial |> group_by(arm) |>
  summarise(n = n(), mean = mean(age), sd = sd(age))
```

. . .

Four blocks. **One name differs.**


## What copy-paste actually costs {.smaller}

::: {style="font-size:0.88em"}
| Cost | What it looks like in a week |
|---|---|
| **Reading** | A reader has to diff four blocks to find the one word that differs |
| **Changing** | The data gains missing values. You need `na.rm = TRUE` in four places, and you will find three |
| **Naming** | The block has no name, so you cannot say what it does, reuse it, or check it |
| **Divergence** | One copy already rounds to two digits and the others do not. Nobody decided that |
:::

. . .

::: {.callout-important appearance="simple"}
Duplication is not an aesthetic problem. It is the mechanism by which two parts of
one analysis come to disagree.
:::

. . .

By the end of today that block is one line:

```r
summarise_by_arm(trial, sbp_0)
```


# 2 · `if` is control flow, not `ifelse()` {.section-divider}

## Closing a loop from session 3 {.smaller}

::: {style="font-size:0.9em"}
| | `ifelse(test, yes, no)` | `if (cond) ... else ...` |
|---|---|---|
| Is a | **vectorized function** | **control-flow keyword** |
| Test length | Any — one answer per element | **Exactly one** `TRUE` or `FALSE` |
| Returns | A vector as long as `test` | Whatever the chosen branch returns |
| Used for | Deriving a variable from a column | Choosing what a *script* does |
| Taught in | Session 3 | **Today** |
:::

. . .

::: {.callout-note appearance="simple"}
**`ifelse()` derives a variable. `if` decides what the program does.**
:::


## `if`, `else if`, `else` {.smaller}

`if` takes **one** `TRUE` or `FALSE` and picks a branch.

```{r}
#| label: if-basic
n_missing <- sum(is.na(trial_raw$sbp_12))

if (n_missing == 0) {
  "complete outcome data"
} else if (n_missing < 5) {
  "a few missing — a complete-case summary is defensible"
} else {
  "too much missing outcome data to summarise naively"
}
```

. . .

::: {.smaller}
- The condition is a **single** logical value. Here, `n_missing == 0`.
- `else` must sit on the same line as the closing `}`, or R decides the `if` has
  ended.
- Braces are optional for one-line bodies. Use them anyway.
:::


## The error you will actually hit {.smaller}

```{r}
#| label: if-error
#| error: true
if (trial$sbp_0 > 140) "high" else "normal"
```

. . .

`trial$sbp_0` has length `r nrow(trial)`. `if` wanted length **one**.

. . .

Two neighbours worth recognising:

```{r}
#| label: if-error-2
#| error: true
if (NA > 1) "yes" else "no"
```

```{r}
#| label: if-error-3
#| error: true
sbp <- numeric(0)          # a filter that matched nobody
if (sbp > 140) "high"
```


## Which one do I want? {.smaller}

::: {style="font-size:0.92em"}
| The question you are asking | The tool |
|---|---|
| "What value does **each patient** get?" | `ifelse()`, `if_else()`, `case_when()` |
| "What should **this script** do next?" | `if` / `else if` / `else` |
:::

. . .

And `if` is an expression, so it returns a value:

```{r}
#| label: if-value
label <- if (n_missing > 0) "complete-case" else "all randomised"
label
```

. . .

::: {.callout-tip appearance="simple"}
If the answer has one value per row, it is not a job for `if`. Not ever.
:::


# 3 · Writing functions {.section-divider}

## A function is three things {.smaller}

```{r}
#| label: anatomy
#| results: hold
sbp_summary <- function(x, digits = 1) {
  round(c(n = length(x), mean = mean(x), sd = sd(x)), digits)
}

formals(sbp_summary)
body(sbp_summary)
environment(sbp_summary)
```

::: {.smaller}
**Formals** — the arguments it accepts, and their defaults ·
**body** — the code it runs · **environment** — where it looks for names it does
not define itself.
:::


## Your first function, from your own code {.smaller}

The duplicated block, in one place:

```{r}
#| label: first-function
# Return n, mean, and SD of a numeric vector, rounded.
sbp_summary <- function(x, digits = 1) {
  round(c(n = length(x), mean = mean(x), sd = sd(x)), digits)
}

sbp_summary(trial$sbp_0)
sbp_summary(trial$sbp_12)
sbp_summary(trial$weight_kg, digits = 2)
```

. . .

::: {.smaller}
Three calls, one definition. The **name** now says what the block does, which the
block itself never did.
:::


## Arguments and defaults {.smaller}

```{r}
#| label: defaults
# n counts non-missing values, so the reported n matches the mean.
sbp_summary <- function(x, digits = 1, na_rm = TRUE) {
  round(c(
    n    = sum(!is.na(x)),
    mean = mean(x, na.rm = na_rm),
    sd   = sd(x, na.rm = na_rm)
  ), digits)
}

sbp_summary(trial_raw$sbp_12)
sbp_summary(trial_raw$sbp_12, na_rm = FALSE)
```

::: {.smaller}
A default is what the argument is **when the caller says nothing**. Choose the
safe, common case — then let the caller override it.
:::


## How R matches arguments {.smaller}

Three ways, tried in this order: **exact name**, then **partial name**, then
**position**.

```{r}
#| label: matching
sbp_summary(trial$sbp_0, 2)                 # position
sbp_summary(digits = 2, x = trial$sbp_0)    # exact name — order stops mattering
```

. . .

Partial matching works, and is a trap:

```{r}
#| label: matching-error
#| error: true
report <- function(x, na_rm = FALSE, names = TRUE) "ok"
report(1, n = TRUE)
```

. . .

::: {.callout-warning appearance="simple"}
**House rule:** data argument by position, everything else by full name. Partial
names break the day somebody adds an argument.
:::


## Return values, and the implicit return {.smaller}

A function returns the value of its **last evaluated expression**.

```{r}
#| label: returns
mean_sbp   <- function(x) mean(x)          # implicit return
mean_sbp_2 <- function(x) return(mean(x))  # identical, more typing
mean_sbp_3 <- function(x) { m <- mean(x) } # returns m, invisibly

mean_sbp(trial$sbp_0)
mean_sbp_3(trial$sbp_0)      # prints nothing
print(mean_sbp_3(trial$sbp_0))
```

. . .

::: {.callout-note appearance="simple"}
Use `return()` for an **early exit**. For the value at the end, let it fall out.
:::


## What happens inside stays inside {.smaller}

```{r}
#| label: locals
centre_sbp <- function(x) {
  offset <- mean(x)      # local: created on the call, destroyed after
  x - offset
}

centred <- centre_sbp(trial$sbp_0)
round(head(centred, 4), 1)
exists("offset")
```

. . .

::: {.smaller}
Every call gets a **fresh environment**. Locals cannot leak into your workspace,
and two calls cannot interfere with each other.
:::

. . .

::: {.callout-warning appearance="simple"}
`<<-` writes to the enclosing environment and defeats all of this. You will meet
it in inherited code. Do not write it.
:::


## `...` and `invisible()` {.smaller}

`...` collects arguments you did not name and passes them on:

```{r}
#| label: dots
mean_sbp <- function(x, ...) mean(x, ...)

mean_sbp(trial_raw$sbp_12)
mean_sbp(trial_raw$sbp_12, na.rm = TRUE)
```

. . .

`invisible()` returns a value without printing it:

```{r}
#| label: invisible-demo
report_n <- function(data) {
  cat("n =", nrow(data), "\n")
  invisible(nrow(data))
}

report_n(trial)          # prints the message, not the value
n_rows <- report_n(trial)
n_rows
```


# 4 · Lexical scoping, and globals {.section-divider}

## Where R looks for a name {.smaller}

```{mermaid}
%%| echo: false
%%| fig-width: 12
flowchart LR
  L["<b>1. Inside the call</b><br/>locals and arguments"] --> D["<b>2. Where the function<br/>was DEFINED</b><br/>usually the global env"]
  D --> P["<b>3. Attached packages</b><br/>dplyr, ggplot2, …"]
  P --> B["<b>4. base</b>"]
  B --> E["Error:<br/>object not found"]
  style L fill:#CFAE70,stroke:#8a7448,color:#fff
  style D fill:#CFAE70,stroke:#8a7448,color:#fff
  style E fill:#f4efe4,stroke:#CFAE70
```

. . .

::: {.callout-important appearance="simple"}
Step 2 is **where the function was written**, not where it was called from. That
is what "lexical" means.
:::


## The global that vanishes {.smaller}

```{r}
#| label: scope-works
threshold <- 140

is_high <- function(sbp) sbp > threshold   # threshold is not an argument

sum(is_high(trial$sbp_0))
```

. . .

It works. Now restart R — or hand the script to a colleague:

```{r}
#| label: scope-breaks
#| error: true
rm(threshold)
sum(is_high(trial$sbp_0))
```

. . .

::: {.smaller}
This is the honest failure. It is annoying, and it tells you immediately.
:::


## The global that is stale {.smaller}

```{r}
#| label: scope-silent
threshold <- 140
sum(is_high(trial$sbp_0))
```

. . .

Fifty lines further down the document, somebody — possibly you, last Tuesday —
writes:

```{r}
#| label: scope-silent-2
threshold <- 160        # for the sensitivity analysis
sum(is_high(trial$sbp_0))
```

. . .

::: {.callout-important appearance="simple"}
No error. No warning. A different number, and the function looks untouched.
:::


## Everything comes in through an argument {.smaller}

```{r}
#| label: scope-fixed
# Flag patients above a systolic threshold. Threshold is an argument so the
# result depends only on what the caller passed.
is_high <- function(sbp, threshold = 140) {
  sbp > threshold
}

sum(is_high(trial$sbp_0))
sum(is_high(trial$sbp_0, threshold = 160))
```

. . .

::: {.callout-note appearance="simple"}
**The rule:** a function's return value should depend only on its arguments.
Nothing else.
:::

. . .

::: {.smaller}
The sensitivity analysis is now a second *call*, visible on the page, instead of a
reassignment fifty lines away.
:::


# 5 · Input validation {.section-divider}

## Fail loudly, or be wrong quietly {.smaller}

```{r}
#| label: silent-wrong
mean_weight <- function(weight_kg) mean(weight_kg)

mean_weight(trial_raw$weight_kg)   # one weight is recorded as -99
mean_weight(trial$weight_kg)       # the same cohort, repaired
```

. . .

::: {.smaller}
`r round(mean(trial_raw$weight_kg), 1)` kg against
`r round(mean(trial$weight_kg), 1)` kg. Both are plausible. Only one is right, and
the function had no opinion about which.
:::

. . .

::: {.callout-important}
A function that fails loudly on bad input is safer than one that returns a
plausible wrong answer.
:::


## `stopifnot()` {.smaller}

```{r}
#| label: stopifnot-demo
#| error: true
sbp_summary <- function(x, digits = 1) {
  stopifnot(is.numeric(x), length(x) > 0, digits >= 0)

  round(c(n = sum(!is.na(x)), mean = mean(x, na.rm = TRUE)), digits)
}

sbp_summary(trial$arm)
```

. . .

Name the checks and you choose the message:

```{r}
#| label: stopifnot-named
#| error: true
sbp_summary <- function(x, digits = 1) {
  stopifnot(
    "x must be a numeric vector" = is.numeric(x),
    "digits must not be negative" = digits >= 0
  )
  round(mean(x, na.rm = TRUE), digits)
}

sbp_summary(trial$arm)
```


## `stop()` and guard clauses {.smaller}

```{r}
#| label: guard-clause
#| error: true
# Mean weight, refusing sentinel codes. Site 3 used -99 for "not recorded",
# and averaging those understates the mean by about 3 kg.
mean_weight <- function(weight_kg) {
  if (any(weight_kg < 20, na.rm = TRUE)) {
    stop("weight_kg contains values below 20: check for sentinel codes",
         call. = FALSE)
  }
  mean(weight_kg, na.rm = TRUE)
}

mean_weight(trial_raw$weight_kg)
```

. . .

::: {.smaller}
A **guard clause** is a check at the top that exits early. It keeps the happy path
unindented and the failure cases visible.
:::


## How much validation is enough {.smaller}

::: {style="font-size:0.9em"}
| Check | Because |
|---|---|
| **Type and shape** | `is.numeric()`, `is.data.frame()`, `length()`, column names present |
| **Values that cannot be real** | Negative weights, ages over 120, proportions above 1 |
| **What your code silently assumes** | Sorted rows, one row per patient, non-empty groups |
:::

. . .

::: {.smaller}
Do not re-check what R already checks. `mean("a")` errors on its own.
:::

. . .

::: {.callout-tip appearance="simple"}
Rule of thumb: validate what a **collaborator's file** could plausibly get wrong.
Two or three checks is usually right.
:::


# 6 · Functions that take data frames {.section-divider}

## Data argument first {.smaller}

When the column names are fixed, nothing special is needed:

```{r}
#| label: df-function
# Add change from baseline. Fails if the two SBP columns are absent, because
# mutate() would otherwise create a column full of nothing useful.
add_change <- function(data) {
  stopifnot(is.data.frame(data),
            all(c("sbp_0", "sbp_12") %in% names(data)))

  data |> mutate(change = sbp_12 - sbp_0)
}

trial_raw |> add_change() |> summarise(mean_change = mean(change, na.rm = TRUE))
```

. . .

::: {.smaller}
**Data first**, so the function drops into a pipeline like any dplyr verb.
:::


## The wrinkle: `{{ var }}` {.smaller}

When the **column itself** is the argument, dplyr needs to be told:

```{r}
#| label: tidyeval
summarise_by_arm <- function(data, outcome) {
  data |>
    group_by(arm) |>
    summarise(
      n    = sum(!is.na({{ outcome }})),
      mean = round(mean({{ outcome }}, na.rm = TRUE), 1)
    )
}

summarise_by_arm(trial_raw, sbp_12)
summarise_by_arm(trial_raw, weight_kg)
```

::: {.smaller}
```r
# the base R way — no tidy evaluation involved
tapply(trial_raw$sbp_12, trial_raw$arm, mean, na.rm = TRUE)
```
:::


## Without `{{ }}`: two failures {.smaller}

```{r}
#| label: tidyeval-fail
#| error: true
bad_summary <- function(data, outcome) {
  data |> group_by(arm) |> summarise(mean = mean(outcome))
}

bad_summary(trial, sbp_0)
```

. . .

Now the same call, with a stray object in the workspace:

```{r}
#| label: tidyeval-silent
sbp_0 <- c(1000, 2000)      # left over from something else entirely
bad_summary(trial, sbp_0)
```

```{r}
#| label: tidyeval-cleanup
#| include: false
rm(sbp_0)
```


# 7 · Refactoring, and house style {.section-divider}

## Before: the script we all wrote {.smaller}

```r
# baseline
tab_0 <- trial |> group_by(arm) |>
  summarise(n = n(), mean = round(mean(sbp_0), 1), sd = round(sd(sbp_0), 1))

# 12 weeks
tab_12 <- trial |> group_by(arm) |>
  summarise(n = n(), mean = round(mean(sbp_12), 1), sd = round(sd(sbp_12), 1))

# change — and this one has NAs, so it needs na.rm
tab_ch <- trial |> group_by(arm) |>
  summarise(n = n(), mean = round(mean(change, na.rm = TRUE), 1),
            sd = round(sd(change), 1))     # <- na.rm missing here

ggplot(trial, aes(arm, sbp_0)) + geom_boxplot() + labs(y = "SBP baseline")
ggplot(trial, aes(arm, sbp_12)) + geom_boxplot() + labs(y = "SBP 12 wk")
ggplot(trial, aes(arm, change)) + geom_boxplot() + labs(y = "Change")
```

. . .

::: {.smaller}
Six blocks, two shapes, **one bug** — and `n` is wrong in the third block for the
same reason.
:::


## After: three functions {.smaller}

```{r}
#| label: refactor
#| results: hold
# n counts patients contributing to the mean, not patients randomised.
summarise_by_arm <- function(data, outcome, digits = 1) {
  stopifnot(is.data.frame(data), "arm" %in% names(data))

  data |>
    group_by(arm) |>
    summarise(
      n    = sum(!is.na({{ outcome }})),
      mean = round(mean({{ outcome }}, na.rm = TRUE), digits),
      sd   = round(sd({{ outcome }}, na.rm = TRUE), digits)
    )
}

# One boxplot of an outcome by arm, labelled for a clinical reader.
plot_by_arm <- function(data, outcome, y_label) {
  ggplot(data, aes(arm, {{ outcome }}, fill = arm)) +
    geom_boxplot(alpha = 0.7, width = 0.5) +
    labs(x = NULL, y = y_label) +
    theme(legend.position = "none")
}

summarise_by_arm(trial_raw, change)
```


## Verify the refactor {.smaller}

A refactor that changes the output is not a refactor. **Check.**

```{r}
#| label: verify
old <- trial |>
  group_by(arm) |>
  summarise(n = sum(!is.na(sbp_0)),
            mean = round(mean(sbp_0), 1),
            sd = round(sd(sbp_0), 1))

new <- summarise_by_arm(trial, sbp_0)

identical(old, new)
all.equal(old, new)
```

. . .

::: {.smaller}
`identical()` is exact and unforgiving. `all.equal()` tolerates floating-point
difference — which is why it is the right one for numbers. Both in session 12.
:::


## House style for functions {.smaller}

::: {style="font-size:0.88em"}
| Rule | In practice |
|---|---|
| **Verbs for functions, nouns for objects** | `summarise_by_arm()`, not `arm_summary()` |
| **`snake_case`, no exceptions** | `plot_by_arm`, not `plotByArm` |
| **One job per function** | If the name needs "and", it is two functions |
| **About 30 lines** | Longer means it is doing more than one thing |
| **Data argument first** | So it pipes |
| **Validate inputs** | Two or three checks, at the top |
| **One line above saying what it does** | What it takes, what it returns |
| **Comments say *why*** | The code already says what |
:::

. . .

::: {.callout-note appearance="simple"}
All of this is in `style-guide.md`, §7. **Code quality is graded on it.**
:::


## Summary {.smaller}

- `if` controls one decision; vectorized conditionals classify many values.
- A function is an interface: arguments in, one predictable result out.
- Local objects and explicit arguments avoid stale global dependencies.
- Guard clauses turn silent wrong answers into informative failures.
- Refactoring is complete only after the new result matches a trusted result.

. . .

**Next connection:** Session 12 applies these interfaces to debugging and reliable code.

