---
title: "Homework 1: Reproducible R objects"
author: "Your name"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S %Z')`"
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE)
```

```{r study-vectors}
patient_id <- sprintf("P%02d", 1:12)
arm <- rep(c("control", "treated"), 6)
baseline_sbp_text <- c(
  "142", "138", "not recorded", "151", "133", "145",
  "128", "160", "139", "147", "136", "154"
)
week4_sbp_raw <- c(140, 128, 131, 143, 129, 136, 126, 18, 135, 139, NA, 142)
```

## Part 1: Inspect before changing anything

```{r}
# Add class(), typeof(), and length() checks here.
```

Write your coercion prediction and explanation here.

```{r}
# Test your coercion prediction here.
```

## Part 2: Preserve and clean the raw values

```{r}
# Preserve the raw vector, replace the text missing-value code with NA,
# and convert the cleaned values.
```

Explain how you verified that the raw vector was preserved.

## Part 3: Named indexing and a documented correction

```{r}
# Copy week4_sbp_raw, add patient names, correct P08 by name,
# and count missing week-4 measurements.
```

Explain why you preserved the raw vector.

## Part 4: Select patients with logical conditions

```{r}
# Report treated IDs and IDs with both measurements observed.
```

## Part 5: Calculate and interpret change

```{r}
# Calculate named patient-level changes, the two requested means,
# and their contributing sample sizes.
```

Write your three- or four-sentence interpretation here.

## Part 6: Reproducibility

```{r}
sessionInfo()
```

After completing the work, restart R and render this document again. Confirm
that the render succeeds before submission.

Private repository address: `https://github.com/YOUR_GITHUB_USERNAME/BIOS6301-mywork`

Final short commit identifier: `REPLACE_ME`
