---
title: "Random Numbers and Simulation"
subtitle: "BIOS 6301 · Introduction to Statistical Computing"
author: "Department of Biostatistics, Vanderbilt University Medical Center"
date: "Session 14"
date-format: "[Session 14]"
format:
  revealjs:
    theme: simple
    css: ../../slides.css
    slide-number: c/t
    show-slide-number: all
    incremental: false
    width: 1280
    height: 720
    fig-align: center
    code-line-numbers: false
    highlight-style: github
    footer: "BIOS 6301 · Session 14 · Random Numbers and Simulation"
    navigation-mode: linear
execute:
  enabled: true
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(ggplot2)
library(knitr)

# Synthetic teaching data for BIOS 6301. Not real patients.
set.seed(6301)
n <- 60
trial <- data.frame(
  id        = sprintf("P%03d", 1:n),
  arm       = rep(c("Placebo", "Treatment"), each = n / 2),
  site      = sample(1:4, n, replace = TRUE),
  age       = round(rnorm(n, 61, 9)),
  sex       = sample(c("F", "M"), n, replace = TRUE),
  weight_kg = round(rnorm(n, 82, 15), 1),
  sbp_0     = round(rnorm(n, 148, 14)),
  stringsAsFactors = FALSE
)
trial$sbp_12 <- round(trial$sbp_0 -
  ifelse(trial$arm == "Treatment", rnorm(n, 14, 8), rnorm(n, 3, 8)))

theme_set(theme_minimal(base_size = 15))
```

## Where we are {.smaller}

| Course arc | Sessions | |
|---|---:|---|
| Setup and workflow | 0–2 |  |
| Data foundations | 3–8 |  |
| Modern tools | 9–10 |  |
| R programming | 11–13 |  |
| **Statistical computing** | **14–16** | **← today** |

Today's deck develops one part of the current arc; later-session tools are named only when they clarify where this idea will be used.


## What you will be able to do {.smaller}

::: {.smaller}
[1]{.objective-num} **Explain** what a seed is, and why reproducible randomness
is not a contradiction.

[2]{.objective-num} **Draw** from a distribution using the `r*` functions, and
use `d*`, `p*`, `q*` for the other three questions.

[3]{.objective-num} **Structure** a simulation study with the ADEMP scaffold.

[4]{.objective-num} **Write** one function that simulates a dataset and one that
analyses it, and iterate over replicates.

[5]{.objective-num} **Compute** bias, empirical standard error, and **Monte Carlo
error**.

[6]{.objective-num} **Choose** the number of replicates from a target Monte Carlo
SE.

[7]{.objective-num} **Estimate** type I error and power, and say what to do when
the type I error is not what it should be.
:::


## Roadmap for today {.smaller}

| | Block | Minutes |
|---|---|---|
| 1 | Pseudorandom numbers, seeds, and streams | 12 |
| 2 | Drawing from distributions: `sample` and the four prefixes | 13 |
| 3 | ADEMP — a protocol for a simulation study | 10 |
| 4 | Building it: simulate one, analyse one, replicate | 20 |
| 5 | Bias, empirical SE, and Monte Carlo error | 12 |
| 6 | Type I error and power, worked | 18 |
| 7 | Practice and summary | 5 |

: {tbl-colwidths="[6,64,30]"}

. . .

::: {.callout-note appearance="simple"}
**Style policy, from today: your choice.** Base R, tidyverse, or a defensible mix —
consistently, within a document. Today's slides use base R for the simulation
machinery and dplyr for summarising results, and says why when it matters.
:::


# 1 · Pseudorandom numbers {.section-divider}

## There is no randomness in your laptop {.smaller}

`rnorm()` does not produce random numbers. It produces a **deterministic
sequence** that passes the statistical tests we ask of randomness: the right
distribution, no detectable serial correlation, an astronomically long period
before it repeats.

. . .

::: {.smaller}
R's default generator is the **Mersenne–Twister**. Its state is 624 32-bit
integers and its period is $2^{19937}-1$.
:::

```{r}
#| label: rngkind
RNGkind()
```

. . .

::: {.callout-important appearance="simple"}
The word is **pseudorandom**, and the determinism is not a defect. It is the
property the rest of this session exploits.
:::


## The seed picks your starting point {.smaller}

```{r}
#| label: seed-demo
set.seed(42)
rnorm(3)

set.seed(42)
rnorm(3)
```

. . .

```{r}
#| label: seed-demo-2
set.seed(42)
rnorm(3)
rnorm(3)
```

. . .

::: {.callout-note appearance="simple"}
**Reproducible randomness is not a contradiction.** The numbers are random with
respect to your *data* — no structure, no correlation with anything you care
about. They are fixed with respect to your *machine*.
:::


## The state is an ordinary object {.smaller}

```{r}
#| label: random-seed
set.seed(42)
length(.Random.seed)
.Random.seed[1:3]

invisible(rnorm(1))     # one draw ...
.Random.seed[1:3]       # ... and the state has moved
```

. . .

::: {.smaller}
Two consequences that catch people:

- Every draw **advances** the stream. Nothing is reset until you call `set.seed()`
  again.
- **Inserting a chunk changes everything below it.** If a result must be
  reproducible, seed it immediately before it — not once at the top and hope.
:::


## Where to put `set.seed()` {.smaller}

::: {style="font-size:0.86em"}
| Placement | Verdict |
|---|---|
| Once near the top of the document | **Fine** for an analysis. The whole render is reproducible |
| Immediately before a simulation or a resample | **Best.** Stable even if you edit above it |
| Inside a function that gets called repeatedly | **Catastrophic.** Every replicate is identical, and nothing errors |
| `set.seed(i)` inside a loop over `i` | Reproducible, but the streams are not *provably* independent |
| Nowhere | Not reproducible. Not submittable |
:::

. . .

::: {.callout-warning appearance="simple"}
**Running in parallel?** Workers do not inherit a usable stream. Use
`RNGkind("L'Ecuyer-CMRG")` and seed once before `parallel::mclapply()`, or hand
each worker a stream from `parallel::nextRNGStream()`. Know the name; the details
are Advanced Statistical Computing.
:::


# 2 · Drawing from distributions {.section-divider}

## `sample()` {.smaller}

```{r}
#| label: sample-demo
set.seed(13)
sample(5)                              # a permutation of 1:5
sample(trial$id, 4)                    # 4 patients, no repeats
sample(c("A", "B"), 8, replace = TRUE) # allocation, with replacement
```

. . .

::: {.smaller}
- Default is **without replacement**, and `size` defaults to `length(x)`.
- `replace = TRUE` is the one-word difference between *permuting* and
  *resampling* — which is the whole of session 15.
- `prob =` gives unequal selection probabilities.
:::


## The `sample()` trap {.smaller}

```{r}
#| label: sample-trap
set.seed(13)
sites <- c(7)          # one site left after exclusions
sample(sites)          # you wanted 7
```

. . .

::: {.smaller}
`sample(x)` with a length-1 numeric `x` samples from `1:x`. **No error, no
warning, plausible output** — the exact failure mode session 12 was about.
:::

. . .

```{r}
#| label: sample-safe
# The idiom from ?sample. Correct for every length, including 0 and 1.
resample <- function(x, ...) x[sample.int(length(x), ...)]
set.seed(13)
resample(sites)
```


## Four questions, four prefixes {.smaller}

::: {style="font-size:0.88em"}
| Prefix | Question it answers | Example |
|---|---|---|
| `d` | What is the **density** (or pmf) at `x`? | `dnorm(0)` |
| `p` | What is $P(X \le q)$ — the **CDF**? | `pnorm(1.96)` |
| `q` | Which `q` has $P(X \le q) = p$ — the **quantile**? | `qnorm(0.975)` |
| `r` | Give me `n` **random** draws | `rnorm(10)` |
:::

. . .

::: {.smaller}
Attach any prefix to any root: `norm`, `unif`, `binom`, `pois`, `t`, `chisq`,
`exp`, `gamma`, `beta`, `nbinom`, `weibull`, …
:::

```{r}
#| label: prefix-demo
c(dnorm(0), pnorm(1.96), qnorm(0.975), qt(0.975, df = 18))
```


## `rnorm`, `runif`, `rbinom` {.smaller}

```{r}
#| label: r-funs
set.seed(13)
round(rnorm(3, mean = 148, sd = 14), 1)  # n, then the parameters
round(runif(3, min = 0, max = 1), 3)
rbinom(3, size = 1, prob = 0.3)          # 3 coin flips
rbinom(3, size = 10, prob = 0.3)         # 3 counts out of 10
```

. . .

::: {.smaller}
**Parameters recycle**, which is how you generate a whole trial in one call:
:::

```{r}
#| label: r-vectorised
set.seed(13)
round(rnorm(6, mean = rep(c(148, 134), each = 3), sd = 14))
```


# 3 · ADEMP — a protocol for a simulation {.section-divider}

## A simulation is an experiment {.smaller}

You would not run a trial by collecting data until the result looked right. A
simulation has the same failure mode and none of the oversight.

. . .

::: {.smaller}
Written down in advance, a simulation study is evidence. Written down afterwards,
it is a story about the settings that worked.
:::

. . .

::: {.callout-note appearance="simple"}
**ADEMP** — Morris, White & Crowther, "Using simulation studies to evaluate
statistical methods," *Statistics in Medicine* **38**:2074–2102, 2019. Read it
before you design one for real; it is the clearest thing written on the subject.
:::


## ADEMP {.smaller}

::: {style="font-size:0.86em"}
| | | You must be able to write down |
|---|---|---|
| **A** | Aims | The question, in one sentence |
| **D** | Data-generating mechanism | Exactly how **one** dataset is produced. Every parameter, every value |
| **E** | Estimands | The quantity being estimated — a number that exists before any method touches it |
| **M** | Methods | The procedures being compared, including the naive one |
| **P** | Performance measures | How you decide which method won, **and with what precision** |
:::

. . .

::: {.smaller}
Five headings, and the discipline is that you fill them in **before** you write
any code.
:::


## ADEMP for today {.smaller}

::: {style="font-size:0.82em"}
| | |
|---|---|
| **A** | Does a two-sample *t*-test hold its nominal 5% type I error when the arms have different variances **and** different sizes? |
| **D** | $y_{\text{ctrl}} \sim N(0, 12^2)$, $n_{\text{ctrl}}$ patients; $y_{\text{trt}} \sim N(\delta, 4^2)$, $n_{\text{trt}}$ patients. Allocations 25/25, 10/40, 40/10 |
| **E** | $\delta$, the difference in mean SBP change. Null case $\delta = 0$; power cases $\delta = 3, 6, 9, 12$ mmHg |
| **M** | Pooled *t*-test (`var.equal = TRUE`) vs Welch (`var.equal = FALSE`) |
| **P** | Rejection rate at $\alpha = 0.05$; bias and empirical SE of $\hat\delta$. $n_{\text{sim}} = 10{,}000$, chosen for a Monte Carlo SE of 0.0022 |
:::

. . .

::: {.smaller}
The SDs are not invented: 12 and 4 mmHg bracket what our `trial` cohort shows for
`sbp_12` by arm.
:::


## Estimands — the letter people skip {.smaller}

::: {.callout-important appearance="simple"}
The estimand is a property of the **data-generating mechanism**, not of a method.
If you cannot write down its true value before running anything, you cannot
compute bias.
:::

. . .

::: {.smaller}
- Two methods can both be "unbiased" — for **different** estimands. A marginal
  risk difference and a conditional odds ratio are not the same target, and a
  simulation that mixes them will report bias that is an artefact of the
  question, not the method.
- In today's study the estimand is $\delta$, and we know it because we **set**
  it. That is the entire reason simulation can measure bias at all.
:::


# 4 · Building the simulation {.section-divider}

## The shape of every simulation {.smaller}

```{mermaid}
%%| echo: false
%%| fig-width: 10
flowchart LR
  P["<b>parameters</b><br/>n, sd, delta"] --> S["<b>simulate_one()</b><br/>one dataset"]
  S --> A["<b>analyse_one()</b><br/>one row of results"]
  A --> R["<b>replicate</b><br/>× n_sim"]
  R --> T["<b>results</b><br/>one row per replicate"]
  T --> M["<b>performance</b><br/>bias, SE, rejection rate"]
  style S fill:#CFAE70,stroke:#8a7448,color:#fff
  style A fill:#CFAE70,stroke:#8a7448,color:#fff
  style P fill:#f4efe4,stroke:#CFAE70
```

. . .

::: {.smaller}
Session 11 gave you **functions**. Session 13 gave you **iteration**. This is what
they were for.
:::


## `simulate_one()` {.smaller}

```{r}
#| label: sim-one
# Simulate one two-arm trial. Returns a data frame with `arm` and `y`.
# Control is N(0, sd_ctrl^2); treatment is N(delta, sd_trt^2).
simulate_trial <- function(n_ctrl, n_trt, sd_ctrl, sd_trt, delta = 0) {
  stopifnot(n_ctrl >= 2, n_trt >= 2, sd_ctrl > 0, sd_trt > 0)

  data.frame(
    arm = rep(c("ctrl", "trt"), c(n_ctrl, n_trt)),
    y   = c(rnorm(n_ctrl, mean = 0,     sd = sd_ctrl),
            rnorm(n_trt,  mean = delta, sd = sd_trt))
  )
}

set.seed(6301)
head(simulate_trial(3, 3, 12, 4, delta = 8), 3)
```


## `analyse_one()` {.smaller}

```{r}
#| label: analyse-one
# Analyse one simulated trial with a two-sample t-test.
# Returns a named numeric vector: estimate, p-value, CI limits.
analyse_trial <- function(data, pooled = TRUE) {
  stopifnot(is.data.frame(data), all(c("arm", "y") %in% names(data)))

  fit <- t.test(y ~ arm, data = data, var.equal = pooled)
  c(est = unname(diff(fit$estimate)),    # trt minus ctrl
    p   = fit$p.value,
    lo  = -fit$conf.int[2],              # t.test's CI is ctrl minus trt
    hi  = -fit$conf.int[1])
}

set.seed(6301)
round(analyse_trial(simulate_trial(20, 20, 12, 4, delta = 8)), 3)
```


## One replicate, then ten thousand {.smaller}

```{r}
#| label: replicate-demo
run_scenario <- function(n_sim, n_ctrl, n_trt, sd_ctrl, sd_trt,
                         delta = 0, pooled = TRUE) {
  out <- replicate(n_sim, {
    d <- simulate_trial(n_ctrl, n_trt, sd_ctrl, sd_trt, delta)
    analyse_trial(d, pooled = pooled)
  })
  as.data.frame(t(out))          # n_sim rows, one per replicate
}

set.seed(6301)
demo <- run_scenario(5, 20, 20, 12, 4, delta = 8)
round(demo, 3)
```


## Store results tidily {.smaller}

Keep **one row per replicate**, and carry every parameter that varies:

```{r}
#| label: tidy-results
set.seed(6301)
res <- run_scenario(200, 10, 40, 12, 4, delta = 0, pooled = TRUE)
res$n_ctrl <- 10; res$n_trt <- 40; res$method <- "pooled"

res |>
  group_by(method) |>
  summarise(reject = mean(p < 0.05), bias = mean(est), emp_se = sd(est))
```

::: {.smaller}
::: {.callout-tip appearance="simple"}
**The base R way**
```r
c(reject = mean(res$p < 0.05), bias = mean(res$est), emp_se = sd(res$est))
```
:::
:::


## Many scenarios {.smaller}

```{r}
#| label: grid
grid <- expand.grid(
  alloc  = c("25 / 25", "10 / 40", "40 / 10"),
  method = c("pooled", "welch"),
  delta  = c(0, 6),
  stringsAsFactors = FALSE
)
nrow(grid)
head(grid, 3)
```

. . .

::: {.smaller}
`nrow(grid)` scenarios × `n_sim` replicates is the whole cost of the study.
**Compute that number before you start the run**, not while it is running.
:::


## Cache the run {.smaller}

::: {style="font-size:0.88em"}
| Level | How | Use when |
|---|---|---|
| Chunk cache | `#\| cache: true` | Seconds to a minute. Needs `codetools` |
| Results file | `saveRDS(res, "output/sim.rds")` | Minutes to hours |
| Separate script | `01-run-sim.R` writes; the `.qmd` reads | Long enough that it should never run on render |
:::

. . .

```r
sim_file <- here::here("output", "sim-typeI.rds")
if (!file.exists(sim_file)) {
  set.seed(6301)
  saveRDS(run_all(grid, n_sim = 10000), sim_file)
}
res <- readRDS(sim_file)
```

. . .

::: {.callout-warning appearance="simple"}
Save the **seed and the parameter grid** with the results. A cached object you
cannot regenerate is not reproducible — it is just a file.
:::


## A buggy simulation does not crash {.smaller}

```{r}
#| label: seed-bug
simulate_bad <- function(n_ctrl, n_trt, sd_ctrl, sd_trt, delta = 0) {
  set.seed(6301)                       # <- the bug
  data.frame(
    arm = rep(c("ctrl", "trt"), c(n_ctrl, n_trt)),
    y   = c(rnorm(n_ctrl, 0, sd_ctrl), rnorm(n_trt, delta, sd_trt))
  )
}

bad <- as.data.frame(t(replicate(500, analyse_trial(simulate_bad(25, 25, 12, 4)))))
c(reject = mean(bad$p < 0.05), bias = mean(bad$est), emp_se = sd(bad$est))
```

. . .

::: {.smaller}
No error. No warning. A type I error of **0**, a bias of exactly zero, and a
perfectly plausible-looking estimate — from 500 copies of the same dataset.
:::


## Assertions for simulations {.smaller}

::: {style="font-size:0.86em"}
| Where | Check |
|---|---|
| **Arguments** | `stopifnot(n_ctrl >= 2, sd_ctrl > 0, length(delta) == 1)` |
| **The DGM** | Simulate once at $n = 10^6$; the empirical mean and SD must match the parameters |
| **The results object** | `nrow(res) == n_sim`, no `NA`, and **`sd(res$est) > 0`** |
| **A known case** | A setting where you know the answer must return it, to within Monte Carlo error |
:::

. . .

```{r}
#| label: dgm-check
set.seed(6301)
check <- simulate_trial(1e6, 1e6, 12, 4, delta = 8)
stopifnot(
  abs(mean(check$y[check$arm == "trt"]) - 8) < 0.05,
  abs(sd(check$y[check$arm == "ctrl"]) - 12) < 0.05
)
"DGM checks passed"
```


# 5 · Performance measures {.section-divider}

## What you actually report {.smaller}

::: {style="font-size:0.84em"}
| Measure | Estimated by | Answers |
|---|---|---|
| **Bias** | $\bar{\hat\theta} - \theta$ | Is the estimator systematically off? |
| **Empirical SE** | $\mathrm{sd}(\hat\theta)$ | How variable is it, run to run? |
| **Coverage** | $\frac{1}{n_{\text{sim}}}\sum \mathbb{1}\{\hat\theta_{\text{lo}} \le \theta \le \hat\theta_{\text{hi}}\}$ | Does the 95% CI cover 95% of the time? |
| **Type I error** | $\mathrm{mean}(p < \alpha)$, with $\theta = 0$ | Does the test hold its size? |
| **Power** | $\mathrm{mean}(p < \alpha)$, with $\theta \ne 0$ | Will the study detect a real effect? |
:::

. . .

::: {.smaller}
Type I error and power are the **same computation** on data from the same DGM
with `delta` set differently. Nothing else changes.
:::


## Monte Carlo error {.smaller}

Every number on the previous slide is **itself an estimate**, from $n_{\text{sim}}$
draws. It has a standard error.

::: {style="font-size:0.86em"}
| Quantity | Monte Carlo SE |
|---|---|
| Bias | $\mathrm{EmpSE}/\sqrt{n_{\text{sim}}}$ |
| Empirical SE | $\mathrm{EmpSE}/\sqrt{2(n_{\text{sim}}-1)}$ |
| A rejection rate $\hat p$ | $\sqrt{\hat p(1-\hat p)/n_{\text{sim}}}$ |
:::

. . .

```{r}
#| label: mcse
p_hat <- 0.052; n_sim <- 1000
sqrt(p_hat * (1 - p_hat) / n_sim)      # MCSE
p_hat + c(-1.96, 1.96) * sqrt(p_hat * (1 - p_hat) / n_sim)
```

. . .

::: {.smaller}
At $n_{\text{sim}} = 1000$, an observed 0.052 is consistent with anything from
0.038 to 0.066. **That simulation cannot tell 5% from 6.5%.**
:::


## Choosing $n_{\text{sim}}$ on purpose {.smaller}

Invert the formula. For a rejection rate near $p$:

$$n_{\text{sim}} = \frac{p(1-p)}{\text{MCSE}^2}$$

::: {style="font-size:0.86em"}
| Target MCSE at $p = 0.05$ | $n_{\text{sim}}$ | Resolves |
|---|---|---|
| 0.010 | 475 | 0.05 vs 0.08 |
| 0.005 | 1,900 | 0.05 vs 0.065 |
| **0.0025** | **7,600** | 0.05 vs 0.057 |
| 0.001 | 47,500 | 0.05 vs 0.053 |
:::

. . .

::: {.smaller}
For **bias**, $n_{\text{sim}} = (\mathrm{EmpSE}/\text{target})^2$. With an
empirical SE of 3.9 mmHg and a target of 0.1 mmHg, that is 1,521.
:::

. . .

::: {.callout-important appearance="simple"}
**1000 is a habit, not a decision.** State the difference you need to detect, then
compute $n_{\text{sim}}$ — exactly as you would a sample size for a trial.
:::


# 6 · Type I error and power {.section-divider}

## The question, restated {.smaller}

Two arms. **Different variances** — 12 vs 4 mmHg. **Different sizes** — and it is
the *combination* that does the damage.

::: {.smaller}
The pooled *t*-test assumes a common variance. It estimates one $s_p$ from both
arms, weighting by degrees of freedom, and gets $n_1 + n_2 - 2$ df. Welch
estimates each arm's variance separately and adjusts the df downwards.
:::

. . .

::: {.smaller}
`t.test()` in R defaults to `var.equal = FALSE`. Most software does not, most
textbooks teach the pooled version first, and most people tick the box that says
"assume equal variances" because it sounds like the simpler model.
:::

. . .

::: {.callout-note appearance="simple"}
Four scenarios, $\delta = 0$, $n_{\text{sim}} = 10{,}000$. If the test is behaving,
every number should be 0.05 ± 0.0044.
:::


## Run it {.smaller}

```{r}
#| label: type1-run
n_sim <- 10000
set.seed(6301)

scenarios <- list(
  list(lab = "25 / 25", nc = 25, nt = 25, pooled = TRUE),
  list(lab = "10 / 40", nc = 10, nt = 40, pooled = TRUE),
  list(lab = "40 / 10", nc = 40, nt = 10, pooled = TRUE),
  list(lab = "10 / 40", nc = 10, nt = 40, pooled = FALSE)
)

type1 <- do.call(rbind, lapply(scenarios, function(s) {
  r <- run_scenario(n_sim, s$nc, s$nt, sd_ctrl = 12, sd_trt = 4,
                    delta = 0, pooled = s$pooled)
  data.frame(alloc = s$lab, method = ifelse(s$pooled, "pooled", "Welch"),
             reject = mean(r$p < 0.05), bias = mean(r$est), emp_se = sd(r$est))
}))
type1$mcse <- sqrt(type1$reject * (1 - type1$reject) / n_sim)
```


## The type I error is not 0.05 {.smaller}

```{r}
#| label: type1-table
#| echo: false
type1 |>
  transmute(
    Allocation = alloc, Method = method,
    `Rejection rate` = sprintf("%.4f", reject),
    `MC SE` = sprintf("%.4f", mcse),
    Bias = sprintf("%+.3f", bias),
    `Emp. SE` = sprintf("%.2f", emp_se)
  ) |>
  kable(align = "llrrrr")
```

::: {.smaller}
$\sigma_{\text{ctrl}} = 12$, $\sigma_{\text{trt}} = 4$, $\delta = 0$,
$n_{\text{sim}} = 10{,}000$. Nominal $\alpha = 0.05$.
:::

. . .

::: {.smaller}
- **10 / 40 pooled** — the small arm is the noisy one. Rejects **five times too
  often**.
- **40 / 10 pooled** — reverse the allocation and the test becomes absurdly
  conservative.
- **Bias is essentially zero in every row.** The estimate was never the problem.
:::


## Under the null, p-values must be uniform {.smaller}

```{r}
#| label: fig-pvals
#| echo: false
#| fig-width: 9.5
#| fig-height: 3.4
#| fig-cap: "Null p-values, 10,000 replicates, n = 10 / 40, sd = 12 / 4."
set.seed(97)
pv <- rbind(
  data.frame(method = "Pooled t-test",
             p = run_scenario(10000, 10, 40, 12, 4, 0, TRUE)$p),
  data.frame(method = "Welch t-test",
             p = run_scenario(10000, 10, 40, 12, 4, 0, FALSE)$p)
)
ggplot(pv, aes(p)) +
  geom_histogram(breaks = seq(0, 1, 0.05), fill = "#CFAE70", colour = "white") +
  geom_hline(yintercept = 500, linetype = 2) +
  facet_wrap(~method) +
  labs(x = "p-value under the null", y = "count")
```

::: {.smaller}
The dashed line is what a correct test gives: 500 per bin. The pooled test piles
up near zero — and the excess *is* the inflated type I error.
:::


## Welch, and what it costs {.smaller}

::: {.smaller}
Welch does not pool. It uses $\sqrt{s_1^2/n_1 + s_2^2/n_2}$ and a
Satterthwaite-approximated degrees of freedom, which is smaller — often much
smaller — than $n_1 + n_2 - 2$.
:::

::: columns
::: {.column width="48%"}
### Under unequal variances

Holds its size. **0.0485** against a nominal 0.05, with a Monte Carlo SE of
0.0021.
:::

::: {.column width="48%"}
### Under equal variances

Costs almost nothing. A percent or two of power in the worst case, and nothing at
all in most.
:::
:::

. . .

::: {.callout-important appearance="simple"}
`var.equal = TRUE` is a **choice you have to justify**. R's default is the safe
one; use it unless you can say why the variances are equal.
:::


## Power, from the same machinery {.smaller}

```{r}
#| label: fig-power
#| echo: false
#| fig-width: 9
#| fig-height: 3.6
#| fig-cap: "Welch power. Total N = 50 in every scenario. sd = 12 (ctrl) / 4 (trt)."
set.seed(555)
alloc <- data.frame(lab = c("10 / 40", "25 / 25", "38 / 12"),
                    nc = c(10, 25, 38), nt = c(40, 25, 12))
pw_grid <- expand.grid(i = 1:3, delta = c(0, 3, 6, 9, 12))
pw <- do.call(rbind, lapply(seq_len(nrow(pw_grid)), function(k) {
  i <- pw_grid$i[k]
  r <- run_scenario(2000, alloc$nc[i], alloc$nt[i], 12, 4,
                    delta = pw_grid$delta[k], pooled = FALSE)
  data.frame(alloc = alloc$lab[i], delta = pw_grid$delta[k],
             power = mean(r$p < 0.05))
}))
ggplot(pw, aes(delta, power, colour = alloc, shape = alloc)) +
  geom_line(linewidth = 0.9) + geom_point(size = 2.6) +
  geom_hline(yintercept = 0.8, linetype = 2, colour = "grey40") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "true difference in means (mmHg)", y = "power",
       colour = "ctrl / trt", shape = "ctrl / trt")
```

::: {.smaller}
Same DGM, same functions, one argument changed: `delta`. Fifty patients in every
scenario — only the **allocation** differs.
:::


## What the simulation actually told us {.smaller}

::: {.smaller}
- The pooled *t*-test's **type I error was 0.26**, not 0.05, when the smaller arm
  had the larger variance.
- The **estimate was unbiased throughout**. Only the inference broke.
- Welch fixed it, at negligible cost.
- With unequal variances, **allocate more patients to the noisier arm**.
:::

. . .

::: {.smaller}
None of that required any new mathematics. It required a DGM you could write
down, two short functions, ten thousand replicates, and the discipline to report a
Monte Carlo SE.
:::

. . .

::: {.callout-note appearance="simple"}
This is the general move: **when you do not know whether a method works under
your conditions, build your conditions and find out.**
:::


## Summary {.smaller}

- A seed identifies a pseudorandom stream; it does not make a design valid.
- Write ADEMP before writing the repetition.
- Test one simulated replicate before running thousands.
- Report Monte Carlo uncertainty with every simulated probability.
- Type I error and power differ by the data-generating parameter, not by the code skeleton.

. . .

**Next connection:** Session 15 changes the simulated world into bootstrap, permutation, and cross-validation worlds.

