---
title: "Visualization"
subtitle: "BIOS 6301 · Introduction to Statistical Computing"
author: "Department of Biostatistics, Vanderbilt University Medical Center"
date: "Session 8"
date-format: "[Session 8]"
format:
  revealjs:
    theme: simple
    css: ../../slides.css
    slide-number: c/t
    show-slide-number: all
    incremental: false
    width: 1280
    height: 720
    fig-align: center
    code-line-numbers: false
    highlight-style: github
    footer: "BIOS 6301 · Session 8 · Visualization"
    navigation-mode: linear
    auto-stretch: false
execute:
  enabled: true
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(dplyr)
library(knitr)

# Synthetic teaching data for BIOS 6301. Not real patients.
set.seed(6301)
n <- 60
trial <- data.frame(
  id        = sprintf("P%03d", 1:n),
  arm       = rep(c("Placebo", "Treatment"), each = n / 2),
  site      = sample(1:4, n, replace = TRUE),
  age       = round(rnorm(n, 61, 9)),
  sex       = sample(c("F", "M"), n, replace = TRUE),
  weight_kg = round(rnorm(n, 82, 15), 1),
  sbp_0     = round(rnorm(n, 148, 14)),
  stringsAsFactors = FALSE
)
trial$sbp_12 <- round(trial$sbp_0 -
  ifelse(trial$arm == "Treatment", rnorm(n, 14, 8), rnorm(n, 3, 8)))

# Added for this session only, and nothing else is changed:
#   sbp_6  - a middle visit, so a ribbon has somewhere to run
#   crp    - a deliberately right-skewed lab value, for the log-scale slide
#   site_f - site as a labelled factor, so facet strips read as words
trial$sbp_6 <- round(trial$sbp_0 -
  ifelse(trial$arm == "Treatment", rnorm(n, 8, 7), rnorm(n, 2, 7)))
trial$crp    <- round(exp(rnorm(n, 1.1, 0.9)), 2)
trial$site_f <- factor(trial$site, levels = 1:4,
                       labels = paste("Site", 1:4))

# Long form, one row per patient-visit. Session 7 would build this with
# tidyr::pivot_longer(); assembled here in base R because this deck's R
# install is deliberately minimal.
trial_long <- data.frame(
  id   = rep(trial$id, 3),
  arm  = rep(trial$arm, 3),
  week = rep(c(0, 6, 12), each = n),
  sbp  = c(trial$sbp_0, trial$sbp_6, trial$sbp_12)
)

theme_set(theme_minimal(base_size = 14))

# Every figure fills the width of its container; fig-width and fig-height then
# control the aspect ratio and the text size relative to the panel.
knitr::opts_chunk$set(out.width = "100%")
```

## Where we are {.smaller}

| Course arc | Sessions | |
|---|---:|---|
| Setup and workflow | 0–2 |  |
| **Data foundations** | **3–8** | **← today** |
| Modern tools | 9–10 |  |
| R programming | 11–13 |  |
| Statistical computing | 14–16 |  |

Today's deck develops one part of the current arc; later-session tools are named only when they clarify where this idea will be used.


## What you will be able to do {.smaller}

::: {.smaller}
[1]{.objective-num} **Read** base-R plotting code and produce a fast diagnostic
panel with it.

[2]{.objective-num} **Name** the components of the grammar of graphics and map an
existing figure onto them.

[3]{.objective-num} **Build** a plot in layers — distributions, comparisons,
relationships, and **uncertainty**.

[4]{.objective-num} **Control** facets, scales, transformations, and position
adjustments.

[5]{.objective-num} **Apply** the perceptual hierarchy and a colour-blind-safe
palette on purpose.

[6]{.objective-num} **Critique** a figure against explicit criteria — including
your own.

[7]{.objective-num} **Export** a publication figure reproducibly at a stated size
and resolution.
:::


## Roadmap for today

| | Block | Minutes |
|---|---|---|
| 1 | Base graphics, and why we then move on | 10 |
| 2 | The grammar of graphics | 15 |
| 3 | Layers: geoms, facets, scales, positions | 30 |
| 4 | **Design principles** | 25 |
| 5 | Publication output | 8 |
| 6 | Practice and summary | 2 |

: {tbl-colwidths="[6,64,30]"}

. . .

Every figure in this deck was **rendered when the deck was built**. None of them
is a screenshot.


# 1 · Base graphics, briefly {.section-divider}


## Base graphics in four functions {.smaller}

```{r}
#| label: base-four
#| fig-width: 9
#| fig-height: 2.8
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
hist(trial$sbp_0, main = "hist", xlab = "SBP baseline")
boxplot(sbp_12 ~ arm, data = trial, main = "boxplot", xlab = "")
plot(trial$sbp_0, trial$sbp_12, main = "plot", xlab = "Session 0",
     ylab = "Session 12")
```

::: {.smaller}
`par(mfrow = c(1, 3))` splits the device into a 1 × 3 grid. It is **device
state**: it stays set until you change it back.
:::


## The one base-graphics habit worth keeping {.smaller}

```{r}
#| label: base-diagnostic
#| fig-width: 9
#| fig-height: 2.5
par(mfrow = c(1, 4), mar = c(4, 4, 2, 1))
hist(trial$age,       main = "age")
hist(trial$weight_kg, main = "weight_kg")
hist(trial$sbp_0,     main = "sbp_0")
hist(trial$crp,       main = "crp")
```

. . .

::: {.smaller}
Four columns, one screen, five seconds. You are not making a figure — you are
**checking that the file is what you think it is**. `crp` is right-skewed and
you now know it before you fit anything.
:::


## Why we then move on {.smaller}

::: {style="font-size:0.88em"}
| Base graphics | Consequence |
|---|---|
| Each function has its own arguments | `hist(breaks=)`, `boxplot(names=)`, `plot(pch=)` — nothing transfers |
| A plot is a **sequence of side effects** on a device | You cannot store it, modify it, or return it from anything |
| Grouping means splitting the data yourself | Colour by arm is your job, and so is the legend |
| Faceting has no equivalent | You manage `mfrow` and the axis limits by hand |
| Styling is argument soup | Twenty arguments and a `par()` call to get one journal figure |
:::

. . .

**None of this is a bug.** Base graphics is thirty years old, it draws anything,
and it is in every R installation. It simply has no *grammar* — no shared
vocabulary that composes.


# 2 · The grammar of graphics {.section-divider}


## A plot is a specification {.smaller}

Wilkinson's idea, by way of Wickham: stop naming chart *types*. Describe the
**mapping from data to visual channels**, and the chart type falls out.

. . .

::: {style="font-size:0.9em"}
| Instead of | Say |
|---|---|
| "a grouped bar chart" | x = arm, y = mean, fill = sex, bars, dodged |
| "a scatter plot with a trend line" | x = Session 0, y = Session 12, points **and** a smooth |
| "a boxplot" | x = arm, y = SBP, boxplot statistic |
:::

. . .

::: {.callout-important appearance="simple"}
There is no list of chart types to memorise. There is a small set of components
that combine, which is why the learning curve is worth it: the fifth plot type
costs you nothing new.
:::


## Seven components, plus a theme {.smaller}

::: {style="font-size:0.84em"}
| Component | Question it answers | Code |
|---|---|---|
| **data** | What is being plotted? | `ggplot(trial, ...)` |
| **aesthetics** | Which variable drives which channel? | `aes(x = arm, y = sbp_12)` |
| **geoms** | What marks appear? | `geom_point()`, `geom_col()` |
| **stats** | What is computed before drawing? | `stat_smooth()`, the count in a histogram |
| **scales** | How do data values become positions, colours, sizes? | `scale_y_log10()` |
| **coords** | What is the coordinate system? | `coord_flip()`, `coord_cartesian()` |
| **facets** | How is the plot split into panels? | `facet_wrap(~ site_f)` |
| **themes** | What does everything *not* data-driven look like? | `theme_minimal()` |
:::

. . .

::: {.smaller}
Only **data**, **aesthetics** and one **geom** are required. Everything else has
a default that is usually right.
:::


## Minimum viable ggplot, then a layer {.smaller}

::: columns
::: {.column width="49%"}
```{r}
#| label: mvp
#| fig-width: 4.6
#| fig-height: 2.9
ggplot(trial, aes(sbp_0, sbp_12)) +
  geom_point()
```
:::

::: {.column width="49%"}
```{r}
#| label: mvp-plus
#| fig-width: 4.6
#| fig-height: 2.9
ggplot(trial, aes(sbp_0, sbp_12)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm")
```
:::
:::

::: {.smaller}
Same data, same aesthetics, **one more layer**. The second geom inherits both.
:::


## `aes()` means *map*. Outside means *set* {.smaller}

::: columns
::: {.column width="49%"}
```{r}
#| label: aes-inside
#| fig-width: 4.6
#| fig-height: 2.6
ggplot(trial, aes(sbp_0, sbp_12,
                  colour = arm)) +
  geom_point()
```
:::

::: {.column width="49%"}
```{r}
#| label: aes-outside
#| fig-width: 4.6
#| fig-height: 2.6
ggplot(trial, aes(sbp_0, sbp_12)) +
  geom_point(colour = "steelblue")
```
:::
:::

::: {.smaller}
Inside `aes()`: colour **carries information**, and you get a legend.
Outside: colour is decoration, and no legend is possible.
:::


# 3 · Layers in practice {.section-divider}


## Distributions: histogram and density {.smaller}

::: columns
::: {.column width="49%"}
```{r}
#| label: geom-hist
#| fig-width: 4.6
#| fig-height: 2.6
ggplot(trial, aes(sbp_0)) +
  geom_histogram(binwidth = 8,
                 fill = "grey35")
```
:::

::: {.column width="49%"}
```{r}
#| label: geom-density
#| fig-width: 4.6
#| fig-height: 2.6
ggplot(trial, aes(sbp_0)) +
  geom_density(fill = "grey80")
```
:::
:::

::: {.smaller}
Set `binwidth` in the units of the variable — "8 mmHg" is a decision you can
defend. `bins = 30` is a default you cannot.

**The base R way:** `hist(trial$sbp_0)` · `plot(density(trial$sbp_0))`
:::


## Distributions compared: boxplot and violin {.smaller}

::: columns
::: {.column width="49%"}
```{r}
#| label: geom-box
#| fig-width: 4.6
#| fig-height: 2.6
ggplot(trial, aes(arm, sbp_12)) +
  geom_boxplot(width = 0.5)
```
:::

::: {.column width="49%"}
```{r}
#| label: geom-violin
#| fig-width: 4.6
#| fig-height: 2.6
ggplot(trial, aes(arm, sbp_12)) +
  geom_violin(fill = "grey85")
```
:::
:::

::: {.smaller}
A boxplot shows five numbers. A violin shows a density. **Neither shows n**, and
neither would look different if half the data were missing.

**The base R way:** `boxplot(sbp_12 ~ arm, data = trial)`
:::


## Show the data, not only the summary {.smaller}

```{r}
#| label: geom-jitter
#| fig-width: 9
#| fig-height: 2.9
ggplot(trial, aes(arm, sbp_12)) +
  geom_boxplot(width = 0.45, outlier.shape = NA) +
  geom_jitter(width = 0.12, alpha = 0.7, size = 2)
```

::: {.smaller}
`outlier.shape = NA` stops the boxplot drawing outliers that `geom_jitter()`
already drew. Two marks for one observation is a real error, not a style choice.
:::


## Comparisons: bars, and position adjustments {.smaller}

```{r}
#| label: geom-bar
#| fig-width: 9
#| fig-height: 2.7
ggplot(trial, aes(site_f, fill = sex)) +
  geom_bar(position = "dodge")
```

::: {.smaller}
| `position =` | Shows | Good for |
|---|---|---|
| `"stack"` *(default)* | Counts, one bar per group | Totals |
| `"dodge"` | Counts, side by side | Comparing subgroups |
| `"fill"` | Proportions, all bars to 1 | Composition — but hides n |
:::


## Relationships: points and a smooth {.smaller}

```{r}
#| label: geom-smooth
#| fig-width: 9
#| fig-height: 2.8
ggplot(trial, aes(age, sbp_12, colour = arm)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.15)
```

::: {.smaller}
`method = "lm"` is a model you have chosen. The default, `"loess"`, is a model
you have not — and the band is a confidence band for the *fit*, not a prediction
interval.
:::


## Uncertainty: an estimate needs an interval {.smaller}

```{r}
#| label: geom-pointrange
#| fig-width: 9
#| fig-height: 2.6
by_arm <- trial |>
  group_by(arm) |>
  summarise(mean_sbp = mean(sbp_12),
            se = sd(sbp_12) / sqrt(n()))

ggplot(by_arm, aes(arm, mean_sbp)) +
  geom_pointrange(aes(ymin = mean_sbp - 1.96 * se,
                      ymax = mean_sbp + 1.96 * se))
```


## Uncertainty over time: the ribbon {.smaller}

```{r}
#| label: geom-ribbon
#| fig-width: 9
#| fig-height: 2.4
by_week <- trial_long |>
  group_by(arm, week) |>
  summarise(m = mean(sbp), se = sd(sbp) / sqrt(n()),
            .groups = "drop")

ggplot(by_week, aes(week, m, colour = arm, fill = arm)) +
  geom_ribbon(aes(ymin = m - 1.96 * se, ymax = m + 1.96 * se),
              alpha = 0.2, colour = NA) +
  geom_line(linewidth = 1) +
  scale_x_continuous(breaks = c(0, 6, 12))
```


## Facets: the same plot, split {.smaller}

```{r}
#| label: facet-wrap
#| fig-width: 9
#| fig-height: 2.6
ggplot(trial, aes(sbp_0, sbp_12, colour = arm)) +
  geom_point(alpha = 0.8) +
  facet_wrap(~ site_f, nrow = 1)
```

::: {.smaller}
`facet_wrap(~ x)` wraps one variable into panels · `facet_grid(a ~ b)` makes a
matrix of two · `scales = "free_y"` lets each panel set its own axis, **which
breaks comparability**.
:::


## Scales: where the transformation belongs {.smaller}

::: columns
::: {.column width="49%"}
```{r}
#| label: scale-raw
#| fig-width: 4.6
#| fig-height: 2.5
ggplot(trial, aes(arm, crp)) +
  geom_jitter(width = 0.1)
```
:::

::: {.column width="49%"}
```{r}
#| label: scale-log
#| fig-width: 4.6
#| fig-height: 2.5
ggplot(trial, aes(arm, crp)) +
  geom_jitter(width = 0.1) +
  scale_y_log10()
```
:::
:::

::: {.smaller}
`scale_y_log10()` transforms the **scale** and keeps the axis in original units.
`aes(y = log10(crp))` transforms the **data** and labels the axis `log10(crp)`.
Prefer the scale.
:::


# 4 · Design principles {.section-divider}


## The perceptual hierarchy {.smaller}

How accurately a reader can decode a quantity, best first:

::: {style="font-size:0.9em"}
| Rank | Channel | Use it for |
|---|---|---|
| 1 | **Position** on a common scale | The number that carries your argument |
| 2 | **Length** | Second choice; needs a zero baseline |
| 3 | Angle, slope | Trends, reluctantly |
| 4 | **Area** | Rough magnitude only |
| 5 | Colour intensity, saturation | A third variable, not the main one |
:::

. . .

::: {.callout-important appearance="simple"}
**Position beats length beats area beats colour.** Put your primary comparison on
a shared axis, and never encode the effect you care about in colour.
:::


## Critique: what is wrong here? {.smaller}

```{r}
#| label: bad-figure
#| echo: false
#| fig-width: 8.4
#| fig-height: 3.3
ggplot(by_arm, aes(arm, mean_sbp, fill = arm)) +
  geom_col(width = 0.7) +
  coord_cartesian(ylim = c(128, 146)) +
  labs(title = "SBP results", x = "Group", y = "SBP") +
  theme_grey(base_size = 14)
```

::: {.smaller}
Take thirty seconds. Name three things.
:::


## The same data, honestly {.smaller}

```{r}
#| label: good-figure
#| echo: false
#| fig-width: 8.4
#| fig-height: 3.4
ggplot(trial, aes(arm, sbp_12)) +
  geom_jitter(width = 0.12, alpha = 0.45, size = 1.8,
              colour = "grey40") +
  geom_pointrange(
    data = by_arm,
    aes(y = mean_sbp,
        ymin = mean_sbp - 1.96 * se,
        ymax = mean_sbp + 1.96 * se),
    colour = "#0072B2", linewidth = 0.9, size = 0.7
  ) +
  labs(x = NULL, y = "Systolic BP at 12 weeks (mmHg)",
       title = "SBP at 12 weeks by arm, with mean and 95% CI",
       subtitle = "n = 30 per arm; grey points are individual patients") +
  theme_minimal(base_size = 14)
```


## Two chart types to refuse {.smaller}

::: columns
::: {.column width="46%"}
```{r}
#| label: bad-pie
#| echo: false
#| fig-width: 3.6
#| fig-height: 2.7
ggplot(trial, aes(x = "", fill = site_f)) +
  geom_bar(width = 1) +
  coord_polar(theta = "y") +
  labs(fill = NULL) +
  theme_void(base_size = 12)
```
:::

::: {.column width="52%"}
::: {.smaller}
**Pie charts** ask the reader to compare **angles and areas** — ranks 3 and 4.
Rank the four sites from that. Now imagine a bar chart.

**Dual axes** are worse, because the reader cannot check them:

```r
# do not
p + scale_y_continuous(
  sec.axis = sec_axis(~ . * 3.2, name = "CRP")
)
```

The crossing point is set by *your* choice of scaling. Two facets, or two
panels, say the same thing without the implied claim.
:::
:::
:::


## Colour-blind-safe, on purpose {.smaller}

::: columns
::: {.column width="49%"}
```{r}
#| label: pal-okabe
#| echo: false
#| fig-width: 4.4
#| fig-height: 1.15
ok <- data.frame(x = 1:8,
                 hex = unname(palette.colors(8, "Okabe-Ito")))
ggplot(ok, aes(x, 1, fill = hex)) +
  geom_tile(colour = "white", linewidth = 1) +
  scale_fill_identity() +
  labs(title = "Okabe-Ito (categorical)") +
  theme_void(base_size = 11)
```
:::

::: {.column width="49%"}
```{r}
#| label: pal-viridis
#| echo: false
#| fig-width: 4.4
#| fig-height: 1.15
ggplot(data.frame(x = 1:8), aes(x, 1, fill = factor(x))) +
  geom_tile(colour = "white", linewidth = 1) +
  scale_fill_viridis_d(guide = "none") +
  labs(title = "viridis (ordered / continuous)") +
  theme_void(base_size = 11)
```
:::
:::

::: {.smaller}
```r
# categorical - built in to R since 4.0
palette.colors(8, "Okabe-Ito")
#> #000000 #E69F00 #56B4E9 #009E73 #F0E442 #0072B2 #D55E00 #CC79A7
scale_colour_manual(values = palette.colors(3, "Okabe-Ito"))

# ordered or continuous - built in to ggplot2
scale_colour_viridis_d()   # discrete
scale_fill_viridis_c()     # continuous
```

Then print it in greyscale. If the groups stop being distinguishable, colour
was doing work that shape or a direct label should have been doing.
:::


## Label the thing, not the legend {.smaller}

```{r}
#| label: direct-label
#| echo: false
#| fig-width: 8.6
#| fig-height: 3.1
ggplot(by_week, aes(week, m, colour = arm)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  annotate("text", x = 12.4, y = 133, label = "Treatment",
           colour = "#0072B2", hjust = 0, fontface = "bold") +
  annotate("text", x = 12.4, y = 145, label = "Placebo",
           colour = "#D55E00", hjust = 0, fontface = "bold") +
  scale_colour_manual(values = c("#D55E00", "#0072B2")) +
  scale_x_continuous(breaks = c(0, 6, 12),
                     limits = c(0, 16)) +
  labs(x = "Week", y = "Mean SBP (mmHg)") +
  theme(legend.position = "none")
```

::: {.smaller}
A legend makes the reader look away, decode, and look back — for every series.
A label next to the line costs them nothing.
:::


## Ink that is not information {.smaller}

Every mark on a figure is either **carrying information** or **competing with
it**.

::: {style="font-size:0.88em"}
| Remove | Because |
|---|---|
| Heavy grid lines, panel background | They are as visually loud as the data |
| A legend that repeats the x axis | Zero information |
| A third dimension on a 2-D quantity | Distorts area and hides marks |
| Decorative colour | Spends the strongest labelling channel on nothing |
| Gridlines on a categorical axis | There is nothing to read off |

:::

. . .

::: {.callout-tip appearance="simple"}
`theme_minimal()` gets you most of the way for free. The rest is deletion, and
the test is: **remove it. Did the figure lose anything?**
:::


# 5 · Publication output {.section-divider}


## Finish it: `labs()` and a theme {.smaller}

```{r}
#| label: publication
#| fig-width: 8.6
#| fig-height: 2.5
p_final <- ggplot(trial, aes(arm, sbp_12)) +
  geom_jitter(width = 0.12, alpha = 0.45, colour = "grey40") +
  geom_pointrange(data = by_arm,
                  aes(y = mean_sbp,
                      ymin = mean_sbp - 1.96 * se,
                      ymax = mean_sbp + 1.96 * se),
                  colour = "#0072B2") +
  labs(x = NULL, y = "SBP at 12 weeks (mmHg)",
       caption = "Mean and 95% CI. n = 30 per arm.") +
  theme_minimal(base_size = 13)
p_final
```


## `ggsave()`, with the arguments filled in {.smaller}

```r
ggsave(
  filename = here::here("output", "fig-sbp-arm.png"),
  plot     = p_final,
  width    = 6.5,     # inches - the column width of your journal
  height   = 4,
  dpi      = 300,     # 300 for raster; journals ask for 300-600
  bg       = "white"
)
```

::: {.smaller}
| Argument | Leave it out and you get |
|---|---|
| `width`, `height` | The last size of your **plot pane**. Irreproducible by construction |
| `dpi` | 300 by default, but state it — reviewers ask |
| `bg` | A transparent background that prints as grey in Word |
| `.pdf` / `.svg` extension | Vector output: infinitely scalable, and what a typesetter wants |
:::


## Why the plot pane is not a workflow {.smaller}

::: columns
::: {.column width="48%"}
### Export → paste

Right-click, copy, paste into Word.

**Nothing records what produced it.** The data changes, the figure does not, and
no warning appears anywhere.
:::

::: {.column width="48%"}
### `ggsave()` in the document

The figure is regenerated from the data on every render.

**It cannot disagree with the analysis**, because there is only one copy of the
truth.
:::
:::

. . .

::: {.callout-warning appearance="simple"}
A pasted figure is the same class of error as a number typed by hand — and it is
harder to spot, because a stale figure still looks like a figure.
:::


## `patchwork`: several panels, one figure {.smaller}

```r
library(patchwork)

p_box  <- ggplot(trial, aes(arm, sbp_12)) + geom_boxplot()
p_scat <- ggplot(trial, aes(sbp_0, sbp_12)) + geom_point()
p_time <- ggplot(by_week, aes(week, m, colour = arm)) + geom_line()

(p_box | p_scat) / p_time +
  plot_annotation(tag_levels = "A")
```

::: {.smaller}
`|` side by side · `/` stacked · `plot_annotation(tag_levels = "A")` labels the
panels **A**, **B**, **C** the way a journal wants them.

Not installed on this machine — `install.packages("patchwork")`.
:::



## Summary {.smaller}

- Choose a figure from the statistical question and observation unit.
- A ggplot is a specification of mappings, layers, scales, coordinates, facets, and theme.
- Mapping a variable differs from setting a display property.
- Show distributions and uncertainty rather than only summary bars.
- Exported dimensions, labels, accessibility, and reproducibility are part of the figure.

. . .

**Next connection:** Session 9 introduces Positron and GitHub Copilot; course-approved AI use begins only under that session's policy.

